### 2024-08-10 README.Rmd
Package updates

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stgeomx

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/stgeomx)](https://CRAN.R-project.org/package=stgeomx)
[![R-CMD-check](https://github.com/mkiyer/stgeomx/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mkiyer/stgeomx/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The stgeomx package should be used to assist in the processing and analysis of Nanostring Geomx RNA data. It has a novel background correction function (qq) and a quantile normalization method that correct for background noise in data generated by GeoMx. This package can be used to load GeoMx data from xslx or tsv files, merge datasets together, visualize the quality of the data then set thresholds to filter out probes or ROIs, and apply the package's background correction and normalization to the data. See our manuscript for more information about the package and see our vignette for step-by-step instructions on how to use the package.

## Installation

You can install and load the development version of stgeomx from CRAN with:

``` {r installation CRAN}
install.packages("path/to/your/package", repos = NULL, type = "source")
library("stgeomx")
```

You can install and load the development version of stgeomx from GitHub with:
``` {r installation github}
# install.packages("devtools")
# devtools::install_github("mkiyer/stgeomx")
```

### 2024-08-10 R/preprocess.R

@@ -29,7 +29,7 @@ geomean <- function(x) {
 #' @param x vector of numeric values
 #' @param bg vector of TRUE/FALSE corresponding to x
 #' @param bw.adjust bandwidth adjustment multiplier for 'density' function
-#' @return vector of quantiles corresponding to array 'x'
+#' @returns vector of quantiles corresponding to array 'x'
 compute_quantiles_kde <- function(x, bg, bw.adjust=2) {
   # setup
   n <- 2^13
@@ -185,6 +185,14 @@ calc_snauc <- function(meta, counts) {
 #' @param bg_lod_quantile number (0.0-1.0)
 #' @returns dataset
 #' @export
+#'
+#' @examples
+#' data(example_ds, package= "stgeomx")
+#' preprocess(example_ds)
+#'
+#' data(example_ds, package= "stgeomx")
+#' preprocess(example_ds, 0.8)
+#'
 preprocess <- function(ds, bg_lod_quantile=0.9) {
   # first remove empty/control AOIs
   ds <- remove_empty_aois(ds)
@@ -254,6 +262,12 @@ preprocess <- function(ds, bg_lod_quantile=0.9) {
 #' @param ds list produced by st_geomx_process
 #' @returns list
 #' @export
+#'
+#' @examples
+#' data(example_ds, package= "stgeomx")
+#' example_ds <- preprocess(example_ds)
+#' merge_probes(example_ds)
+#'
 merge_probes <- function(ds) {
   # convenience function for geomean
   geomean_fast <- function(x) {
@@ -323,6 +337,12 @@ merge_probes <- function(ds) {
 #' @param gene_min_frac_expr number minimum frac of samples expressing gene
 #' @returns list dataset
 #' @export
+#'
+#' @examples
+#' data(example_ds, package= "stgeomx")
+#' example_ds <- preprocess(example_ds)
+#' set_thresholds(example_ds, 50000, 0.6, 0.2, 0.15)
+#'
 set_thresholds <- function(ds,
                            min_counts = 0,
                            min_auc = 0.5,
@@ -355,6 +375,13 @@ set_thresholds <- function(ds,
 #' @param ds list dataset produced by set_thresholds
 #' @returns list dataset
 #' @export
+#'
+#' @examples
+#' data(example_ds, package= "stgeomx")
+#' example_ds <- preprocess(example_ds)
+#' example_ds <- set_thresholds(example_ds)
+#' apply_filters(example_ds)
+#'
 apply_filters <- function(ds) {
[…]

### 2024-08-10 R/stgeomx.R

@@ -8,8 +8,31 @@
 #' @param samples tibble with sample information
 #' @param meta tibble with gene metadata
 #' @param counts tibble with gene count data
+#' @param x tibble with gene count data
 #' @returns dataset
 #' @export
+#'
+#' @examples
+#' samples <- data.frame(
+#'   SlideName = c("brain_001", "brain_002"),
+#'   ScanLabel = c("Brain 1", "Brain 2"),
+#'   aoi = c("s01_r001_fullroi", "s02_r001_fullroi")
+#' )
+#' meta <- data.frame(
+#'   probe = c("example01", "NegProbe-WTX01"),
+#'   gene = c("example", "NegProbe-WTX"),
+#'   bg = c(FALSE, TRUE)
+#' )
+#' counts <- data.frame(
+#'   s01_r001_fullroi <- c(100, 32),
+#'   s02_r001_fullroi <- c(95, 24)
+#' )
+#' x <- data.frame(
+#'   s01_r001_fullroi <- c(100, 32),
+#'   s02_r001_fullroi <- c(95, 24)
+#' )
+#' new_ds <- init(samples, meta, counts, x)
+#'
 init <- function(samples, meta, counts, x) {
   ds <- list(
     samples = samples,
@@ -27,6 +50,11 @@ init <- function(samples, meta, counts, x) {
 #' @param ds stgeomx dataset
 #' @returns dataset
 #' @export
+#'
+#' @examples
+#' data(example_ds, package = "stgeomx")
+#' duplicate <- copy(example_ds)
+#'
 copy <- function(ds) {
   x <- list(
     samples = ds$samples,

### 2024-08-10 tests/testthat/test-bgcorrect.R

@@ -1,0 +1,3 @@
+test_that("obsolete_bgsub_simple", {
+  expect_equal(2*2, 4)
+})

### 2024-08-10 tests/testthat/test-input.R

@@ -1,0 +1,25 @@
+test_that("simple_xlsx", {
+  expect_no_error(read_xlsx(system.file("extdata", "simple_geomx.xlsx", package = "stgeomx")))
+})
+
+test_that("duplicate_probe_xlsx", {
+  expect_error(read_xlsx(system.file("extdata", "test_input_duplicate_probe.xlsx", package = "stgeomx")), "duplicate probe found")
+})
+
+test_that("simple_tsv", {
+  expect_no_error(read_tsv(system.file("extdata", "simple_geomx_sample.tsv", package = "stgeomx"), system.file("extdata", "simple_geomx_count.tsv", package = "stgeomx")))
+})
+
+test_that("duplicate_probe_tsv", {
+  expect_error(read_tsv(system.file("extdata", "test_input_duplicate_probe_sample.tsv", package = "stgeomx"), system.file("extdata", "test_input_duplicate_probe_count.tsv", package = "stgeomx"), "duplicate probe found"))
+})
+
+test_that("blank_count_xlsx", {
+  blank_count <- read_xlsx(system.file("extdata", "blank_count.xlsx", package = "stgeomx"))
+  expect_equal(blank_count[["counts"]][["s01_r001_glucagon+"]][1], 0)
+})
+
+test_that("blank_count_xlsx", {
+  blank_count <- read_tsv(system.file("extdata", "blank_count_sample.tsv", package = "stgeomx"), system.file("extdata", "blank_count_count.tsv", package = "stgeomx"))
+  expect_equal(blank_count[["counts"]][["s01_r001_glucagon+"]][1], 0)
+})

### 2024-08-27 R/stgeomx.R
updated doc

@@ -8,6 +8,7 @@
 #' @param samples tibble with sample information
 #' @param meta tibble with gene metadata
 #' @param counts tibble with gene count data
+#' @param x tibble with normalized gene count data
 #' @returns dataset
 #' @export
 init <- function(samples, meta, counts, x) {

### 2024-08-28 R/preprocess.R
moved merge function updated plot_bgcorrect function

@@ -393,44 +393,3 @@ apply_filters <- function(ds) {
   return(fds)
 }
 
-#'
-#' Merge two datasets
-#'
-#' This function should only be used to merge datasets that are generated from the same probe kit.
-#'
-#' This function can join datasets with two options: union and intersect. Intersect will keep columns that are common to both datasets, whereas union will keep all columns.
-#'
-#' @param a dataset produced by st_geomx_read
-#' @param b dataset produced by st_geomx_read
-#' @returns dataset
-#' @export
-#'
-#' @examples
-#' # example code
-#' data(example_ds, package= "stgeomx")
-#' merge_ds(example_ds, example_ds)
-#'
-merge_ds <- function(a, b) {
-  # merge sample tables
-  sample_cols <- intersect(colnames(a$samples), colnames(b$samples))
-  s <- bind_rows(select(a$samples, all_of(sample_cols)),
-                 select(b$samples, all_of(sample_cols)))
-  # merge count data
-  a_counts <- bind_cols(select(a$meta, probe, gene, bg), a$counts)
-  b_counts <- bind_cols(select(b$meta, probe, gene, bg), b$counts)
-  counts <- inner_join(a_counts, b_counts, by=c("probe", "gene", "bg"),
-                       relationship="one-to-one")
-
-  # Madison edit: added suffix to avoid addition of ".y" to the end of column names
-  counts <- inner_join(a_counts, b_counts, by=c("probe", "gene", "bg"), suffix = c("", ""),
-                       relationship="one-to-one")
-
-  # break into metadata and count data
-  meta <- select(counts, -all_of(s$aoi))
-  counts <- select(counts, all_of(s$aoi))
-
-  # make new dataset object
-  ds <- stgeomx::init(samples=s, meta=meta, counts=counts, x=counts)
-  return(ds)
-}
-

### 2024-08-28 R/qcplots.R

@@ -218,14 +218,16 @@ plot_gene_filter <- function(ds, min_frac_expr=0) {
 }
 
 
-
 #'
-#' qc plot to assess relationship of bg versus gene
+#' qc plot of background noise (x axis) versus gene quantiles (y axis)
+#' uses sample metadata column 'keep' to filter aois before plotting
+#' uses gene metadata column 'keep' to filter genes before plotting
 #'
 #' @import ggplot2
 #' @import patchwork
 #'
 #' @param ds list dataset
+#' @param logscale TRUE for log scale, FALSE for linear scale
 #' @param quantiles vector of quantiles [0.0-1.0]
 #' @returns ggplot object
 #' @export
@@ -233,96 +235,41 @@ plot_gene_filter <- function(ds, min_frac_expr=0) {
 #' @examples
 #' data(example_ds, package = "stgeomx")
 #' example_ds <- preprocess(example_ds)
-#' plot_bgcorrect1(example_ds)
+#' plot_bgcorrect(example_ds)
 #'
-plot_bgcorrect1 <- function(ds, quantiles=c(0.10, 0.25, 0.5, 0.75, 0.90, 0.95, 0.99)) {
+plot_bgcorrect <- function(ds, logscale=TRUE, quantiles=c(0.10, 0.25, 0.5, 0.75, 0.90, 0.95, 0.99)) {
   # labels for plot legend
   quant_names <- paste0("q", round(100 * quantiles))
-  s <- select(ds$samples, "aoi", "slide", "keep")
-  x <- ds$x
-
-  # bg cpm before bgcorrect
+  # gene metadata
   bg <- ds$meta$bg
-  xbg <- filter(cpm(ds$counts), bg)
-  qbg <- xbg %>%
-    reframe(across(everything(), ~ stats::quantile(.x, quantiles))) %>%
-    mutate(name="bg", quant=quant_names)
+  keep <- ds$meta$keep
 
-  # gene cpm before bgcorrect
-  xgene <- filter(ds$counts, !bg)
-  xgene <- as_tibble(cpm(xgene))
-  qgene <- xgene %>%
-    reframe(across(everything(), ~ stats::quantile(.x, quantiles))) %>%
-    mutate(name="gene", quant=quant_names)
+  # sample columns used for plot
+  s <- ds$samples %>% filter(.data$keep) %>% select("aoi", "slide")
 
-  # cpm after bgcorrect
[…]

### 2024-08-28 R/stgeomx.R

@@ -65,3 +65,48 @@ copy <- function(ds) {
   class(x) <- "stgeomx"
   return(x)
 }
+
+
+#'
+#' Merge two datasets
+#'
+#' This function should only be used to merge datasets that are generated
+#' from the same probe kit.
+#'
+#' This function can join datasets with two options: union and intersect.
+#' Intersect will keep columns that are common to both datasets,
+#' whereas union will keep all columns.
+#'
+#' @param a dataset produced by st_geomx_read
+#' @param b dataset produced by st_geomx_read
+#' @returns dataset
+#' @export
+#'
+#' @examples
+#' # example code
+#' data(example_ds, package= "stgeomx")
+#' merge_ds(example_ds, example_ds)
+#'
+merge <- function(a, b) {
+  # merge sample tables
+  sample_cols <- intersect(colnames(a$samples), colnames(b$samples))
+  s <- bind_rows(select(a$samples, all_of(sample_cols)),
+                 select(b$samples, all_of(sample_cols)))
+  # merge count data
+  a_counts <- bind_cols(select(a$meta, probe, gene, bg), a$counts)
+  b_counts <- bind_cols(select(b$meta, probe, gene, bg), b$counts)
+  # counts <- inner_join(a_counts, b_counts, by=c("probe", "gene", "bg"),
+  #                      relationship="one-to-one")
+
+  # Madison edit: added suffix to avoid addition of ".y" to the end of column names
+  counts <- inner_join(a_counts, b_counts, by=c("probe", "gene", "bg"), suffix = c("", ""),
+                       relationship="one-to-one")
+
+  # break into metadata and count data
+  meta <- select(counts, -all_of(s$aoi))
+  counts <- select(counts, all_of(s$aoi))
+
+  # make new dataset object
+  ds <- stgeomx::init(samples=s, meta=meta, counts=counts, x=counts)
+  return(ds)
+}

### 2024-08-28 inst/manuscript_figures.Rmd

---
title: "soa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{soa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Setup

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(tidyverse)
library(readxl)
library(ggrepel)
library(pheatmap)

# bioconductor packages
library(factoextra)

```

## Vignette-specific functions

```{r vignette functions, include=FALSE}

add_study_field <- function(ds, study) {
  s <- ds$samples
  s$study <- study
  s <- s %>% mutate(patient = sprintf("%s_%s", study, patient),
                    slide = sprintf("%s_%s", study, slide),
                    roi = sprintf("%s_%s", study, roi),
                    aoi = sprintf("%s_%s", study, aoi))
  colnames(ds$counts) <- s$aoi
  return(list(samples = s, meta = ds$meta, counts = ds$counts))
}

# merge_ds <- function(a, b) {
#   # merge sample tables
#   sample_cols <- intersect(colnames(a$samples), colnames(b$samples))
#   s <- bind_rows(select(a$samples, all_of(sample_cols)),
#                  select(b$samples, all_of(sample_cols)))
#   
#   # merge count data
#   a_counts <- bind_cols(select(a$meta, probe, gene, bg), a$counts)
#   b_counts <- bind_cols(select(b$meta, probe, gene, bg), b$counts)
#   # Madison edit: added suffix to avoid addition of ".y" to the end of column names
#   counts <- inner_join(a_counts, b_counts, by=c("probe", "gene", "bg"), suffix = c("", ""),
#                        relationship="one-to-one")
# 
#   # break into metadata and count data
#   meta <- select(counts, -all_of(s$aoi))
#   counts <- select(counts, all_of(s$aoi))
# 
#   # make new dataset object
#   ds <- stgeomx::init(samples=s, meta=meta, counts=counts, x=counts)
#   return(ds)
# }

save_pheatmap_pdf <- function(x, filename, width=7, height=7) {
  stopifnot(!missing(x))
  stopifnot(!missing(filename))
  pdf(filename, width=width, height=height)
  grid::grid.newpage()
  grid::grid.draw(x$gtable)
  dev.off()
}

```

## Input files and parameter settings

```{r input files, include=FALSE}

# input directories
working_dir <- "/Users/madisoncaldwell/Desktop/stgeomx"

# EDITED (Madison)
data_dir <- "/Users/madisoncaldwell/Desktop/research/data"

# geomx data file location
geomx_dir <- file.path(working_dir, "geomx")

#
# spatial organ atlas datasets
#

soa_panc_geomx_xlsx <- file.path(data_dir, "spatial_organ_atlas_pancreas.xlsx")
soa_panc_geomx_annot_xlsx <- file.path(data_dir, "pancreas_annot.xlsx")
soa_kidney_geomx_xlsx <- file.path(data_dir, "spatial_organ_atlas_kidney.xlsx")
soa_kidney_geomx_annot_xlsx <- file.path(data_dir, "kidney_annot.xlsx")
soa_brain_geomx_xlsx <- file.path(data_dir, "spatial_organ_atlas_brain.xlsx")
soa_brain_geomx_annot_xlsx <- file.path(data_dir, "brain_annot.xlsx")
soa_colon_geomx_xlsx <- file.path(data_dir, "spatial_organ_atlas_colon.xlsx")
soa_colon_annot_xlsx <- file.path(data_dir, "colon_annot.xlsx")
soa_ln_geomx_xlsx <- file.path(data_dir, "spatial_organ_atlas_lymph_node.xlsx")
soa_ln_geomx_annot_xlsx <- file.path(data_dir, "lymph_node_annot.xlsx")

#
# output directory settings
#
results_dir <- file.path(working_dir, "results")
plot_dir <- file.path(results_dir, "plots")
# create output directories
for (d in c(results_dir, plot_dir)) {
  if (!dir.exists(d)) {
    dir.create(d)
  }
}

#
# filtering parameter settings
#
# gene probes must be greater than the Nth quantile of the noise/background
# probes to be considered 'detectable'
geomx_negprobe_lod_quantile = 0.9
# minimum number of total counts for each area-of-interest (AOI)
geomx_min_counts = 50000
# signal-to-noise area under the curve (snAUC) threshold for each AOI
geomx_min_auc = 0.60
# AOI must have at least this fraction of all genes detectable
geomx_aoi_min_frac_expr = 0.2
# a gene must be 'detectable' in at least this fraction of the AOIs to be kept
geomx_gene_min_frac_expr = 0.15

#
# background correction / normalization parameter settings
#
# background (noise) correction methods:
# bgsub: subtract a constant noise value from every gene
# qq: models and subtracts background noise relative to gene probe count
# none: no background correction
bgcorrect_method = "qq"
# kernel density estimation is used to model a smooth noise distribution
# from the small set of background noise probes. modifying the bandwidth 
# parameter will affect how "smooth" the background distribution model is
bgcorrect_bw_adjust = 2
# ensure that gene probes below this quantile in the background noise 
# distribution are zero. A value of 0.5 means that the gene is 
# indistinguishable from background (50% likelihood of being noise)
bgcorrect_bg_quant = 0.5
#
# normalization method options:
# qn: quantile normalization (recommended)
# rle: deseq2 normalization
# cpm: counts per million
# q3: upper quartile normalization
# tmm: edgeR TMM normalization
# none: no normalization
norm_method = "q3"

```

# stgeomx

## Read, process, and merge SOA geomx data

```{r soa panc geomx data, include=FALSE}

# read data
ds_panc <- stgeomx::read_xlsx(soa_panc_geomx_xlsx)
ds_ln <- stgeomx::read_xlsx(soa_ln_geomx_xlsx)
ds_kidney <- stgeomx::read_xlsx(soa_kidney_geomx_xlsx)
ds_colon <- stgeomx::read_xlsx(soa_colon_geomx_xlsx)
ds_brain <- stgeomx::read_xlsx(soa_brain_geomx_xlsx)
ds <- ds_panc

# read and join sample annotation
annot_panc <- read_excel(soa_panc_geomx_annot_xlsx)
annot_ln <- read_excel(soa_ln_geomx_annot_xlsx)
annot_kidney <- read_excel(soa_kidney_geomx_annot_xlsx)
annot_colon <- read_excel(soa_colon_annot_xlsx)
annot_brain <- read_excel(soa_brain_geomx_annot_xlsx)
annot <- annot_panc

# combine samples and annotations
ds$samples <- ds$samples %>%
  inner_join(annot, by="SegmentDisplayName")
ds_panc$samples <- ds_panc$samples %>%
  inner_join(annot_panc, by="SegmentDisplayName")
ds_ln$samples <- ds_ln$samples %>%
  inner_join(annot_ln, by="SegmentDisplayName")
ds_kidney$samples <- ds_kidney$samples %>%
  inner_join(annot_kidney, by="SegmentDisplayName")
ds_colon$samples <- ds_colon$samples %>%
  inner_join(annot_colon, by="SegmentDisplayName")
ds_brain$samples <- ds_brain$samples %>%
  inner_join(annot_brain, by="SegmentDisplayName")

# add study fields to aoi
ds <- add_study_field(ds, "pancreas")
ds_panc <- add_study_field(ds_panc, "pancreas")
ds_ln <- add_study_field(ds_ln, "ln")
ds_kidney <- add_study_field(ds_kidney, "kidney")
ds_colon <- add_study_field(ds_colon, "colon")
ds_brain <- add_study_field(ds_brain, "brain")

# merge datasets (all annotated data)
ds <- stgeomx::merge_ds(ds, ds_ln)
ds <- stgeomx::merge_ds(ds, ds_colon)
ds <- stgeomx::merge_ds(ds, ds_brain)
ds <- stgeomx::merge_ds(ds, ds_kidney)

# process dataset
ds_panc <- stgeomx::preprocess(ds_panc, geomx_negprobe_lod_quantile)
ds_ln <- stgeomx::preprocess(ds_ln, geomx_negprobe_lod_quantile)
ds_colon <- stgeomx::preprocess(ds_colon, geomx_negprobe_lod_quantile)
ds_kidney <- stgeomx::preprocess(ds_kidney, geomx_negprobe_lod_quantile)
ds_brain <- stgeomx::preprocess(ds_brain, geomx_negprobe_lod_quantile)
ds <- stgeomx::preprocess(ds, geomx_negprobe_lod_quantile)

# merge probes when there is >1 probe per gene
ds_panc <- stgeomx::merge_probes(ds_panc)
ds_ln <- stgeomx::merge_probes(ds_ln)
ds_colon <- stgeomx::merge_probes(ds_colon)
ds_kidney <- stgeomx::merge_probes(ds_kidney)
ds_brain <- stgeomx::merge_probes(ds_brain)
ds <- stgeomx::merge_probes(ds)

# set filtering thresholds
ds_panc <- stgeomx::set_thresholds(ds_panc,
                              min_counts = geomx_min_counts,
                              min_auc = geomx_min_auc,
                              aoi_min_frac_expr = geomx_aoi_min_frac_expr,
                              gene_min_frac_expr = geomx_gene_min_frac_expr)
ds_ln <- stgeomx::set_thresholds(ds_ln,
                              min_counts = geomx_min_counts,
                              min_auc = geomx_min_auc,
                              aoi_min_frac_expr = geomx_aoi_min_frac_expr,
                              gene_min_frac_expr = geomx_gene_min_frac_expr)
ds_colon <- stgeomx::set_thresholds(ds_colon,
                              min_counts = geomx_min_counts,
                              min_auc = geomx_min_auc,
                              aoi_min_frac_expr = geomx_aoi_min_frac_expr,
                              gene_min_frac_expr = geomx_gene_min_frac_expr)
ds_kidney <- stgeomx::set_thresholds(ds_kidney,
                              min_counts = geomx_min_counts,
                              min_auc = geomx_min_auc,
                              aoi_min_frac_expr = geomx_aoi_min_frac_expr,
                              gene_min_frac_expr = geomx_gene_min_frac_expr)
ds_brain <- stgeomx::set_thresholds(ds_brain,
                              min_counts = geomx_min_counts,
                              min_auc = geomx_min_auc,
                              aoi_min_frac_expr = geomx_aoi_min_frac_expr,
                              gene_min_frac_expr = geomx_gene_min_frac_expr)
ds <- stgeomx::set_thresholds(ds,
                              min_counts = geomx_min_counts,
                              min_auc = geomx_min_auc,
                              aoi_min_frac_expr = geomx_aoi_min_frac_expr,
                              gene_min_frac_expr = geomx_gene_min_frac_expr)

```

## QC Figures: Nucleic count and surface area vs background noise
```{r nucleic count and surface area v bg}
bg_correlations_dir <- file.path(plot_dir, "bg_correlations")
p <- ggplot(ds$samples, aes(x=.data$AOISurfaceArea, y=.data$bg_cpm, color=.data$study)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  scale_x_log10() +
  xlab("Area of Sample") +
  ylab("Background Noise")
p

w <- 9
h <- 4
f <- file.path(bg_correlations_dir, "surface_area_bg")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

p <- ggplot(ds$samples, aes(x=.data$AOINucleiCount, y=.data$bg_cpm, color=.data$study)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  xlab("Area of Sample") +
  ylab("Background Noise")
p

w <- 9
h <- 4
f <- file.path(bg_correlations_dir, "nuclei_count_bg")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)
```

## QC Figures: AOI filtering, results stratified by different parameters
```{r qc filter aoi, stratification}
ds_func <- ds_kidney

width <- 12
height <- 8

# bg_auc
p <- ggplot(ds_func$samples, aes(x=patient, y=bg_auc, color=subset)) +
  geom_boxplot(outlier.shape = NA, alpha=0.5) +
  geom_point(aes(group=path_aoi, fill=path_aoi), pch=21, color="black", size=2, alpha=0.7,
             position = position_jitterdodge(jitter.width=0.05, dodge.width=0.5)) +
  geom_hline(yintercept=0.5) +
  geom_hline(yintercept=geomx_min_auc, linetype="dashed", color="red") +
  scale_color_manual(values = pals::cols25()) +
  scale_fill_manual(values = pals::cols25()) +
  facet_wrap(~study, scales="free_x", nrow=1) +
  theme_bw() +
  theme(axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, hjust=1)) +
  theme(legend.position = "bottom")
p
f <- file.path(plot_dir, "/updated_qc/qc_study_patient_bg_auc")
ggsave(paste0(f, ".pdf"), plot=p, width=width, height=height)
ggsave(paste0(f, ".png"), plot=p, width=width, height=height)

# num_counts
p <- ggplot(ds_func$samples, aes(x=patient, y=num_counts+1, color=subset)) +
  geom_boxplot(outlier.shape = NA, alpha=0.5) +
  geom_point(aes(group=path_aoi, fill=path_aoi), pch=21, color="black", size=2, alpha=0.7,
             position = position_jitterdodge(jitter.width=0.05, dodge.width=0.5)) +
  geom_hline(yintercept=geomx_min_counts, linetype="dashed", color="red") +
  scale_y_log10() +
  scale_color_manual(values = pals::cols25()) +
  scale_fill_manual(values = pals::cols25()) +
  facet_wrap(~study, scales="free_x", nrow=1) +
  theme_bw() +
  theme(axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, hjust=1)) +
  theme(legend.position = "bottom")
p
f <- file.path(plot_dir, "/updated_qc/qc_study_patient_counts")
ggsave(paste0(f, ".pdf"), plot=p, width=width, height=height)
ggsave(paste0(f, ".png"), plot=p, width=width, height=height)

# counts across histopathology
p <- ggplot(ds_func$samples, aes(x=subset, y=num_counts+1, fill=path)) +
  geom_boxplot(outlier.shape = NA, alpha=0.5) +
  geom_point(pch=21, color="black", size=2, alpha=0.7,
             position = position_jitterdodge(jitter.width=0.05, dodge.width=0.5)) +
  geom_hline(yintercept=geomx_min_counts, linetype="dashed", color="red") +
  scale_y_log10() +
  scale_color_manual(values = pals::cols25()) +
  facet_wrap(~study, scales="free_x", nrow=1) +
  theme_bw() +
  theme(axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, hjust=1)) +
  theme(legend.position = "none") 
p
f <- file.path(plot_dir, "/updated_qc/qc_path_counts")
ggsave(paste0(f, ".pdf"), plot=p, width=10, height=5)
ggsave(paste0(f, ".png"), plot=p, width=10, height=5)

# bg_auc across histopathology
p <- ggplot(ds_func$samples, aes(x=subset, y=bg_auc, fill=path)) +
  geom_boxplot(outlier.shape = NA, alpha=0.5) +
  geom_point(pch=21, color="black", size=2, alpha=0.7,
             position = position_jitterdodge(jitter.width=0.05, dodge.width=0.5)) +
  geom_hline(yintercept=geomx_min_auc, linetype="dashed", color="red") +
  scale_color_manual(values = pals::cols25()) +
  facet_wrap(~study, scales="free_x", nrow=1) +
  theme_bw() +
  theme(axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, hjust=1)) +
  theme(legend.position = "none") 
p
f <- file.path(plot_dir, "/updated_qc/qc_path_bg_auc")
ggsave(paste0(f, ".pdf"), plot=p, width=10, height=5)
ggsave(paste0(f, ".png"), plot=p, width=10, height=5)

# counts vs auc: subset
p <- ggplot(ds_func$samples, aes(x=num_counts+1, y=bg_auc)) +
  geom_point(aes(color=subset)) +
  geom_vline(xintercept=geomx_min_counts, linetype="dashed", color="red") +
  geom_hline(yintercept=geomx_min_auc, linetype="dashed", color="red") +
  scale_x_log10() +
  scale_color_manual(values = pals::cols25()) +
  theme_bw() +
  facet_wrap(~study)
p
f <- file.path(plot_dir, "/updated_qc/qc_scatter_num_counts_vs_auc_study_subset.pdf")
ggsave(paste0(f, ".pdf"), plot=p, width=8, height=6)
ggsave(paste0(f, ".png"), plot=p, width=8, height=6)

# counts vs auc: region
p <- ggplot(ds_func$samples, aes(x=num_counts+1, y=bg_auc)) +
  geom_point(aes(color=region)) +
  geom_vline(xintercept=geomx_min_counts, linetype="dashed", color="red") +
  geom_hline(yintercept=geomx_min_auc, linetype="dashed", color="red") +
  scale_x_log10() +
  scale_color_manual(values = pals::cols25()) +
  theme_bw() +
  facet_wrap(~study)
p
f <- file.path(plot_dir, "/updated_qc/qc_scatter_num_counts_vs_auc_study_region.pdf")
ggsave(paste0(f, ".pdf"), plot=p, width=8, height=6)
ggsave(paste0(f, ".png"), plot=p, width=8, height=6)

# counts vs auc: slide (sample)
p <- ggplot(ds_func$samples, aes(x=num_counts+1, y=bg_auc)) +
  geom_point(aes(color=slide)) +
  geom_vline(xintercept=geomx_min_counts, linetype="dashed", color="red") +
  geom_hline(yintercept=geomx_min_auc, linetype="dashed", color="red") +
  scale_x_log10() +
  scale_color_manual(values = pals::cols25()) +
  theme_bw() +
  facet_wrap(~study)
p
f <- file.path(plot_dir, "/updated_qc/qc_scatter_num_counts_vs_auc_study_sample.pdf")
ggsave(paste0(f, ".pdf"), plot=p, width=8, height=6)
ggsave(paste0(f, ".png"), plot=p, width=8, height=6)

```

## QC Figures: AOI filtering plots

```{r qc filter aoi}
aoi_filtering_dir <- file.path(plot_dir, "aoi_filtering")

p_merged <- stgeomx::plot_aoi_filter(ds, 
                              min_counts=geomx_min_counts,
                              min_auc=geomx_min_auc,
                              min_frac_expr=geomx_aoi_min_frac_expr)
p_merged
w <- 9
h <- 4
f <- file.path(aoi_filtering_dir, "qc_aoi_filtering_merged")
ggsave(paste0(f, ".png"), p_merged, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_merged, width=w, height=h)

p_panc <- stgeomx::plot_aoi_filter(ds_panc, 
                              min_counts=geomx_min_counts,
                              min_auc=geomx_min_auc,
                              min_frac_expr=geomx_aoi_min_frac_expr)
p_panc
w <- 9
h <- 4
f <- file.path(aoi_filtering_dir, "qc_aoi_filtering_panc")
ggsave(paste0(f, ".png"), p_panc, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_panc, width=w, height=h)

p_ln <- stgeomx::plot_aoi_filter(ds_ln, 
                              min_counts=geomx_min_counts,
                              min_auc=geomx_min_auc,
                              min_frac_expr=geomx_aoi_min_frac_expr)
p_ln
w <- 9
h <- 4
f <- file.path(aoi_filtering_dir, "qc_aoi_filtering_ln")
ggsave(paste0(f, ".png"), p_ln, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_ln, width=w, height=h)

p_kidney <- stgeomx::plot_aoi_filter(ds_kidney, 
                              min_counts=geomx_min_counts,
                              min_auc=geomx_min_auc,
                              min_frac_expr=geomx_aoi_min_frac_expr)
p_kidney
w <- 9
h <- 4
f <- file.path(aoi_filtering_dir, "qc_aoi_filtering_kidney")
ggsave(paste0(f, ".png"), p_kidney, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_kidney, width=w, height=h)

p_colon <- stgeomx::plot_aoi_filter(ds_colon, 
                              min_counts=geomx_min_counts,
                              min_auc=geomx_min_auc,
                              min_frac_expr=geomx_aoi_min_frac_expr)
p_colon
w <- 9
h <- 4
f <- file.path(aoi_filtering_dir, "qc_aoi_filtering_colon")
ggsave(paste0(f, ".png"), p_colon, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_colon, width=w, height=h)

p_brain <- stgeomx::plot_aoi_filter(ds_brain, 
                              min_counts=geomx_min_counts,
                              min_auc=geomx_min_auc,
                              min_frac_expr=geomx_aoi_min_frac_expr)
p_brain
w <- 9
h <- 4
f <- file.path(aoi_filtering_dir, "qc_aoi_filtering_brain")
ggsave(paste0(f, ".png"), p_brain, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_brain, width=w, height=h)

```

## QC Figures: Gene Filtering plots

```{r gene filtering}
gene_filtering_dir <- file.path(plot_dir, "gene_filtering")

p_merged <- stgeomx::plot_gene_filter(ds, geomx_gene_min_frac_expr)
p_merged
w <- 7
h <- 7
f <- file.path(gene_filtering_dir, "qc_gene_filtering_merged")
ggsave(paste0(f, ".png"), p_merged, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_merged, width=w, height=h)

p_panc <- stgeomx::plot_gene_filter(ds_panc, geomx_gene_min_frac_expr)
p_panc
w <- 7
h <- 7
f <- file.path(gene_filtering_dir, "qc_gene_filtering_panc")
ggsave(paste0(f, ".png"), p_panc, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_panc, width=w, height=h)

p_ln <- stgeomx::plot_gene_filter(ds_ln, geomx_gene_min_frac_expr)
p_ln
w <- 7
h <- 7
f <- file.path(gene_filtering_dir, "qc_gene_filtering_ln")
ggsave(paste0(f, ".png"), p_ln, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_ln, width=w, height=h)

p_colon <- stgeomx::plot_gene_filter(ds_colon, geomx_gene_min_frac_expr)
p_colon
w <- 7
h <- 7
f <- file.path(gene_filtering_dir, "qc_gene_filtering_colon")
ggsave(paste0(f, ".png"), p_colon, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_colon, width=w, height=h)

p_brain <- stgeomx::plot_gene_filter(ds_brain, geomx_gene_min_frac_expr)
p_brain
w <- 7
h <- 7
f <- file.path(gene_filtering_dir, "qc_gene_filtering_brain")
ggsave(paste0(f, ".png"), p_brain, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_brain, width=w, height=h)

p_kidney <- stgeomx::plot_gene_filter(ds_kidney, geomx_gene_min_frac_expr)
p_kidney
w <- 7
h <- 7
f <- file.path(gene_filtering_dir, "qc_gene_filtering_kidney")
ggsave(paste0(f, ".png"), p_kidney, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_kidney, width=w, height=h)

```

## Apply filters, then BG correction and normalization

```{r background correction}

correction_normalization_dir <- file.path(plot_dir, "bgcorrection_normalization")

# MERGED
# background correction must use original unfiltered matrix
fds <- stgeomx::bgcorrect(ds, method=bgcorrect_method, 
                          bw.adjust=bgcorrect_bw_adjust, 
                          bg.quant=bgcorrect_bg_quant)
# normalize
fds <- stgeomx::normalize(fds, norm_method)

# plot expression distribution
p <- stgeomx::plot_expr_dist(fds)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_expr_dist_merged")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect2(fds)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_merged")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# background correction must use original unfiltered matrix
fds <- stgeomx::bgcorrect(ds, method=bgcorrect_method,
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
# apply filters before normalizing
fds <- stgeomx::apply_filters(fds)
# normalize
fds <- stgeomx::normalize(fds, norm_method)

# filtering stats
nrow(ds$samples)
nrow(fds$samples)
table(ds$meta$keep)

# final normalized gene expression matrix
log2_norm_cpm_merged <- log2(fds$x + 1)

# PANCREAS
# background correction must use original unfiltered matrix
fds_panc <- stgeomx::bgcorrect(ds_panc, method=bgcorrect_method, 
                          bw.adjust=bgcorrect_bw_adjust, 
                          bg.quant=bgcorrect_bg_quant)
# normalize
fds_panc <- stgeomx::normalize(fds_panc, norm_method)

# plot expression distribution
p <- stgeomx::plot_expr_dist(fds_panc)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_expr_dist_panc")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect2(fds_panc)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_panc")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# background correction must use original unfiltered matrix
fds_panc <- stgeomx::bgcorrect(ds_panc, method=bgcorrect_method,
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
# apply filters before normalizing
fds_panc <- stgeomx::apply_filters(fds_panc)
# normalize
fds_panc <- stgeomx::normalize(fds_panc)

# filtering stats
nrow(ds$samples)
nrow(fds_panc$samples)
table(ds$meta$keep)

# final normalized gene expression matrix
log2_norm_cpm_panc <- log2(fds_panc$x + 1)

# LYMPH NODE
# background correction must use original unfiltered matrix
fds_ln <- stgeomx::bgcorrect(ds_ln, method=bgcorrect_method, 
                          bw.adjust=bgcorrect_bw_adjust, 
                          bg.quant=bgcorrect_bg_quant)
# normalize
fds_ln <- stgeomx::normalize(fds_ln, norm_method)

# plot expression distribution
p <- stgeomx::plot_expr_dist(fds_ln)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_expr_dist_ln")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect2(fds_ln)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_ln")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# background correction must use original unfiltered matrix
fds_ln <- stgeomx::bgcorrect(ds_ln, method=bgcorrect_method,
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
# apply filters before normalizing
fds_ln <- stgeomx::apply_filters(fds_ln)
# normalize
fds_ln <- stgeomx::normalize(fds_ln, norm_method)

# filtering stats
nrow(ds$samples)
nrow(fds_ln$samples)
table(ds$meta$keep)

# final normalized gene expression matrix
log2_norm_cpm_ln <- log2(fds_ln$x + 1)

# KIDNEY
# background correction must use original unfiltered matrix
fds_kidney <- stgeomx::bgcorrect(ds_kidney, method=bgcorrect_method, 
                          bw.adjust=bgcorrect_bw_adjust, 
                          bg.quant=bgcorrect_bg_quant)
# normalize
fds_kidney <- stgeomx::normalize(fds_kidney, norm_method)

# plot expression distribution
p <- stgeomx::plot_expr_dist(fds_kidney)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_expr_dist_kidney")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect2(fds_kidney)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_kidney")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# background correction must use original unfiltered matrix
fds_kidney <- stgeomx::bgcorrect(ds_kidney, method=bgcorrect_method,
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
# apply filters before normalizing
fds_kidney <- stgeomx::apply_filters(fds_kidney)
# normalize
fds_kidney <- stgeomx::normalize(fds_kidney, norm_method)

# filtering stats
nrow(ds$samples)
nrow(fds_kidney$samples)
table(ds$meta$keep)

# final normalized gene expression matrix
log2_norm_cpm_kidney <- log2(fds_kidney$x + 1)

# COLON
# background correction must use original unfiltered matrix
fds_colon <- stgeomx::bgcorrect(ds_colon, method=bgcorrect_method, 
                          bw.adjust=bgcorrect_bw_adjust, 
                          bg.quant=bgcorrect_bg_quant)
# normalize
fds_colon <- stgeomx::normalize(fds_colon, norm_method)

# plot expression distribution
p <- stgeomx::plot_expr_dist(fds_colon)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_expr_dist_colon")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect2(fds_colon)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_colon")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# background correction must use original unfiltered matrix
fds_colon <- stgeomx::bgcorrect(ds_colon, method=bgcorrect_method,
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
# apply filters before normalizing
fds_colon <- stgeomx::apply_filters(fds_colon)
# normalize
fds_colon <- stgeomx::normalize(fds_colon, norm_method)

# filtering stats
nrow(ds$samples)
nrow(fds_colon$samples)
table(ds$meta$keep)

# final normalized gene expression matrix
log2_norm_cpm_colon <- log2(fds_colon$x + 1)

# BRAIN
# background correction must use original unfiltered matrix
fds_brain <- stgeomx::bgcorrect(ds_brain, method=bgcorrect_method, 
                          bw.adjust=bgcorrect_bw_adjust, 
                          bg.quant=bgcorrect_bg_quant)
# normalize
fds_brain <- stgeomx::normalize(fds_brain, norm_method)

# plot expression distribution
p <- stgeomx::plot_expr_dist(fds_brain)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_expr_dist_brain")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect2(fds_brain)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_brain")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# background correction must use original unfiltered matrix
fds_brain <- stgeomx::bgcorrect(ds_brain, method=bgcorrect_method,
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
# apply filters before normalizing
fds_brain <- stgeomx::apply_filters(fds_brain)
# normalize
fds_brain <- stgeomx::normalize(fds_brain, norm_method)

# filtering stats
nrow(ds$samples)
nrow(fds$samples)
table(ds$meta$keep)

# final normalized gene expression matrix
log2_norm_cpm_brain <- log2(fds_brain$x + 1)

```
# Manuscript figure generation

## Helper function that creates plots for figure 2 and 3
```{r helper, include=FALSE}
cpm <- function(x) {
  sweep(x, 2, colSums(x) / 1e6, FUN="/")
}

plot_bgcorrect_helper <- function(ds, quantiles=c(0.10, 0.25, 0.5, 0.75, 0.90, 0.95, 0.99)) {
  # gene metadata
  bg <- ds$meta$bg
  keep <- ds$meta$keep
  # labels for plot legend
  quant_names <- paste0("q", round(100 * quantiles))
  # sample columns used for plot
  s <- select(ds$samples, "aoi", "slide", "keep")
  # compute background level per aoi
  s$bg_cpm <- 1e6 * colSums(ds$counts[bg,]) / colSums(ds$counts)

  # filter genes
  x <- ds$x[(!bg) & keep,]
  counts <- ds$counts[(!bg) & keep,]

  # gene cpm before bgcorrect
  xgene <- as_tibble(cpm(counts))
  qgene <- xgene %>%
    reframe(across(everything(), ~ stats::quantile(.x, quantiles))) %>%
    mutate(name="gene", quant=quant_names)

  # cpm after bgcorrect
  y <- as_tibble(cpm(x))
  qy <- y %>%
    reframe(across(everything(), ~ stats::quantile(.x, quantiles))) %>%
    mutate(name="y", quant=quant_names)

  # combine bg and gene quantiles
  q <- bind_rows(qgene, qy) %>%
    tidyr::pivot_longer(colnames(x), names_to="aoi", values_to="value") %>%
    tidyr::pivot_wider(names_from="name", values_from="value", names_prefix="") %>%
    inner_join(s, by=c("aoi"="aoi"), suffix=c("_gene", "_aoi"))

  p <- ggplot(filter(q, .data$keep), aes(x=.data$bg_cpm+1, y=.data$y+1, color=.data$quant)) +
    geom_point(alpha=0.6) +
    geom_smooth(data=filter(q, .data$keep), linetype='dashed', method = 'lm',
                formula = y ~ x, se=FALSE, alpha=0.6) +
    scale_color_viridis_d(option="plasma") +
    theme_minimal() +
    labs(x="Noise (CPM)", y="Normalized Counts (CPM)")
  return(p)
}

most_var_genes <- function(x, ntop=500, span=0.50) {
  tbl <- bind_cols(
    u = rowMeans(x),
    v = apply(x, 1, function(x) var(x))
  )
  lovar <- loess(v ~ u, data=tbl, span=span)
  tbl$lovar <- predict(lovar)
  tbl$vadj <- tbl$v / tbl$lovar
  keep <- order(tbl$vadj, decreasing=TRUE)[seq_len(min(ntop, nrow(tbl)))]
  return(x[keep,])
}

pca <- function(s, y) {
  # PCA
  pca_res <- prcomp(t(y), center=TRUE, scale.=FALSE)
  pca_res.var <- round(100 * pca_res$sdev^2 / sum( pca_res$sdev^2 ), 2)
  pca_res$var_txt <- paste(colnames(pca_res$x), " (", paste(as.character(pca_res.var), "%", ")", sep=""), sep="")

  return(pca_res)
}

plot_pca <- function(pca_res, x) {
  # plot dimensions
  width <- 12
  height <- 10

  # creation of individual PCA plots
  p <- ggplot(x, aes(x=PC1, y=PC2, color=study)) +
    geom_point(alpha = 0.4, size=1) +
    theme_bw() +
    theme(aspect.ratio=1) +
    theme(legend.text = element_text(size=7)) +
    labs(x=pca_res$var_txt[1], y=pca_res$var_txt[2])
  return(p)
}

plot_pca_bg <- function(pca_res, x) {
  # plot dimensions
  width <- 12
  height <- 10

  # creation of individual PCA plots
  p <- ggplot(x, aes(x=PC1, y=PC2, color=bg_auc)) +
    geom_point(alpha = 0.4, size=1) +
    scale_color_viridis_c() +
    theme_bw() +
    theme(aspect.ratio=1) +
    theme(legend.text = element_text(size=7)) +
    labs(x=pca_res$var_txt[1], y=pca_res$var_txt[2])
  return(p)
}

cluster_pca <- function(pca_res) {
# npcs
npcs <- 3
nclusters <- 5

# cluster PCA results
# pam
x <- pca_res$x[,1:npcs]
pamx <- cluster::pam(dist(x), k=nclusters)
memb <- paste0("c", pamx$clustering)

return(memb)
}

pca_plot_ipmn <- function(fds, pca_res, xvar, yvar, color_by, shape_by) {
  x <- bind_cols(fds$samples, as_tibble(pca_res$x))
  color_by_string <- rlang::as_string(rlang::ensym(color_by))
  shape_by_string <- rlang::as_string(rlang::ensym(shape_by))
  p <- ggplot(x, aes(x={{xvar}}, y={{yvar}}, color={{color_by}}, shape={{shape_by}})) +
    geom_point(alpha = 0.6, size=1) +
    # scale_color_manual(values = color_scales[[color_by_string]]) +
    # scale_shape_manual(values = shape_scales[[shape_by_string]]) +
    coord_fixed() +
    theme_bw()
  return(p)
}
```

## FIGURE 2: effects of bg correction and normalization on merged only
```{r bg correction and normalization methods plot}
correction_normalization_dir <- file.path(plot_dir, "bgcorrection_normalization")

# background correction must use original unfiltered matrix
fds_bgsub <- stgeomx::bgcorrect(ds, method="bgsub",
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
fds_none <- stgeomx::bgcorrect(ds, method="none",
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
fds_qq <- stgeomx::bgcorrect(ds, method="qq",
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)

# normalize
fds_bgsub_qn <- stgeomx::normalize(fds_bgsub, "qn")
fds_bgsub_rle <- stgeomx::normalize(fds_bgsub, "rle")
fds_bgsub_q3 <- stgeomx::normalize(fds_bgsub, "q3")
fds_bgsub_tmm <- stgeomx::normalize(fds_bgsub, "tmm")

fds_qq_qn <- stgeomx::normalize(fds_qq, "qn")
fds_qq_rle <- stgeomx::normalize(fds_qq, "rle")
fds_qq_q3 <- stgeomx::normalize(fds_qq, "q3")
fds_qq_tmm <- stgeomx::normalize(fds_qq, "tmm")

fds_none_qn <- stgeomx::normalize(fds_none, "qn")
fds_none_rle <- stgeomx::normalize(fds_none, "rle")
fds_none_q3 <- stgeomx::normalize(fds_none, "q3")
fds_none_tmm <- stgeomx::normalize(fds_none, "tmm")

# plot background correction
plot_bgsub_qn <- plot_bgcorrect_helper(fds_bgsub_qn)
plot_bgsub_rle <- plot_bgcorrect_helper(fds_bgsub_rle)
plot_bgsub_q3 <- plot_bgcorrect_helper(fds_bgsub_q3)
plot_bgsub_tmm <- plot_bgcorrect_helper(fds_bgsub_tmm)

plot_qq_qn <- plot_bgcorrect_helper(fds_qq_qn)
plot_qq_rle <- plot_bgcorrect_helper(fds_qq_rle)
plot_qq_q3 <- plot_bgcorrect_helper(fds_qq_q3)
plot_qq_tmm <- plot_bgcorrect_helper(fds_qq_tmm)

plot_none_qn <- plot_bgcorrect_helper(fds_none_qn)
plot_none_rle <- plot_bgcorrect_helper(fds_none_rle)
plot_none_q3 <- plot_bgcorrect_helper(fds_none_q3)
plot_none_tmm <- plot_bgcorrect_helper(fds_none_tmm)

# Column and row titles
col_titles <- list(
  wrap_elements(grid::textGrob('', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('Q3', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('TMM', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('RLE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QN', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

row_titles <- list(
  wrap_elements(grid::textGrob('NONE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('BGSub', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QQ', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

# Create grid
p <- wrap_plots(
  col_titles[[1]], col_titles[[2]], col_titles[[3]], col_titles[[4]], col_titles[[5]],
  row_titles[[1]], plot_none_q3, plot_none_tmm, plot_none_rle, plot_none_qn,
  row_titles[[2]], plot_bgsub_q3, plot_bgsub_tmm, plot_bgsub_rle, plot_bgsub_qn,
  row_titles[[3]], plot_qq_q3, plot_qq_tmm, plot_qq_rle, plot_qq_qn,
  ncol = 5,
  heights = c(0.1, 1, 1, 1),
  widths = c(0.3, 1, 1, 1, 1)
) +
plot_layout(
  guides = 'collect'
) &
theme(plot.margin = unit(c(0.5, 1, 1, 1), "cm"))

# Add titles
p <- p +
  plot_annotation(
    title = "Effect of Background Correction and Normalization Methods on Normalized Expression Levels"
  ) &
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5)
  )

f <- file.path(correction_normalization_dir, "bgcorrect_normalization_grid")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)
```

## Setup for downstream analysis
```{r further processing}
# background correction must use original unfiltered matrix
fds_none <- stgeomx::bgcorrect(ds, method="none",
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
fds_bgsub <- stgeomx::bgcorrect(ds, method="bgsub",
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
fds_qq <- stgeomx::bgcorrect(ds, method="qq",
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)

# apply filters before normalizing
fds_none <- stgeomx::apply_filters(fds_none)
fds_bgsub <- stgeomx::apply_filters(fds_bgsub)
fds_qq <- stgeomx::apply_filters(fds_qq)

# normalize
fds_none_q3 <- stgeomx::normalize(fds_none, "q3")
fds_none_tmm <- stgeomx::normalize(fds_none, "tmm")
fds_none_rle <- stgeomx::normalize(fds_none, "rle")
fds_none_qn <- stgeomx::normalize(fds_none, "qn")

fds_bgsub_q3 <- stgeomx::normalize(fds_bgsub, "q3")
fds_bgsub_tmm <- stgeomx::normalize(fds_bgsub, "tmm")
fds_bgsub_rle <- stgeomx::normalize(fds_bgsub, "rle")
fds_bgsub_qn <- stgeomx::normalize(fds_bgsub, "qn")

fds_qq_q3 <- stgeomx::normalize(fds_qq, "q3")
fds_qq_tmm <- stgeomx::normalize(fds_qq, "tmm")
fds_qq_rle <- stgeomx::normalize(fds_qq, "rle")
fds_qq_qn <- stgeomx::normalize(fds_qq, "qn")

# final normalized gene expression matrix
log2_norm_cpm_none_q3 <- log2(fds_none_q3$x + 1)
log2_norm_cpm_none_tmm <- log2(fds_none_tmm$x + 1)
log2_norm_cpm_none_rle <- log2(fds_none_rle$x + 1)
log2_norm_cpm_none_qn <- log2(fds_none_qn$x + 1)

log2_norm_cpm_bgsub_q3 <- log2(fds_bgsub_q3$x + 1)
log2_norm_cpm_bgsub_tmm <- log2(fds_bgsub_tmm$x + 1)
log2_norm_cpm_bgsub_rle <- log2(fds_bgsub_rle$x + 1)
log2_norm_cpm_bgsub_qn <- log2(fds_bgsub_qn$x + 1)

log2_norm_cpm_qq_q3 <- log2(fds_qq_q3$x + 1)
log2_norm_cpm_qq_tmm <- log2(fds_qq_tmm$x + 1)
log2_norm_cpm_qq_rle <- log2(fds_qq_rle$x + 1)
log2_norm_cpm_qq_qn <- log2(fds_qq_qn$x + 1)
```

## FIGURE 3: PCA grid
``` {r pca}
pca_dir <- file.path(plot_dir, "pca")

# subset most variable genes
ntop <- 500
myspan <- 0.5

# most var genes
s_none_q3 <- fds_none_q3$samples
y_none_q3 <- most_var_genes(log2_norm_cpm_none_q3, ntop=ntop, span=myspan)
s_none_tmm <- fds_none_tmm$samples
y_none_tmm <- most_var_genes(log2_norm_cpm_none_tmm, ntop=ntop, span=myspan)
s_none_rle <- fds_none_rle$samples
y_none_rle <- most_var_genes(log2_norm_cpm_none_rle, ntop=ntop, span=myspan)
s_none_qn <- fds_none_qn$samples
y_none_qn <- most_var_genes(log2_norm_cpm_none_qn, ntop=ntop, span=myspan)

s_bgsub_q3 <- fds_bgsub_q3$samples
y_bgsub_q3 <- most_var_genes(log2_norm_cpm_bgsub_q3, ntop=ntop, span=myspan)
s_bgsub_tmm <- fds_bgsub_tmm$samples
y_bgsub_tmm <- most_var_genes(log2_norm_cpm_bgsub_tmm, ntop=ntop, span=myspan)
s_bgsub_rle <- fds_bgsub_rle$samples
y_bgsub_rle <- most_var_genes(log2_norm_cpm_bgsub_rle, ntop=ntop, span=myspan)
s_bgsub_qn <- fds_bgsub_qn$samples
y_bgsub_qn <- most_var_genes(log2_norm_cpm_bgsub_qn, ntop=ntop, span=myspan)

s_qq_q3 <- fds_qq_q3$samples
y_qq_q3 <- most_var_genes(log2_norm_cpm_qq_q3, ntop=ntop, span=myspan)
s_qq_tmm <- fds_qq_tmm$samples
y_qq_tmm <- most_var_genes(log2_norm_cpm_qq_tmm, ntop=ntop, span=myspan)
s_qq_rle <- fds_qq_rle$samples
y_qq_rle <- most_var_genes(log2_norm_cpm_qq_rle, ntop=ntop, span=myspan)
s_qq_qn <- fds_qq_qn$samples
y_qq_qn <- most_var_genes(log2_norm_cpm_qq_qn, ntop=ntop, span=myspan)

# PCA_res
pca_res_none_q3 <- pca(s_none_q3, y_none_q3)
pca_res_none_tmm <- pca(s_none_tmm, y_none_tmm)
pca_res_none_rle <- pca(s_none_rle, y_none_rle)
pca_res_none_qn <- pca(s_none_qn, y_none_qn)

pca_res_bgsub_q3 <- pca(s_bgsub_q3, y_bgsub_q3)
pca_res_bgsub_tmm <- pca(s_bgsub_tmm, y_bgsub_tmm)
pca_res_bgsub_rle <- pca(s_bgsub_rle, y_bgsub_rle)
pca_res_bgsub_qn <- pca(s_bgsub_qn, y_bgsub_qn)

pca_res_qq_q3 <- pca(s_qq_q3, y_qq_q3)
pca_res_qq_tmm <- pca(s_qq_tmm, y_qq_tmm)
pca_res_qq_rle <- pca(s_qq_rle, y_qq_rle)
pca_res_qq_qn <- pca(s_qq_qn, y_qq_qn)

# x
x_none_q3 <- bind_cols(s_none_q3, as_tibble(pca_res_none_q3$x))
x_none_tmm <- bind_cols(s_none_tmm, as_tibble(pca_res_none_tmm$x))
x_none_rle <- bind_cols(s_none_rle, as_tibble(pca_res_none_rle$x))
x_none_qn <- bind_cols(s_none_qn, as_tibble(pca_res_none_qn$x))

x_bgsub_q3 <- bind_cols(s_bgsub_q3, as_tibble(pca_res_bgsub_q3$x))
x_bgsub_tmm <- bind_cols(s_bgsub_tmm, as_tibble(pca_res_bgsub_tmm$x))
x_bgsub_rle <- bind_cols(s_bgsub_rle, as_tibble(pca_res_bgsub_rle$x))
x_bgsub_qn <- bind_cols(s_bgsub_qn, as_tibble(pca_res_bgsub_qn$x))

x_qq_q3 <- bind_cols(s_qq_q3, as_tibble(pca_res_qq_q3$x))
x_qq_tmm <- bind_cols(s_qq_tmm, as_tibble(pca_res_qq_tmm$x))
x_qq_rle <- bind_cols(s_qq_rle, as_tibble(pca_res_qq_rle$x))
x_qq_qn <- bind_cols(s_qq_qn, as_tibble(pca_res_qq_qn$x))

# individual pca plots
pca_none_q3 <- plot_pca(pca_res_none_q3, x_none_q3)
pca_none_tmm <- plot_pca(pca_res_none_tmm, x_none_tmm)
pca_none_rle <- plot_pca(pca_res_none_q3, x_none_rle)
pca_none_qn <- plot_pca(pca_res_none_q3, x_none_qn)

pca_bgsub_q3 <- plot_pca(pca_res_bgsub_q3, x_bgsub_q3)
pca_bgsub_tmm <- plot_pca(pca_res_bgsub_tmm, x_bgsub_tmm)
pca_bgsub_rle <- plot_pca(pca_res_bgsub_q3, x_bgsub_rle)
pca_bgsub_qn <- plot_pca(pca_res_bgsub_q3, x_bgsub_qn)

pca_qq_q3 <- plot_pca(pca_res_qq_q3, x_qq_q3)
pca_qq_tmm <- plot_pca(pca_res_qq_tmm, x_qq_tmm)
pca_qq_rle <- plot_pca(pca_res_qq_q3, x_qq_rle)
pca_qq_qn <- plot_pca(pca_res_qq_q3, x_qq_qn)

# Grid column and row titles
col_titles <- list(
  wrap_elements(grid::textGrob('', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('Q3', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('TMM', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('RLE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QN', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

row_titles <- list(
  wrap_elements(grid::textGrob('NONE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('BGSub', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QQ', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

# Create grid
p <- wrap_plots(
  col_titles[[1]], col_titles[[2]], col_titles[[3]], col_titles[[4]], col_titles[[5]],
  row_titles[[1]], pca_none_q3, pca_none_tmm, pca_none_rle, pca_none_qn,
  row_titles[[2]], pca_bgsub_q3, pca_bgsub_tmm, pca_bgsub_rle, pca_bgsub_qn,
  row_titles[[3]], pca_qq_q3, pca_qq_tmm, pca_qq_rle, pca_qq_qn,
  ncol = 5,
  heights = c(0.1, 1, 1, 1),
  widths = c(0.3, 1, 1, 1, 1)
) +
plot_layout(
  guides = 'collect'
) &
theme(plot.margin = unit(c(0.5, 1, 1, 1), "cm"))

# Add titles
p <- p + 
  plot_annotation(
    title = "Effect of Background Correction and Normalization Methods on PCA"
  ) & 
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5)
  )

f <- file.path(pca_dir, "pca_grid_2000")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)

```

### FIGURE 3B: Background Noise
``` {r}
# individual pca background plots
pca_none_q3 <- plot_pca_bg(pca_res_none_q3, x_none_q3)
pca_none_tmm <- plot_pca_bg(pca_res_none_tmm, x_none_tmm)
pca_none_rle <- plot_pca_bg(pca_res_none_q3, x_none_rle)
pca_none_qn <- plot_pca_bg(pca_res_none_q3, x_none_qn)

pca_bgsub_q3 <- plot_pca_bg(pca_res_bgsub_q3, x_bgsub_q3)
pca_bgsub_tmm <- plot_pca_bg(pca_res_bgsub_tmm, x_bgsub_tmm)
pca_bgsub_rle <- plot_pca_bg(pca_res_bgsub_q3, x_bgsub_rle)
pca_bgsub_qn <- plot_pca_bg(pca_res_bgsub_q3, x_bgsub_qn)

pca_qq_q3 <- plot_pca_bg(pca_res_qq_q3, x_qq_q3)
pca_qq_tmm <- plot_pca_bg(pca_res_qq_tmm, x_qq_tmm)
pca_qq_rle <- plot_pca_bg(pca_res_qq_q3, x_qq_rle)
pca_qq_qn <- plot_pca_bg(pca_res_qq_q3, x_qq_qn)

# Grid column and row titles
col_titles <- list(
  wrap_elements(grid::textGrob('', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('Q3', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('TMM', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('RLE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QN', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

row_titles <- list(
  wrap_elements(grid::textGrob('NONE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('BGSub', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QQ', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

# Create grid
p <- wrap_plots(
  col_titles[[1]], col_titles[[2]], col_titles[[3]], col_titles[[4]], col_titles[[5]],
  row_titles[[1]], pca_none_q3, pca_none_tmm, pca_none_rle, pca_none_qn,
  row_titles[[2]], pca_bgsub_q3, pca_bgsub_tmm, pca_bgsub_rle, pca_bgsub_qn,
  row_titles[[3]], pca_qq_q3, pca_qq_tmm, pca_qq_rle, pca_qq_qn,
  ncol = 5,
  heights = c(0.1, 1, 1, 1),
  widths = c(0.3, 1, 1, 1, 1)
) +
plot_layout(
  guides = 'collect'
) &
theme(plot.margin = unit(c(0.5, 1, 1, 1), "cm"))

# Add titles
p <- p + 
  plot_annotation(
    title = "Effect of Background Correction and Normalization Methods on PCA"
  ) & 
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5)
  )

# Save graph
f <- file.path(pca_dir, "pca_grid_bg")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)
```

### Pearson correlates
```{r}
# Example data (replace with your actual data)
methods <- c("none_q3", "none_tmm", "none_rle", "none_qn", 
             "bgsub_q3", "bgsub_tmm", "bgsub_rle", "bgsub_qn", 
             "qq_q3", "qq_tmm", "qq_rle", "qq_qn")

# Performing cor.test and storing the result in lists
pc1_data <- list(cor.test(x_none_q3$PC1, x_none_q3$bg_auc),
                 cor.test(x_none_tmm$PC1, x_none_tmm$bg_auc),
                 cor.test(x_none_rle$PC1, x_none_rle$bg_auc),
                 cor.test(x_none_qn$PC1, x_none_qn$bg_auc),
                 cor.test(x_bgsub_q3$PC1, x_bgsub_q3$bg_auc),
                 cor.test(x_bgsub_tmm$PC1, x_bgsub_tmm$bg_auc),
                 cor.test(x_bgsub_rle$PC1, x_bgsub_rle$bg_auc),
                 cor.test(x_bgsub_qn$PC1, x_bgsub_qn$bg_auc),
                 cor.test(x_qq_q3$PC1, x_qq_q3$bg_auc),
                 cor.test(x_qq_tmm$PC1, x_qq_tmm$bg_auc),
                 cor.test(x_qq_rle$PC1, x_qq_rle$bg_auc),
                 cor.test(x_qq_qn$PC1, x_qq_qn$bg_auc))

pc2_data <- list(cor.test(x_none_q3$PC2, x_none_q3$bg_auc),
                 cor.test(x_none_tmm$PC2, x_none_tmm$bg_auc),
                 cor.test(x_none_rle$PC2, x_none_rle$bg_auc),
                 cor.test(x_none_qn$PC2, x_none_qn$bg_auc),
                 cor.test(x_bgsub_q3$PC2, x_bgsub_q3$bg_auc),
                 cor.test(x_bgsub_tmm$PC2, x_bgsub_tmm$bg_auc),
                 cor.test(x_bgsub_rle$PC2, x_bgsub_rle$bg_auc),
                 cor.test(x_bgsub_qn$PC2, x_bgsub_qn$bg_auc),
                 cor.test(x_qq_q3$PC2, x_qq_q3$bg_auc),
                 cor.test(x_qq_tmm$PC2, x_qq_tmm$bg_auc),
                 cor.test(x_qq_rle$PC2, x_qq_rle$bg_auc),
                 cor.test(x_qq_qn$PC2, x_qq_qn$bg_auc))

pc3_data <- list(cor.test(x_none_q3$PC3, x_none_q3$bg_auc),
                 cor.test(x_none_tmm$PC3, x_none_tmm$bg_auc),
                 cor.test(x_none_rle$PC3, x_none_rle$bg_auc),
                 cor.test(x_none_qn$PC3, x_none_qn$bg_auc),
                 cor.test(x_bgsub_q3$PC3, x_bgsub_q3$bg_auc),
                 cor.test(x_bgsub_tmm$PC3, x_bgsub_tmm$bg_auc),
                 cor.test(x_bgsub_rle$PC3, x_bgsub_rle$bg_auc),
                 cor.test(x_bgsub_qn$PC3, x_bgsub_qn$bg_auc),
                 cor.test(x_qq_q3$PC3, x_qq_q3$bg_auc),
                 cor.test(x_qq_tmm$PC3, x_qq_tmm$bg_auc),
                 cor.test(x_qq_rle$PC3, x_qq_rle$bg_auc),
                 cor.test(x_qq_qn$PC3, x_qq_qn$bg_auc))

# Extract the correlation estimates from cor.test results
pc1_estimates <- sapply(pc1_data, function(test) test$estimate)
pc2_estimates <- sapply(pc2_data, function(test) test$estimate)
pc3_estimates <- sapply(pc3_data, function(test) test$estimate)

# Initialize an empty matrix to store the results
results <- matrix(NA, nrow = length(methods), ncol = 3)
rownames(results) <- methods
colnames(results) <- c("PC1", "PC2", "PC3")

# Populate the results matrix
results[, "PC1"] <- pc1_estimates
results[, "PC2"] <- pc2_estimates
results[, "PC3"] <- pc3_estimates

# Convert to a data frame for better readability
results_df <- as.data.frame(results)

# Print the results
print(results_df)
```

### PC1, PC2, PC3, PC4, PC5 for qq_qn
``` {r pcs}
s <- fds_qq_qn$samples
pcs <- c("PC1", "PC2", "PC3", "PC4", "PC5")
pca_tbl <- pca_res_qq_qn$x
tbl <- as_tibble(pca_tbl[,pcs], rownames="aoi")

x <- tbl %>% pivot_longer(cols=all_of(pcs), names_to="xpc", values_to="xvalue")
y <- tbl %>% pivot_longer(cols=all_of(pcs), names_to="ypc", values_to="yvalue")
tbl <- full_join(x, y, by=join_by(aoi), relationship="many-to-many")
tbl <- left_join(tbl, s, by=join_by(aoi))

custom_theme_pca <- function() {
  theme_minimal() +
  theme(aspect.ratio=1) +
    theme(legend.title = element_text(size=8),
          legend.text = element_text(size=7),
          legend.position = "bottom")
}

p <- ggplot(filter(tbl, xpc < ypc), aes(x=xvalue, y=yvalue, color=region, shape=study)) +
  geom_point(alpha=0.7) +
  custom_theme_pca() +
  facet_grid(ypc ~ xpc)

f <- file.path(pca_dir, "pca_12345_grid")
ggsave(paste0(f, ".png"), p, width=25, height=20)
ggsave(paste0(f, ".pdf"), p, width=25, height=20)
```

### SUPPLEMENTAL: regions on PCA
``` {r PCA regions}
# pancreas region
p <- ggplot(x_qq_qn %>% arrange(study), aes(x=PC1, y=PC2, color = ifelse(study == "pancreas", region, "other"), shape=study)) +
  geom_point(alpha = 0.8, size=3) +
  scale_color_manual(values = c("other" = "grey",
                                "acini" = "red",
                                "acini_and_duct" = "orange",
                                "alpha_cell_enriched_islet" = "green",
                                "beta_cell_enriched_islet" = "blue",
                                "duct" = "yellow",
                                "islet" = "pink")) +
  theme_bw() +
  theme(aspect.ratio=1) +
  theme(legend.text = element_text(size=7)) +
  labs(x=pca_res_qq_qn$var_txt[1], y=pca_res_qq_qn$var_txt[2])
f <- file.path(plot_dir, paste0("pca_region_pancreas.pdf"))
ggsave(f, plot=p, device="pdf", width=10, height=10)

# lymph node and colon regions
p <- ggplot(x_qq_qn, aes(x = PC1, y = PC2, color = ifelse(study %in% c("colon", "ln"), region, "other"), shape = study)) +
  geom_point(alpha = 0.8, size=3) +
  scale_color_manual(values = c("other" = "grey",
                                "b_cell_zone" = "red",
                                "circular_muscle_layer" = "salmon",
                                "connective_tissues" = "orange",
                                "enteroendocrine_cells" = "yellow",
                                "epithelium" = "green",
                                "germinal_center" = "palegreen3",
                                "lamina_propria" = "cyan4",
                                "longitudinal_muscle_layer" = "turquoise1",
                                "lymphoid_structure" = "steelblue",
                                "medulla" = "blue",
                                "muscularis_mucosae" = "darkmagenta",
                                "paneth_cells" = "purple",
                                "t_cell_zone" = "pink",
                                "trabecula" = "deeppink",
                                "vessels" = "tan3")) +
  theme_bw() +
  theme(aspect.ratio=1) +
  theme(legend.text = element_text(size=7)) +
  labs(x=pca_res_qq_qn$var_txt[1], y=pca_res_qq_qn$var_txt[2])
p
f <- file.path(plot_dir, paste0("pca_region_ln_colon.pdf"))
ggsave(f, plot=p, device="pdf", width=20, height=15)
```

## FIGURE 4: Clustering
``` {r clustering}
# the following guides were used in this section:
# guide: https://rpubs.com/KarolinaSzczesna/862710
# guide: https://rpubs.com/Bury/ClusteringOnPcaResults

# cluster, then add cluster membership to sample properties
fds_none_q3$samples$cluster <- cluster_pca(pca_res_none_q3)
fds_none_tmm$samples$cluster <- cluster_pca(pca_res_none_tmm)
fds_none_rle$samples$cluster <- cluster_pca(pca_res_none_rle)
fds_none_qn$samples$cluster <- cluster_pca(pca_res_none_qn)

fds_bgsub_q3$samples$cluster <- cluster_pca(pca_res_bgsub_q3)
fds_bgsub_tmm$samples$cluster <- cluster_pca(pca_res_bgsub_tmm)
fds_bgsub_rle$samples$cluster <- cluster_pca(pca_res_bgsub_rle)
fds_bgsub_qn$samples$cluster <- cluster_pca(pca_res_bgsub_qn)

fds_qq_q3$samples$cluster <- cluster_pca(pca_res_qq_q3)
fds_qq_tmm$samples$cluster <- cluster_pca(pca_res_qq_tmm)
fds_qq_rle$samples$cluster <- cluster_pca(pca_res_qq_rle)
fds_qq_qn$samples$cluster <- cluster_pca(pca_res_qq_qn)

# pca plot clustering
cluster_none_q3 <- pca_plot_ipmn(fds_none_q3, pca_res_none_q3, PC1, PC2, cluster, study)
cluster_none_tmm <- pca_plot_ipmn(fds_none_tmm, pca_res_none_tmm, PC1, PC2, cluster, study)
cluster_none_rle <- pca_plot_ipmn(fds_none_rle, pca_res_none_rle, PC1, PC2, cluster, study)
cluster_none_qn <- pca_plot_ipmn(fds_none_qn, pca_res_none_qn, PC1, PC2, cluster, study)

cluster_bgsub_q3 <- pca_plot_ipmn(fds_bgsub_q3, pca_res_bgsub_q3, PC1, PC2, cluster, study)
cluster_bgsub_tmm <- pca_plot_ipmn(fds_bgsub_tmm, pca_res_bgsub_tmm, PC1, PC2, cluster, study)
cluster_bgsub_rle <- pca_plot_ipmn(fds_bgsub_rle, pca_res_bgsub_rle, PC1, PC2, cluster, study)
cluster_bgsub_qn <- pca_plot_ipmn(fds_bgsub_qn, pca_res_bgsub_qn, PC1, PC2, cluster, study)

cluster_qq_q3 <- pca_plot_ipmn(fds_qq_q3, pca_res_qq_q3, PC1, PC2, cluster, study)
cluster_qq_tmm <- pca_plot_ipmn(fds_qq_tmm, pca_res_qq_tmm, PC1, PC2, cluster, study)
cluster_qq_rle <- pca_plot_ipmn(fds_qq_rle, pca_res_qq_rle, PC1, PC2, cluster, study)
cluster_qq_qn <- pca_plot_ipmn(fds_qq_qn, pca_res_qq_qn, PC1, PC2, cluster, study)

# Grid column and row titles
col_titles <- list(
  wrap_elements(grid::textGrob('', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('Q3', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('TMM', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('RLE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QN', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

row_titles <- list(
  wrap_elements(grid::textGrob('NONE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('BGSub', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QQ', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

# Create grid
p <- wrap_plots(
  col_titles[[1]], col_titles[[2]], col_titles[[3]], col_titles[[4]], col_titles[[5]],
  row_titles[[1]], cluster_none_q3, cluster_none_tmm, cluster_none_rle, cluster_none_qn,
  row_titles[[2]], cluster_bgsub_q3, cluster_bgsub_tmm, cluster_bgsub_rle, cluster_bgsub_qn,
  row_titles[[3]], cluster_qq_q3, cluster_qq_tmm, cluster_qq_rle, cluster_qq_qn,
  ncol = 5,
  heights = c(0.1, 1, 1, 1),
  widths = c(0.3, 1, 1, 1, 1)
) +
plot_layout(
  guides = 'collect'
) &
theme(plot.margin = unit(c(0.5, 1, 1, 1), "cm"))

# Add titles
p <- p + 
  plot_annotation(
    title = "Effect of Background Correction and Normalization Methods on PCA Clustering"
  ) & 
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5)
  )

f <- file.path(pca_dir, "pca_clustering_grid")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)
```

### Optimal clustering
``` {r optimal clustering}
npcs <- 3

# assess optimal number of clusters
x <- pca_res_qq_qn$x[,1:npcs]

p <- fviz_nbclust(x, FUNcluster=cluster::pam, method="wss")
f <- file.path(pca_dir, "pca_fviz_nbclust_wss")
ggsave(paste0(f, ".png"), p, width=4, height=4)
```

### Assign names to clusters
``` {r}

fds_none_q3$samples <- fds_none_q3$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "brain",
                           cluster == "c5" ~ "kidney")
)
x <- bind_cols(fds_none_q3$samples, as_tibble(pca_res_none_q3$x))

fds_none_tmm$samples <- fds_none_tmm$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "colon",
                           cluster == "c3" ~ "ln",
                           cluster == "c4" ~ "brain",
                           cluster == "c5" ~ "kidney")
)
x <- bind_cols(fds_none_tmm$samples, as_tibble(pca_res_none_tmm$x))

fds_none_rle$samples <- fds_none_rle$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "colon",
                           cluster == "c3" ~ "ln",
                           cluster == "c4" ~ "brain",
                           cluster == "c5" ~ "kidney")
)
x <- bind_cols(fds_none_rle$samples, as_tibble(pca_res_none_rle$x))

fds_none_qn$samples <- fds_none_qn$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "brain",
                           cluster == "c5" ~ "kidney")
)
x <- bind_cols(fds_none_qn$samples, as_tibble(pca_res_none_qn$x))

fds_bgsub_q3$samples <- fds_bgsub_q3$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "kidney",
                           cluster == "c5" ~ "brain")
)
x <- bind_cols(fds_bgsub_q3$samples, as_tibble(pca_res_bgsub_q3$x))

fds_bgsub_tmm$samples <- fds_bgsub_tmm$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "kidney",
                           cluster == "c5" ~ "brain")
)
x <- bind_cols(fds_bgsub_tmm$samples, as_tibble(pca_res_bgsub_tmm$x))

fds_bgsub_rle$samples <- fds_bgsub_rle$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "kidney",
                           cluster == "c5" ~ "brain")
)
x <- bind_cols(fds_bgsub_rle$samples, as_tibble(pca_res_bgsub_rle$x))

fds_bgsub_qn$samples <- fds_bgsub_qn$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "kidney",
                           cluster == "c5" ~ "brain")
)
x <- bind_cols(fds_bgsub_qn$samples, as_tibble(pca_res_bgsub_qn$x))

fds_qq_q3$samples <- fds_qq_q3$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "kidney",
                           cluster == "c5" ~ "brain")
)
x <- bind_cols(fds_qq_q3$samples, as_tibble(pca_res_qq_q3$x))

fds_qq_tmm$samples <- fds_qq_tmm$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "brain",
                           cluster == "c5" ~ "kidney")
)
x <- bind_cols(fds_qq_tmm$samples, as_tibble(pca_res_qq_tmm$x))

fds_qq_rle$samples <- fds_qq_rle$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "brain",
                           cluster == "c5" ~ "kidney")
)
x <- bind_cols(fds_qq_rle$samples, as_tibble(pca_res_qq_rle$x))

fds_qq_qn$samples <- fds_qq_qn$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "kidney",
                           cluster == "c5" ~ "brain")
)
x <- bind_cols(fds_qq_qn$samples, as_tibble(pca_res_qq_qn$x))
```

### Chi-squared analysis of clusters
``` {r}
# List of methods
methods <- list(fds_none_q3, fds_none_tmm, fds_none_rle, fds_none_qn, 
                fds_bgsub_q3, fds_bgsub_tmm, fds_bgsub_rle, fds_bgsub_qn, 
                fds_qq_q3, fds_qq_tmm, fds_qq_rle, fds_qq_qn)

# Initialize vectors to store Chi-squared statistics and p-values
chi_squared_stats <- numeric(length(methods))
method_names <- c("none_q3", "none_tmm", "none_rle", "none_qn",
                   "bgsub_q3", "bgsub_tmm", "bgsub_rle", "bgsub_qn",
                   "qq_q3", "qq_tmm", "qq_rle", "qq_qn")

# Loop through each method
for (i in seq_along(methods)) {
  # Create the contingency table
  contingency_table <- table(methods[[i]]$samples$study, methods[[i]]$samples$cluster_name)

  # Print the contingency table
  print(contingency_table)

  # Perform the Chi-squared test
  chi_squared_result <- chisq.test(contingency_table)
  print(chi_squared_result)

  # Store the Chi-squared statistic and p-value
  chi_squared_stats[i] <- chi_squared_result$statistic
}

# Format p-values to scientific notation
p_values_formatted <- format(p_values, scientific = TRUE)

# Combine results into a data frame
results_df <- data.frame(
  Method = method_names,
  Chi_Squared_Statistic = chi_squared_stats
 )

# Print the results
print(results_df)

```

# FIGURE 5: DE
## DE analysis

```{r}
de_setup <- function(method_name, de_padj_cutoff=0.05, de_log2fc_cutoff=1) {
  # access fds and log2_norm_cpm
  fds_name <- paste0("fds_", method_name)
  fds <- get(fds_name)
  log2_name <- paste0("log2_norm_cpm_", method_name)
  log2_norm_cpm <- get(log2_name)

  # init de results
  de <- NULL

  # DE analysis of clusters
  s <- fds$samples
  y <- log2_norm_cpm[, s$aoi]
  s$cluster <- factor(s$cluster_name, levels=c("pancreas", "ln", "brain", "colon", "kidney"))

  # compare each cluster to all of the rest
  design <- model.matrix(~0 + cluster, s)
  contrasts <- makeContrasts(clusterPanc = clusterpancreas - (clusterln + clusterbrain + clusterkidney + clustercolon)/4,
                             clusterLn = clusterln - (clusterpancreas + clusterbrain + clusterkidney + clustercolon)/4,
                             clusterBrain = clusterbrain - (clusterln + clusterpancreas + clusterkidney + clustercolon)/4,
                             clusterKidney = clusterkidney - (clusterln + clusterbrain + clusterpancreas + clustercolon)/4,
                             clusterColon = clustercolon - (clusterln + clusterbrain + clusterkidney + clusterpancreas)/4,
                             levels = design)

  res <- run_limma_trend(y, design, contrasts,
                         padj_cutoff=de_padj_cutoff,
                         log2fc_cutoff=de_log2fc_cutoff)

  # write cluster results
  file_name <- paste0("de_clusters_", method_name, ".xlsx")
  write_xlsx(res, file.path(results_dir, file_name))
  return(res)
}

run_limma_trend <- function(y, design, contrasts, padj_cutoff, log2fc_cutoff) {
  fit <- lmFit(y, design)
  fit <- contrasts.fit(fit, contrasts)
  fit <- eBayes(fit, trend=TRUE)

  x <- NULL
  for (coef in colnames(contrasts)) {
    res <- topTable(fit, coef=coef, number=Inf, sort.by="none")
    res <- as_tibble(res, rownames="gene")
    res <- res %>%
      select(gene,
             log2fc=logFC,
             avgexpr=AveExpr,
             pval=P.Value,
             padj=adj.P.Val) %>%
      mutate(de = case_when(padj > padj_cutoff ~ "no",
                            log2fc < -log2fc_cutoff ~ "dn",
                            log2fc > log2fc_cutoff ~ "up",
                            TRUE ~ "no"),
             contrast = coef)
    x <- bind_rows(x, res)
  }
  return(x)
}

# res_none_q3 <- de_setup("none_q3")
# res_none_tmm <- de_setup("none_tmm")
# res_none_rle <- de_setup("none_rle")
# res_none_qn <- de_setup("none_qn")
# 
# res_bgsub_q3 <- de_setup("bgsub_q3")
# res_bgsub_tmm <- de_setup("bgsub_tmm")
# res_bgsub_rle <- de_setup("bgsub_rle")
# res_bgsub_qn <- de_setup("bgsub_qn")
# 
# res_qq_q3 <- de_setup("qq_q3")
# res_qq_tmm <- de_setup("qq_tmm")
# res_qq_rle <- de_setup("qq_rle")
# res_qq_qn <- de_setup("qq_qn")

# write tables
# table(res_none_q3$contrast, res_none_q3$de)

```

## DE Volcano plots

```{r de volcano plots}

de <- NULL
de <- bind_rows(de, res_qq_qn)

plot_de_volcano <- function(res, nlabel=50) {
  # label the top n up/down genes
  label_data <- bind_rows(filter(res, de == "up") %>% slice_max(log2fc, n=nlabel),
                          filter(res, de == "dn") %>% slice_min(log2fc, n=nlabel))
  label_data <- distinct(label_data)

  p <- ggplot(res, aes(x=log2fc, y=-log10(padj), color=de, size=de)) +
    geom_point(alpha=0.7) +
    geom_text_repel(data=label_data, color="black", size=3, aes(label=gene), max.overlaps=Inf) +
    scale_color_manual(values=c("no"="grey", "dn"="blue", "up"="red")) +
    scale_size_manual(values=c("no"=1, "dn"=2, "up"=2)) +
    theme_minimal() +
    theme(legend.position="bottom") +
    theme(axis.line = element_line(color = "black")) +
    labs(x="log2fc", y="-log10(padj)")
  return(p)
}

volcano_dir <- file.path(plot_dir, "volcano")

# volcano plots
for (mycontrast in unique(de$contrast)) {
  print(mycontrast)
  x <- filter(de, contrast == mycontrast)
  p <- plot_de_volcano(x) + 
    labs(title="Volcano Plot", subtitle=mycontrast)
  f <- paste0("volcano_de_", mycontrast)
  ggsave(file.path(volcano_dir, paste0(f, ".pdf")), p, width=8, height=8)
  ggsave(file.path(volcano_dir, paste0(f, ".png")), p, width=8, height=8)
}

```

## DE Heatmaps

```{r de heatmaps, include=FALSE}

# heatmap functions
plot_pheatmap <- function(m, 
                          annot_col, 
                          annot_row, 
                          annot_colors,
                          filename, width, height,
                          xmin=-6, xmax=6) {
  pheatmap(m,
           scale = "none",
           color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
           breaks = seq(xmin, xmax, length.out=51),
           annotation_col = annot_col,
           annotation_row = annot_row,
           annotation_colors = annot_colors,
           border_color = NA,
           cluster_rows = TRUE,
           cluster_cols = TRUE,
           clustering_distance_rows = "euclidean",
           clustering_distance_cols = "euclidean",
           clustering_method = "ward.D2",
#           cellwidth = 6,
#           cellheight = 6,
           fontsize = 6,
           fontsize_row = 6,
           fontsize_col = 6,
           filename = filename, 
           width = width,
           height = height)
}

# setup heatmap annotations
gene_annot <- fds_qq_qn$meta 
# %>%
#   left_join(hpa_pdac, by=c("gene"="gene_symbol")) %>%
#   left_join(cptac_de_merged, by=c("gene"="gene_symbol")) %>%
#   mutate(de_cptac_rna = replace_na(de_cptac_rna, "na"),
#          de_cptac_prot = replace_na(de_cptac_prot, "na"))
annot_row <- gene_annot %>%
  select(gene) %>%
  column_to_rownames("gene") %>%
  as.data.frame()
annot_col <- column_to_rownames(fds_qq_qn$samples, "aoi") %>%
  select(patient, study, cluster_name) %>%
  as.data.frame()

#
# IPMN clusters
#
# de gene params
padj_cutoff <- 0.01
log2fc_cutoff <- 3
ntop <- 10
contrasts <- c("clusterPanc", "clusterLn", "clusterKidney", "clusterColon", "clusterBrain")

g <- filter(de_qq_qn, contrast %in% contrasts,
            abs(log2fc) > log2fc_cutoff, 
            padj < padj_cutoff) %>%
  group_by(contrast, de) %>%
  slice_max(abs(log2fc), n=ntop) %>%
  ungroup() %>%
  distinct(gene) %>% 
  pull(gene)
ngenes <- length(g)
cluster_marker_genes <- g

# centered gene expression (subtract median)
s <- fds_qq_qn$samples
x <- log2_norm_cpm_qq_qn[g, s$aoi]
u <- apply(x, 1, mean)
x <- sweep(x, 1, u, FUN="-")
xmin <- -6
xmax <- 6

# heatmap
heatmap_dir <- file.path(plot_dir, "heatmap")

f <- file.path(plot_dir,  paste0("heatmap_de_cluster_markers_", ngenes, "_genes"))
h <- 6
w <- 6
p <- plot_pheatmap(x, 
                   annot_col[s$aoi,], 
                   annot_row[g,], 
                   annot_colors=scale_color_continuous(),
                   filename=paste0(f, ".png"), 
                   width=w, height=h,
                   xmin=xmin, xmax=xmax)
save_pheatmap_pdf(p, filename=paste0(f, ".pdf"), width=w, height=h)

```

## PCA feature plots for marker genes

```{r pca feature plot for each marker gene, include=FALSE}

s <- fds_qq_qn$samples
y <- log2_norm_cpm_qq_qn
feature_plot_genes <- c("MS4A1", "CPA1", "SNAP25", "IGHA1", "IGHG1")

for (gene in feature_plot_genes) {
  print(gene)
  x <- bind_cols(s, as_tibble(pca_res_qq_qn$x), gene=unlist(y[gene,]))
  p <- ggplot(x, aes(x=PC1, y=PC2, color=gene, shape=study)) +
    geom_point(alpha = 0.8, size=3) +
    scale_color_viridis_c() +
    theme_minimal() +
    theme(aspect.ratio=1) +
    labs(x="PC1", y="PC2", title=gene)
  f <- file.path(plot_dir, paste0("pca_featureplot_", gene))
  ggsave(paste0(f, ".pdf"), plot=p, width=5, height=5)
  ggsave(paste0(f, ".png"), plot=p, width=5, height=5)
}

```

## DE 2D scatter plots

```{r ipmn de 2d scatter plots}

plot_de_scatter2d <- function(tbl, xcontrast, ycontrast, nlabel=10, 
                              highlight_genes=NULL) {
  if (is.null(highlight_genes)) {
    highlight_genes <- c()
  }
  xde <- sym(paste0("de_", xcontrast))
  yde <- sym(paste0("de_", ycontrast))
  xlog2fc <- sym(paste0("log2fc_", xcontrast))
  ylog2fc <- sym(paste0("log2fc_", ycontrast))

  x <- tbl %>% 
    mutate(subtype = paste0("x", {{xde}}, "_", "y", {{yde}}),
           log2fc_abs_x = case_when({{xde}} == "no" ~ 0,
                                    TRUE ~ abs({{xlog2fc}})),
           log2fc_abs_y = case_when({{yde}} == "no" ~ 0,
                                    TRUE ~ abs({{ylog2fc}})),
           log2fc_abs_max = pmax(log2fc_abs_x, log2fc_abs_y))

  color_scale = c(xno_yno = "#888888",
                  xup_yup = "#FF0000",
                  xdn_ydn = "#0000aa",
                  xup_ydn = "#eeee00",
                  xdn_yup = "#ff00ff",
                  xup_yno = "#32cd32",
                  xdn_yno = "#670092",
                  xno_yup = "#ff6700",
                  xno_ydn = "#16ccc4")
  size_scale = c(xno_yno = 1, 
                 xup_yup = 2,
                 xdn_ydn = 2,
                 xup_ydn = 2,
                 xdn_yup = 2,
                 xup_yno = 2,
                 xdn_yno = 2,
                 xno_yup = 2,
                 xno_ydn = 2)

  annot_data <- filter(x, subtype != "xno_yno")
  none_data <- filter(x, subtype == "xno_yno")
  label_data <- annot_data %>% 
    group_by(subtype) %>%
    slice_max(log2fc_abs_max, n=nlabel) %>%
    ungroup()
  highlight_data <- filter(x, gene %in% highlight_genes)

  p <- ggplot(x, aes(x={{xlog2fc}}, 
                     y={{ylog2fc}},
                     color=subtype, 
                     size=subtype)) +
    geom_point(data=none_data, alpha=0.4) + 
    geom_point(data=annot_data, alpha=0.8) +
    geom_text_repel(data=label_data, color="black", size=3, aes(label=gene), max.overlaps=Inf) + 
    geom_point(data=highlight_data, pch=21, fill=NA, size=5, colour="#ffff00", stroke=1) + 
    geom_point(data=highlight_data, pch=21, fill=NA, size=4, colour="#222222", stroke=1, alpha=0.5) +
    geom_text_repel(data=highlight_data, color="black", size=3, aes(label=gene), max.overlaps=Inf) +     
    scale_color_manual(values = color_scale) +
    scale_size_manual(values = size_scale, guide = "none") + 
    theme_minimal() +
    xlab(rlang::as_string(xlog2fc)) +
    ylab(rlang::as_string(ylog2fc)) +
    labs(color = "DE Genes")
  return(p)
}

# merge de analysis by gene
de_merged <- de_qq_qn %>% select(gene, contrast, log2fc, padj, de) %>%
  pivot_wider(names_from = contrast, values_from = c(log2fc, padj, de))
nlabel <- 10
width <- 8
height <- 8

#
# IPMN Clusters
#
p <- plot_de_scatter2d(de_merged, 
                       xcontrast="clusterColon", 
                       ycontrast="clusterLn", 
                       nlabel=nlabel) +
  labs(title="IPMN Specimens", 
       subtitle="clusterColon (x-axis) clusterLn (y-axis)")
f <- file.path(plot_dir, "scatter2d_de_x_clusterColon_y_clusterLn")
ggsave(paste0(f, ".pdf"), plot=p, width=8, height=8)
ggsave(paste0(f, ".png"), plot=p, width=8, height=8)

```

## FIGURE 6: GSEA
``` {r gsea}

```

# Downstream data analysis 
## Highly variable genes: updated
```{r updated highly variable genes, include=FALSE}
most_var_genes <- function(x, ntop=500, span=0.50) {
  tbl <- bind_cols(
    u = rowMeans(x),
    v = apply(x, 1, function(x) var(x))
  )
  lovar <- loess(v ~ u, data=tbl, span=span)
  tbl$lovar <- predict(lovar)
  tbl$vadj <- tbl$v - tbl$lovar
  keep <- order(tbl$vadj, decreasing=TRUE)[seq_len(min(ntop, nrow(tbl)))]
  return(x[keep,])
}

maplot <- function(y, ntop=1000, nlabel=100, myspan=0.5) {
  tbl <- bind_cols(
    g = rownames(y),
    u = rowMeans(y),
    v = apply(y, 1, function(x) var(x))
  )
  lovar <- loess(v ~ u, data=tbl, span=myspan)
  tbl$lovar <- predict(lovar)
  tbl$vadj <- tbl$v - tbl$lovar
  tbl <- tbl %>% 
    mutate(rank = min_rank(-vadj)) %>%
    arrange(rank)
  label_data <- filter(tbl, rank <= nlabel)
  top_data <- filter(tbl, rank < ntop)
  bot_data <- filter(tbl, rank >= ntop)
  p <- ggplot(tbl) +
    geom_point(data=bot_data, aes(u, v), alpha=0.5, size=1, color="#aaaaaa") +
    geom_point(data=top_data, aes(u, v), alpha=0.8, size=2, color="#ffaaaa") +
    geom_line(aes(u, lovar), color="blue") +
    geom_text_repel(data=label_data, aes(u, v, label=g), color="black", size=3, max.overlaps=Inf) +
    theme_minimal()
  return(p)
}

heatmap_genes <- function(s, x, filename="heatmap.pdf", width=10, height=10) {
  s <- s %>% mutate(
    noise_quantile = ntile(desc(bg_auc), 8)
  )
  annot_col <- column_to_rownames(s, "aoi") %>%
    select(patient, path_specimen, path_aoi, noise_quantile) %>%
    as.data.frame()
  p <- pheatmap(x,
                scale="row",
                show_rownames=TRUE,
                show_colnames=FALSE,
                border_color=NA,
                clustering_distance_rows = "euclidean",
                clustering_distance_cols = "euclidean",
                clustering_method = "ward.D2",
                # clustering_method = "average",
                # clustering_distance_rows = "correlation",
                # clustering_distance_cols = "correlation",
                breaks = seq(-3, 3, 0.05),
                color = colorRampPalette(c("purple3", "black", "yellow2"))(120),
                annotation_col = annot_col,
                # annotation_colors = color_scales,
                fontsize=4,
                filename=filename,
                width=width,
                height=height)
  return(p)
}

y <- log2_norm_cpm_merged
ntop <- 1000
nlabel <- 100
myspan <- 0.5

# maplot
p <- maplot(log2_norm_cpm_merged, ntop=ntop, nlabel=nlabel, myspan=myspan)
f <- file.path(plot_dir, paste0("maplot"))
ggsave(paste0(f, ".pdf"), plot=p, width=10, height=10)
ggsave(paste0(f, ".png"), plot=p, width=10, height=10)

# select most variable genes
ntop <- 500
myspan <- 0.5
s <- fds$samples
y <- most_var_genes(log2_norm_cpm_merged, ntop=ntop, span=myspan)

# heatmap
f <- file.path(plot_dir, "heatmap_highvar_genes.pdf")
p <- heatmap_genes(s, y, f)
```

## PCA: updated
```{r updated pca}
# subset most variable genes
ntop <- 250
myspan <- 0.5
s <- fds$samples
y <- most_var_genes(log2_norm_cpm_merged, ntop=ntop, span=myspan)

# plot dimensions
width <- 12
height <- 10

# PCA
npcs <- 50
pca_res <- prcomp(t(y), center=TRUE, scale.=TRUE, rank.=npcs)
pca_res.var <- round(100 * pca_res$sdev^2 / sum( pca_res$sdev^2 ), 2)
pca_res.var_txt <- paste(colnames(pca_res$x), " (", paste(as.character(pca_res.var), "%", ")", sep=""), sep="")

# fviz plots
fviz_ntop <- 50

# percent variance of each PC
p <- fviz_eig(pca_res, col.var="blue", addlabels=TRUE)
f <- file.path(plot_dir, paste0("pca_fviz_eig.pdf"))
ggsave(f, p)

# contributions of variables to each PC
p <- fviz_contrib(pca_res, choice="var", axes=1, top=fviz_ntop) +
  theme(axis.text.x = element_text(angle=90))
f <- file.path(plot_dir, paste0("pca_fviz_contrib1.pdf"))
ggsave(f, p)

p <- fviz_contrib(pca_res, choice="var", axes=2, top=fviz_ntop) +
  theme(axis.text.x = element_text(angle=90))
f <- file.path(plot_dir, paste0("pca_fviz_contrib2.pdf"))
ggsave(f, p)

p <- fviz_contrib(pca_res, choice="var", axes=3, top=fviz_ntop) +
  theme(axis.text.x = element_text(angle=90))
f <- file.path(plot_dir, paste0("pca_fviz_contrib3.pdf"))
ggsave(f, p)

# contribution to first 3 PCs
p <- fviz_contrib(pca_res, choice="var", axes = 1:3, top=fviz_ntop) +
  theme(axis.text.x = element_text(angle=90))
f <- file.path(plot_dir, paste0("pca_fviz_contrib123.pdf"))
ggsave(f, p)

# graph of variables contributing to PCA
p <- fviz_pca_var(
  pca_res,
  axes = c(1,2),
  col.var = "contrib", # Color by the quality of representation
  gradient.cols = c("darkorchid4", "gold", "darkorange"),
  repel=TRUE,
  select.var = list(contrib=fviz_ntop)
)
f <- file.path(plot_dir, paste0("pca_fviz_pca_var_pc12.pdf"))
ggsave(f, p, width=7, height=7)

# graph of variables contributing to PCA
p <- fviz_pca_var(
  pca_res,
  axes = c(1,3),
  col.var = "contrib", # Color by the quality of representation
  gradient.cols = c("darkorchid4", "gold", "darkorange"),
  repel=TRUE,
  select.var = list(contrib=fviz_ntop)
)
f <- file.path(plot_dir, paste0("pca_fviz_pca_var_pc13.pdf"))
ggsave(f, p, width=7, height=7)

# graph of variables contributing to PCA
p <- fviz_pca_var(
  pca_res,
  axes = c(2,3),
  col.var = "contrib", # Color by the quality of representation
  gradient.cols = c("darkorchid4", "gold", "darkorange"),
  repel=TRUE,
  select.var = list(contrib=fviz_ntop)
)
f <- file.path(plot_dir, paste0("pca_fviz_pca_var_pc23.pdf"))
ggsave(f, p, width=7, height=7)

#
# PCA plots
#
x <- bind_cols(s, as_tibble(pca_res$x))

# bg_ks_dist
p <- ggplot(x, aes(x=PC1, y=PC2, color=bg_auc)) +
  geom_point(alpha = 0.8, size=3) +
  scale_color_viridis_c() +
  theme_bw() +
  theme(aspect.ratio=1) +
  theme(legend.text = element_text(size=7)) +
  labs(x=pca_res.var_txt[1], y=pca_res.var_txt[2])
f <- file.path(plot_dir, paste0("pca_bg_auc.pdf"))
ggsave(f, plot=p, device="pdf", width=width, height=height)

# patient
p <- ggplot(x, aes(x=PC1, y=PC2, color=patient, shape=study)) +
  geom_point(alpha = 0.8, size=3) +
  # scale_color_manual(values = color_scales$patient) +
  theme_bw() +
  theme(aspect.ratio=1) +
  theme(legend.text = element_text(size=7)) +
  labs(x=pca_res.var_txt[1], y=pca_res.var_txt[2])
f <- file.path(plot_dir, paste0("pca_patient.pdf"))
ggsave(f, plot=p, device="pdf", width=width, height=height)

# region
p <- ggplot(x, aes(x=PC1, y=PC2, color=region, shape=study)) +
  geom_point(alpha = 0.8, size=3) +
  # scale_color_manual(values = color_scales$patient) +
  theme_bw() +
  theme(aspect.ratio=1) +
  theme(legend.text = element_text(size=7)) +
  labs(x=pca_res.var_txt[1], y=pca_res.var_txt[2])
f <- file.path(plot_dir, paste0("pca_region.pdf"))
ggsave(f, plot=p, device="pdf", width=width, height=height)

# subset
p <- ggplot(x, aes(x=PC1, y=PC2, color=subset, shape=study)) +
  geom_point(alpha = 0.8, size=3) +
  # scale_color_manual(values = color_scales$patient) +
  theme_bw() +
  theme(aspect.ratio=1) +
  theme(legend.text = element_text(size=7)) +
  labs(x=pca_res.var_txt[1], y=pca_res.var_txt[2])
f <- file.path(plot_dir, paste0("pca_subset.pdf"))
ggsave(f, plot=p, device="pdf", width=width, height=height)

# study
p <- ggplot(x, aes(x=PC1, y=PC2, color=study)) +
  geom_point(alpha = 0.8, size=3) +
  # scale_color_manual(values=color_scales$path) +
  theme_bw() +
  theme(aspect.ratio=1) +
  theme(legend.text = element_text(size=7)) +
  labs(x=pca_res.var_txt[1], y=pca_res.var_txt[2])
f <- file.path(plot_dir, paste0("pca_study.pdf"))
ggsave(f, plot=p, device="pdf", width=width, height=height)

#
# Plots of PC1, PC2, and PC3
#
pca_plot <- function(x, xvar, yvar, color_by, shape_by) {
  color_by_string <- rlang::as_string(rlang::ensym(color_by))
  shape_by_string <- rlang::as_string(rlang::ensym(shape_by))

  # if (is.character(pull(x, {{color_by}}))) {
  #   scale_color_func <- scale_color_manual(values = color_scales[[color_by_string]])
  # } else {
  #   scale_color_func <- scale_color_viridis_c()
  # }

  p <- ggplot(x, aes(x={{xvar}}, y={{yvar}}, color={{color_by}}, shape={{shape_by}})) +
    geom_point(alpha = 0.8, size=3) +
    #scale_color_func +
    #scale_color_manual(values = color_scales[[color_by_string]]) +
    #scale_shape_manual(values = shape_scales[[shape_by_string]]) +
    theme_bw() +
    theme(aspect.ratio=1,
          legend.text = element_text(size=7))
    # coord_fixed() +
  return(p)
}

# bg_auc
p12 <- pca_plot(x, PC1, PC2, bg_auc, study)
p13 <- pca_plot(x, PC1, PC3, bg_auc, study)
p23 <- pca_plot(x, PC2, PC3, bg_auc, study)
p <- p12 + p13 + p23 + 
  plot_layout(guides = "collect")
f <- file.path(plot_dir, paste0("pca_bg_auc_pc123.pdf"))
ggsave(f, plot=p, device="pdf", width=12, height=6)

# region
p12 <- pca_plot(x, PC1, PC2, region, study)
p13 <- pca_plot(x, PC1, PC3, region, study)
p23 <- pca_plot(x, PC2, PC3, region, study)
p <- p12 + p13 + p23 + 
  plot_layout(guides = "collect")
f <- file.path(plot_dir, paste0("pca_region_pc123.pdf"))
ggsave(f, plot=p, device="pdf", width=12, height=6)

# patient
p12 <- pca_plot(x, PC1, PC2, patient, study)
p13 <- pca_plot(x, PC1, PC3, patient, study)
p23 <- pca_plot(x, PC2, PC3, patient, study)
p <- p12 + p13 + p23 + 
  plot_layout(guides = "collect")
f <- file.path(plot_dir, paste0("pca_patient_pc123.pdf"))
ggsave(f, plot=p, device="pdf", width=12, height=6)

# study
p12 <- pca_plot(x, PC1, PC2, study)
p13 <- pca_plot(x, PC1, PC3, study)
p23 <- pca_plot(x, PC2, PC3, study)
p <- p12 + p13 + p23 + 
  plot_layout(guides = "collect")
f <- file.path(plot_dir, paste0("pca_study_pc123.pdf"))
ggsave(f, plot=p, device="pdf", width=12, height=6)
```

### 2024-08-28 inst/manuscript_figures.Rmd
vignette ds changes

@@
 
-f <- file.path(pca_dir, "pca_grid_2000")
+f <- file.path(pca_dir, "pca_grid_500")
 ggsave(paste0(f, ".png"), p, width=20, height=15)
@@
 ```{r}
-# Example data (replace with your actual data)
+
 methods <- c("none_q3", "none_tmm", "none_rle", "none_qn", 
@@
 s <- fds_qq_qn$samples
-pcs <- c("PC1", "PC2", "PC3", "PC4", "PC5")
+pcs <- c("PC1", "PC2", "PC3", "PC4")
 pca_tbl <- pca_res_qq_qn$x
@@
 
-f <- file.path(pca_dir, "pca_12345_grid")
+f <- file.path(pca_dir, "pca_1234_grid")
 ggsave(paste0(f, ".png"), p, width=25, height=20)
@@
 
-  p <- ggplot(res, aes(x=log2fc, y=-log10(padj), color=de, size=de)) +
+  p <- ggplot(res, aes(x=log2fc, y=-log10(padj + 1e-300), color=de, size=de)) +
     geom_point(alpha=0.7) +
@@
                           annot_col,
-                          annot_row, 
+                          # annot_row, 
                           annot_colors,
@@
            annotation_col = annot_col,
-           annotation_row = annot_row,
-           annotation_colors = annot_colors,
+           # annotation_row = annot_row,
+           # annotation_colors = annot_colors,
            border_color = NA,
@@
 # setup heatmap annotations
+# column annotations: show clusters, study
+# no row annotations
 gene_annot <- fds_qq_qn$meta
@@
 #          de_cptac_prot = replace_na(de_cptac_prot, "na"))
-annot_row <- gene_annot %>%
-  select(gene) %>%
-  column_to_rownames("gene") %>%
-  as.data.frame()
+# annot_row <- gene_annot %>%
+#   select(gene) %>%
+#   column_to_rownames("gene") %>%
+#   as.data.frame()
 annot_col <- column_to_rownames(fds_qq_qn$samples, "aoi") %>%
-  select(patient, study, cluster_name) %>%
+  select(study, cluster_name) %>%
   as.data.frame()
@@
 ``` {r gsea}
-
+# read msigdb data
+f <- file.path(data_dir, "msigdb.v2023.2.Hs.json")
+msigdb_gene_sets <- read_msigdb_json_file(f)
+# determine gene universe
+msigdb_genes <- intersect(fds$meta$gene, msigdb_gene_sets$gene_symbol)
+# filter genes not in our dataset
+msigdb_gene_sets <- msigdb_gene_sets %>%
+  dplyr::filter(gene_symbol %in% fds$meta$gene) %>%
+  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
+
+# gene set categories
+gs_hallmark <- dplyr::filter(msigdb_gene_sets, gs_cat == "H")
+gs_hallmark <- split(x = gs_hallmark$gene_symbol, f = gs_hallmark$gs_name)
+
+# pass log fold change into gsea with hallmark gene sets (sending new code)
+
+get_de_ranks <- function(xde, mycontrast) {
+  ranks <- xde %>% 
+    dplyr::filter(contrast == mycontrast) %>%
+    select(gene, log2fc, padj) %>%
+    mutate(rank = log2fc)
+  ranks = sort(setNames(ranks$rank, ranks$gene), decreasing = TRUE)
+  return(ranks)
+}
+
+plot_gsea_enrichment <- function(x, ranks, a, gs) {
+  txt_stat <- paste0("ES=", round(x$ES,2), " NES=", round(x$NES, 2), " padj=", format(x$padj, scientific=TRUE, digits=3))
+  txt_title <- paste0("ranks: ", a, " gs: ", x$pathway)
+  p <- plotEnrichment(gs[[x$pathway]], ranks) +
+    annotate(geom="text", x=150, y=0.1, label=txt_stat, hjust=0) +
+    labs(title = txt_title, 
+         xlab = "Rank",
+         ylab = "Enrichment Score") +
+    theme(plot.title = element_text(size=8))
+  return(p)
+}
+
+# fgsea_run_batch <- function(xde, xgs, 
+#                             fgsea_eps=0,
+#                             fgsea_nperms=10000, 
+#                             fgsea_nproc=0) {
+#   fgsea_scoretypes <- c("pos", "neg")
+#   gsea <- NULL
+#   for (xcontrast in unique(xde$contrast)) {
+#     ranks <- get_de_ranks(xde, xcontrast)
+#     for (fgsea_scoretype in fgsea_scoretypes) {
+      res <- fgsea(pathways=xgs, 
+                   stats=ranks, 
+                   eps=fgsea_eps, 
+                   scoreType = fgsea_scoretype,
+                   nPermSimple=fgsea_nperms, 
+                   nproc=fgsea_nproc)
+      res <- mutate(res, 
+                    contrast = xcontrast, 
+                    dir = fgsea_scoretype)
+      gsea <- bind_rows(gsea, res)      
+#     }
+#   }
+#   return(gsea)
+# }
+
+#
+# Run full analysis
+#
+gsea <- NULL
+
+# hallmark
+res <- fgsea_run_batch(de, gs_hallmark, 
+                       fgsea_eps=0, fgsea_nperms=gsea_nperms, fgsea_nproc=10)
+gsea <- bind_rows(gsea, res)
+gsea_all <- gsea
 ```

### 2024-08-28 R/stgeomx.R
duke ds updates

@@ -85,7 +85,7 @@ copy <- function(ds) {
 #' @examples
 #' # example code
 #' data(example_ds, package= "stgeomx")
-#' merge_ds(example_ds, example_ds)
+#' merge(example_ds, example_ds)
 #'
 merge <- function(a, b) {
   # merge sample tables

### 2024-08-28 tests/testthat/test-preprocess.R

@@ -3,7 +3,7 @@
 test_that("merge_duplicate_ds", {
   # problem: when reading in expected, it's deleting the extra
   example_ds <- read_xlsx(system.file("extdata", "simple_geomx2.xlsx", package = "stgeomx"))
-  duplicated <- merge_ds(example_ds, example_ds)
+  duplicated <- merge(example_ds, example_ds)
   expected <- read_xlsx(system.file("extdata", "merge_duplicate_result.xlsx", package = "stgeomx"))
   # expect_equal(duplicated, expected)
   expect_equal(4, 4)
@@ -13,7 +13,7 @@ test_that("merge_duplicate_probes", {
   ds_1 <- read_xlsx(system.file("extdata", "simple_geomx2.xlsx", package = "stgeomx"))
   ds_2 <- read_xlsx(system.file("extdata", "simple_geomx4.xlsx", package = "stgeomx"))
   expected <- read_xlsx(system.file("extdata", "merge_same_probes_result.xlsx", package = "stgeomx"))
-  duplicated <- merge_ds(ds_1, ds_2)
+  duplicated <- merge(ds_1, ds_2)
   expect_equal(duplicated, expected)
 })
 
@@ -21,6 +21,6 @@ test_that("merge_different_probes", {
   ds_1 <- read_xlsx(system.file("extdata", "simple_geomx2.xlsx", package = "stgeomx"))
   ds_2 <- read_xlsx(system.file("extdata", "simple_geomx3.xlsx", package = "stgeomx"))
   expected <- read_xlsx(system.file("extdata", "merge_different_probes_result.xlsx", package = "stgeomx"))
-  duplicated <- merge_ds(ds_1, ds_2)
+  duplicated <- merge(ds_1, ds_2)
   expect_equal(duplicated, expected)
 })

### 2024-08-30 R/hvg.R
moved highly variable gene functions into stgeomx package
updated vignette

@@ -1,0 +1,79 @@
+#
+# Spatial Transcriptomics using NanoString GeoMx DSP
+#
+
+
+#'
+#' Model and rank highly variable genes in dataset
+#'
+#' Models mean-variance relationship using loess. Ratio of actual variance
+#' versus modeled variance is used to rank genes
+#'
+#' @importFrom magrittr %>%
+#' @import dplyr
+#' @import rlang
+#'
+#' @param x matrix/data frame of normalized log2 transformed gene expression,
+#' where gene identifiers are present in the row names of the matrix
+#' @param loess.span span parameter for loess model
+#' @param min_var_quantile [0.0-1.0] used as lower bound for predicted
+#' variance by loess model (in rare cases the predicted variance by loess
+#' can be less than zero or very close to zero, resulting in high variance
+#' ratios)
+#' @returns tibble object with per-gene parameters
+#' @export
+#'
+get_hvg_tbl <- function(x, loess.span=0.25, min_var_quantile=0.01) {
+  tbl <- dplyr::bind_cols(
+    gene = rownames(x),
+    u = rowMeans(x),
+    v = apply(x, 1, function(x) { stats::var(x) })
+  )
+  lovar <- stats::loess(v ~ u, data=tbl, span=loess.span)
+  # stabilize fitted variance (loess can produce negative values
+  # and values very close to zero), setting it to a low positive number
+  tbl$vloess <- stats::predict(lovar)
+  tbl$vminbound <- stats::quantile(tbl$v[tbl$v > 0], min_var_quantile)
+  tbl$vloess_bounded <- pmax(tbl$vloess, tbl$vminbound)
+  # compute variance ratios
+  tbl <- dplyr::mutate(tbl,
+    vratio = .data$v / .data$vloess_bounded,
+    rank = dplyr::min_rank(-.data$vratio)
+  )
+  return(tbl)
+}
+
+
+#'
+#' plot of gene mean versus variance with line showing predicted (expected)
+#' variance by loess model. dashed line shows lower bound to predicted
+# 'variance
+#'
+#' @import ggplot2
+#' @importFrom ggrepel geom_text_repel
+#'
+#' @param x matrix/data frame (normalized log2 gene expression)
+#' @param nhvg number of top highly variable genes to color differently
+#' @param nlabel number of top highly variable genes to label
[…]

### 2024-08-30 R/stgeomx.R

@@ -93,8 +93,8 @@ merge <- function(a, b) {
   s <- bind_rows(select(a$samples, all_of(sample_cols)),
                  select(b$samples, all_of(sample_cols)))
   # merge count data
-  a_counts <- bind_cols(select(a$meta, probe, gene, bg), a$counts)
-  b_counts <- bind_cols(select(b$meta, probe, gene, bg), b$counts)
+  a_counts <- bind_cols(select(a$meta, .data$probe, .data$gene, .data$bg), a$counts)
+  b_counts <- bind_cols(select(b$meta, .data$probe, .data$gene, .data$bg), b$counts)
   # counts <- inner_join(a_counts, b_counts, by=c("probe", "gene", "bg"),
   #                      relationship="one-to-one")
 

### 2024-09-24 R/input.R
added functions for reading DCC and PKC files

@@ -183,3 +183,145 @@ read_xlsx <- function(xlsx_file,
                       gene_key_col, negprobe_regex)
   return(ds)
 }
+
+
+
+
+#'
+#' Read Nanostring probe definition file (.pkc)
+#'
+#' @importFrom magrittr %>%
+#' @import dplyr
+#'
+#' @param pkc_file probe definition file (.pkc)
+#' @returns tibble of probe information
+#' @export
+read_pkc <- function(pkc_file) {
+  pkc <- yaml::read_yaml(pkc_file)
+  probes <- tibble("target" = pkc$Targets)
+  probes <- probes %>% tidyr::unnest_wider("target") %>%
+    dplyr::rename("TargetName" = "DisplayName") %>%
+    tidyr::unnest_longer("Probes") %>%
+    tidyr::unnest_wider("Probes") %>%
+    dplyr::rename("ProbeDisplayName" = "DisplayName") %>%
+    dplyr::mutate(bg = startsWith(.data$CodeClass, "Negative"))
+  return(probes)
+}
+
+
+#'
+#' Read Nanostring digital count conversion (DCC) files
+#'
+#' @importFrom magrittr %>%
+#' @import dplyr
+#'
+#' @param dcc_tar_file tarball of DCC files
+#' @param dcc_dir temporary directory to store DCC files
+#' @returns list of tibbles for 'samples' and 'counts'
+#' @export
+read_dcc <- function(dcc_tar_file, dcc_dir) {
+  # cleanup and overwrite files in 'dcc_dir' if exists
+  if (file.exists(dcc_dir)) {
+    unlink(dcc_dir, recursive = TRUE)
+  }
+  # extract dcc tar file
+  utils::untar(dcc_tar_file, exdir=dcc_dir)
+  # read dcc files
+  dcc_files <- list.files(dcc_dir, full.names=TRUE)
+  dcc_data <- sapply(dcc_files, GeomxTools::readDccFile, simplify=FALSE)
+  dcc_data <- tibble(dcc = dcc_data) %>%
+    tidyr::unnest_wider("dcc", simplify=FALSE)
+
+  samples <- dcc_data %>%
+    dplyr::select("Header", "Scan_Attributes", "NGS_Processing_Attributes") %>%
+    tidyr::unnest(cols = c("Header", "Scan_Attributes", "NGS_Processing_Attributes"))
+
+  counts <- dcc_data %>%
[…]

### 2024-09-24 inst/manuscript_figures.Rmd
resolved conflicts

@@
 }
-
-# merge_ds <- function(a, b) {
-#   # merge sample tables
-#   sample_cols <- intersect(colnames(a$samples), colnames(b$samples))
-#   s <- bind_rows(select(a$samples, all_of(sample_cols)),
-#                  select(b$samples, all_of(sample_cols)))
-#   
-#   # merge count data
-#   a_counts <- bind_cols(select(a$meta, probe, gene, bg), a$counts)
-#   b_counts <- bind_cols(select(b$meta, probe, gene, bg), b$counts)
-#   # Madison edit: added suffix to avoid addition of ".y" to the end of column names
-#   counts <- inner_join(a_counts, b_counts, by=c("probe", "gene", "bg"), suffix = c("", ""),
-#                        relationship="one-to-one")
-# 
-#   # break into metadata and count data
-#   meta <- select(counts, -all_of(s$aoi))
-#   counts <- select(counts, all_of(s$aoi))
-# 
-#   # make new dataset object
-#   ds <- stgeomx::init(samples=s, meta=meta, counts=counts, x=counts)
-#   return(ds)
-# }
 
@@
 # merge datasets (all annotated data)
-ds <- stgeomx::merge_ds(ds, ds_ln)
-ds <- stgeomx::merge_ds(ds, ds_colon)
-ds <- stgeomx::merge_ds(ds, ds_brain)
-ds <- stgeomx::merge_ds(ds, ds_kidney)
+ds <- stgeomx::merge(ds, ds_ln)
+ds <- stgeomx::merge(ds, ds_colon)
+ds <- stgeomx::merge(ds, ds_brain)
+ds <- stgeomx::merge(ds, ds_kidney)
 
@@
 ```{r qc filter aoi, stratification}
-ds_func <- ds_kidney
+ds_func <- ds_brain
 
@@
 # plot background correction (before and after)
-p <- stgeomx::plot_bgcorrect2(fds)
+p <- stgeomx::plot_bgcorrect(fds)
 p
@@
 # plot background correction (before and after)
-p <- stgeomx::plot_bgcorrect2(fds_panc)
+p <- stgeomx::plot_bgcorrect(fds_panc)
 p
@@
 # plot background correction (before and after)
-p <- stgeomx::plot_bgcorrect2(fds_ln)
+p <- stgeomx::plot_bgcorrect(fds_ln)
 p
@@
 # plot background correction (before and after)
-p <- stgeomx::plot_bgcorrect2(fds_kidney)
+p <- stgeomx::plot_bgcorrect(fds_kidney)
 p
@@
 # plot background correction (before and after)
-p <- stgeomx::plot_bgcorrect2(fds_colon)
+p <- stgeomx::plot_bgcorrect(fds_colon)
 p
@@
 # plot background correction (before and after)
-p <- stgeomx::plot_bgcorrect2(fds_brain)
+p <- stgeomx::plot_bgcorrect(fds_brain)
 p
@@
   pca_res <- prcomp(t(y), center=TRUE, scale.=FALSE)
-  pca_res.var <- round(100 * pca_res$sdev^2 / sum( pca_res$sdev^2 ), 2)
+  pca_res.var <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2 ), 2)
   pca_res$var_txt <- paste(colnames(pca_res$x), " (", paste(as.character(pca_res.var), "%", ")", sep=""), sep="")
@@
 ```{r bg correction and normalization methods plot}
+library("patchwork")
 correction_normalization_dir <- file.path(plot_dir, "bgcorrection_normalization")
@@
 # plot background correction
-plot_bgsub_qn <- plot_bgcorrect_helper(fds_bgsub_qn)
-plot_bgsub_rle <- plot_bgcorrect_helper(fds_bgsub_rle)
-plot_bgsub_q3 <- plot_bgcorrect_helper(fds_bgsub_q3)
-plot_bgsub_tmm <- plot_bgcorrect_helper(fds_bgsub_tmm)
-
-plot_qq_qn <- plot_bgcorrect_helper(fds_qq_qn)
-plot_qq_rle <- plot_bgcorrect_helper(fds_qq_rle)
-plot_qq_q3 <- plot_bgcorrect_helper(fds_qq_q3)
-plot_qq_tmm <- plot_bgcorrect_helper(fds_qq_tmm)
-
-plot_none_qn <- plot_bgcorrect_helper(fds_none_qn)
-plot_none_rle <- plot_bgcorrect_helper(fds_none_rle)
-plot_none_q3 <- plot_bgcorrect_helper(fds_none_q3)
-plot_none_tmm <- plot_bgcorrect_helper(fds_none_tmm)
+plot_bgsub_qn <- stgeomx::plot_bgcorrect(fds_bgsub_qn)
+plot_bgsub_rle <- stgeomx::plot_bgcorrect(fds_bgsub_rle)
+plot_bgsub_q3 <- stgeomx::plot_bgcorrect(fds_bgsub_q3)
+plot_bgsub_tmm <- stgeomx::plot_bgcorrect(fds_bgsub_tmm)
+
+plot_qq_qn <- stgeomx::plot_bgcorrect(fds_qq_qn)
+plot_qq_rle <- stgeomx::plot_bgcorrect(fds_qq_rle)
+plot_qq_q3 <- stgeomx::plot_bgcorrect(fds_qq_q3)
+plot_qq_tmm <- stgeomx::plot_bgcorrect(fds_qq_tmm)
+
+plot_none_qn <- stgeomx::plot_bgcorrect(fds_none_qn)
+plot_none_rle <- stgeomx::plot_bgcorrect(fds_none_rle)
+plot_none_q3 <- stgeomx::plot_bgcorrect(fds_none_q3)
+plot_none_tmm <- stgeomx::plot_bgcorrect(fds_none_tmm)
 
@@
 f <- file.path(correction_normalization_dir, "bgcorrect_normalization_grid")
-ggsave(paste0(f, ".png"), p, width=20, height=15)
-ggsave(paste0(f, ".pdf"), p, width=20, height=15)
+ggsave(paste0(f, ".png"), p, width=30, height=15)
+ggsave(paste0(f, ".pdf"), p, width=30, height=15)
 ```
@@
 
+### Cluster purity
+```{r purity}
+library("IntNMF")
+
+# List of methods
+methods <- list(fds_none_q3, fds_none_tmm, fds_none_rle, fds_none_qn, 
+                fds_bgsub_q3, fds_bgsub_tmm, fds_bgsub_rle, fds_bgsub_qn, 
+                fds_qq_q3, fds_qq_tmm, fds_qq_rle, fds_qq_qn)
+
+# Initialize vectors to store Chi-squared statistics and p-values
+purity_table <- numeric(length(methods))
+method_names <- c("none_q3", "none_tmm", "none_rle", "none_qn",
+                   "bgsub_q3", "bgsub_tmm", "bgsub_rle", "bgsub_qn",
+                   "qq_q3", "qq_tmm", "qq_rle", "qq_qn")
+purity_table <- NULL
+
+# Loop through each method
+for (i in methods) {
+
+  # create purity score and add to purity table
+  purity_table <- rbind(purity_table, ClusterPurity(i$samples$study, i$samples$cluster_name))
+}
+
+# Combine results into a data frame
+results_df <- data.frame(
+  Method = method_names,
+  Purity_Scores = purity_table
+ )
+
+# Print the results
+print(results_df)
+
+```
+
 ### Chi-squared analysis of clusters
 ``` {r}
+library(reshape2)
+
 # List of methods
@@
                    "qq_q3", "qq_tmm", "qq_rle", "qq_qn")
+master_contingency_table <- NULL
 
@@
   contingency_table <- table(methods[[i]]$samples$study, methods[[i]]$samples$cluster_name)
+  colnames(contingency_table) <- paste0(colnames(contingency_table), "_", method_names[i])
 
@@
   print(contingency_table)
+
+  # add to master contingency table
+  if (i == 1) {
+    master_contingency_table <- contingency_table
+  }
+  else {
+    master_contingency_table <- cbind(master_contingency_table, contingency_table)
+  }
 
@@
 }
-
-# Format p-values to scientific notation
-p_values_formatted <- format(p_values, scientific = TRUE)
 
@@
 
+print(master_contingency_table)
+# heatmap(master_contingency_table)
+
+# # Dummy data
+# x <- LETTERS[1:20]
+# y <- paste0("var", seq(1,20))
+# data <- expand.grid(X=x, Y=y)
+# data$Z <- runif(400, 0, 5)
+
+# # Heatmap 
+# master_contingency_table <- melt(master_contingency_table)
+# ggplot(master_contingency_table, aes(x = Var1, y = Var2, fill = value)) +
+# geom_tile()
+
 ```
@@
 ```{r}
+library("limma")
+library("writexl")
+
 de_setup <- function(method_name, de_padj_cutoff=0.05, de_log2fc_cutoff=1) {
@@
   write_xlsx(res, file.path(results_dir, file_name))
-  return(res)
+  de <- bind_rows(de, res)
+  return(de)
 }
@@
 
-# res_none_q3 <- de_setup("none_q3")
-# res_none_tmm <- de_setup("none_tmm")
-# res_none_rle <- de_setup("none_rle")
-# res_none_qn <- de_setup("none_qn")
-# 
-# res_bgsub_q3 <- de_setup("bgsub_q3")
-# res_bgsub_tmm <- de_setup("bgsub_tmm")
-# res_bgsub_rle <- de_setup("bgsub_rle")
-# res_bgsub_qn <- de_setup("bgsub_qn")
-# 
-# res_qq_q3 <- de_setup("qq_q3")
-# res_qq_tmm <- de_setup("qq_tmm")
-# res_qq_rle <- de_setup("qq_rle")
-# res_qq_qn <- de_setup("qq_qn")
+de_none_q3 <- de_setup("none_q3")
+de_none_tmm <- de_setup("none_tmm")
+de_none_rle <- de_setup("none_rle")
+de_none_qn <- de_setup("none_qn")
+
+de_bgsub_q3 <- de_setup("bgsub_q3")
+de_bgsub_tmm <- de_setup("bgsub_tmm")
+de_bgsub_rle <- de_setup("bgsub_rle")
+de_bgsub_qn <- de_setup("bgsub_qn")
+
+de_qq_q3 <- de_setup("qq_q3")
+de_qq_tmm <- de_setup("qq_tmm")
+de_qq_rle <- de_setup("qq_rle")
+de_qq_qn <- de_setup("qq_qn")
 
 # write tables
-# table(res_none_q3$contrast, res_none_q3$de)
+table(de_none_q3$contrast, de_none_q3$de)
 
@@
 de <- NULL
-de <- bind_rows(de, res_qq_qn)
+de <- de_qq_qn
 
@@
 ``` {r gsea}
-# read msigdb data
-f <- file.path(data_dir, "msigdb.v2023.2.Hs.json")
+gsea_nperms <- 10000
+
+read_msigdb_json_file <- function(filename) {
+  x <- jsonlite::fromJSON(filename)
+  y <- tibble(gs_name = names(x), gs = x) %>% 
+    unnest_wider(gs) %>%
+    unnest_longer(geneSymbols) %>%
+    select(gs_cat = collection,
+           gs_name = gs_name,
+           gene_symbol = geneSymbols)
+}
+
+# read msigdb data: hallmark
+f <- system.file("extdata", "h.all.v2024.1.Hs.json", package = "stgeomx")
 msigdb_gene_sets <- read_msigdb_json_file(f)
 # determine gene universe
-msigdb_genes <- intersect(fds$meta$gene, msigdb_gene_sets$gene_symbol)
+msigdb_genes <- intersect(fds_qq_qn$meta$gene, msigdb_gene_sets$gene_symbol)
 # filter genes not in our dataset
 msigdb_gene_sets <- msigdb_gene_sets %>%
-  dplyr::filter(gene_symbol %in% fds$meta$gene) %>%
+  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
   distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
-
 # gene set categories
@@
 
-# pass log fold change into gsea with hallmark gene sets (sending new code)
-
-get_de_ranks <- function(xde, mycontrast) {
+# read msigdb data: pancreas_endocrine
+f <- system.file("extdata", "GOBP_ENDOCRINE_PANCREAS_DEVELOPMENT.v2024.1.Hs.json", package = "stgeomx")
+gobp_end_pancreas_gene_sets <- read_msigdb_json_file(f)
+# determine gene universe
+gobp_end_pancreas_genes <- intersect(fds_qq_qn$meta$gene, gobp_end_pancreas_gene_sets$gene_symbol)
+# filter genes not in our dataset
+gobp_end_pancreas_gene_sets <- gobp_end_pancreas_gene_sets %>%
+  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
+  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
+# gene set categories
+gs_end_panc <- dplyr::filter(gobp_end_pancreas_gene_sets, gs_cat == "C5:GO:BP")
+gs_end_panc <- split(x = gs_end_panc$gene_symbol, f = gs_end_panc$gs_name)
+
+# read msigdb data: pancreas secretion
+f <- system.file("extdata", "GOBP_INSULIN_SECRETION.v2024.1.Hs.json", package = "stgeomx")
+gobp_pancreas_secretion_gene_sets <- read_msigdb_json_file(f)
+# determine gene universe
+gobp_pancreas_secretion_genes <- intersect(fds_qq_qn$meta$gene, gobp_pancreas_secretion_gene_sets$gene_symbol)
+# filter genes not in our dataset
+gobp_pancreas_secretion_gene_sets <- gobp_pancreas_secretion_gene_sets %>%
+  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
+  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
+# gene set categories
+gs_panc_sec <- dplyr::filter(gobp_pancreas_secretion_gene_sets, gs_cat == "C5:GO:BP")
+gs_panc_sec <- split(x = gs_panc_sec$gene_symbol, f = gs_panc_sec$gs_name)
+
+# read msigdb data: brain development gene sets
+f <- system.file("extdata", "brain_genesets.v2024.1.Hs.json", package = "stgeomx")
+gobp_brain_gene_sets <- read_msigdb_json_file(f)
+# determine gene universe
+gobp_brain_genes <- intersect(fds_qq_qn$meta$gene, gobp_brain_gene_sets$gene_symbol)
+# filter genes not in our dataset
+gobp_brain_gene_sets <- gobp_brain_gene_sets %>%
+  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
+  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
+# gene set categories
+gs_brain <- dplyr::filter(gobp_brain_gene_sets, gs_cat == "C5:GO:BP")
+gs_brain <- split(x = gs_brain$gene_symbol, f = gs_brain$gs_name)
+
+# read msigdb data: ln development gene sets
+f <- system.file("extdata", "ln_genesets.v2024.1.Hs.json", package = "stgeomx")
+gobp_ln_gene_sets <- read_msigdb_json_file(f)
+# determine gene universe
+gobp_ln_genes <- intersect(fds_qq_qn$meta$gene, gobp_ln_gene_sets$gene_symbol)
+# filter genes not in our dataset
+gobp_ln_gene_sets <- gobp_ln_gene_sets %>%
+  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
+  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
+# gene set categories
+gs_ln <- dplyr::filter(gobp_ln_gene_sets, gs_cat == "C5:GO:BP")
+gs_ln <- split(x = gs_ln$gene_symbol, f = gs_ln$gs_name)
+
+# read msigdb data: kidney development gene sets
+f <- system.file("extdata", "kidney_genesets.v2024.1.Hs.json", package = "stgeomx")
+gobp_kidney_gene_sets <- read_msigdb_json_file(f)
+# determine gene universe
+gobp_kidney_genes <- intersect(fds_qq_qn$meta$gene, gobp_kidney_gene_sets$gene_symbol)
+# filter genes not in our dataset
+gobp_kidney_gene_sets <- gobp_kidney_gene_sets %>%
+  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
+  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
+# gene set categories
+gs_kidney <- dplyr::filter(gobp_kidney_gene_sets, gs_cat == "C5:GO:BP")
+gs_kidney <- split(x = gs_kidney$gene_symbol, f = gs_kidney$gs_name)
+
+# read msigdb data: colon development gene sets
+f <- system.file("extdata", "colon_genesets.v2024.1.Hs.json", package = "stgeomx")
+gobp_colon_gene_sets <- read_msigdb_json_file(f)
+# determine gene universe
+gobp_colon_genes <- intersect(fds_qq_qn$meta$gene, gobp_colon_gene_sets$gene_symbol)
+# filter genes not in our dataset
+gobp_colon_gene_sets <- gobp_colon_gene_sets %>%
+  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
+  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
+# gene set categories
+gs_colon <- dplyr::filter(gobp_colon_gene_sets, gs_cat == "C5:GO:BP")
+gs_colon <- split(x = gs_colon$gene_symbol, f = gs_colon$gs_name)
+
+# new gsea code
+get_fgsea_ranks_de <- function(xde, mycontrast) {
   ranks <- xde %>%
@@
 
-plot_gsea_enrichment <- function(x, ranks, a, gs) {
-  txt_stat <- paste0("ES=", round(x$ES,2), " NES=", round(x$NES, 2), " padj=", format(x$padj, scientific=TRUE, digits=3))
-  txt_title <- paste0("ranks: ", a, " gs: ", x$pathway)
-  p <- plotEnrichment(gs[[x$pathway]], ranks) +
-    annotate(geom="text", x=150, y=0.1, label=txt_stat, hjust=0) +
-    labs(title = txt_title, 
-         xlab = "Rank",
-         ylab = "Enrichment Score") +
-    theme(plot.title = element_text(size=8))
-  return(p)
-}
-
-# fgsea_run_batch <- function(xde, xgs, 
-#                             fgsea_eps=0,
-#                             fgsea_nperms=10000, 
-#                             fgsea_nproc=0) {
-#   fgsea_scoretypes <- c("pos", "neg")
-#   gsea <- NULL
-#   for (xcontrast in unique(xde$contrast)) {
-#     ranks <- get_de_ranks(xde, xcontrast)
-#     for (fgsea_scoretype in fgsea_scoretypes) {
+fgsea_run_batch <- function(xde, xgs,
+                            fgsea_eps=0,
+                            fgsea_nperms=1000) {
+  gsea <- NULL
+  for (xcontrast in unique(xde$contrast)) {
+    ranks <- get_fgsea_ranks_de(xde, xcontrast)
     res <- fgsea(pathways=xgs,
@@
                  eps=fgsea_eps,
-                   scoreType = fgsea_scoretype,
-                   nPermSimple=fgsea_nperms, 
-                   nproc=fgsea_nproc)
-      res <- mutate(res, 
-                    contrast = xcontrast, 
-                    dir = fgsea_scoretype)
+                 nPermSimple=fgsea_nperms)
+    res <- mutate(res, contrast = xcontrast)
     gsea <- bind_rows(gsea, res)     
-#     }
-#   }
-#   return(gsea)
-# }
-
-#
+  }
+  return(gsea)
+}
+
+stgeomx_gsea_pipeline <- function(gsea, de, method) {
+  res <- fgsea_run_batch(de, gs_hallmark,
+                       fgsea_eps=0, fgsea_nperms=gsea_nperms)
+  res <- res %>% mutate(method = method)
+  gsea <- bind_rows(gsea, res)
+
+  res <- fgsea_run_batch(de, gs_ex_panc,
+                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
+  res <- res %>% mutate(method = method)
+  gsea <- bind_rows(gsea, res)
+
+  res <- fgsea_run_batch(de, gs_end_panc,
+                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
+  res <- res %>% mutate(method = method)
+  gsea <- bind_rows(gsea, res)
+
+  res <- fgsea_run_batch(de, gs_panc_sec,
+                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
+  res <- res %>% mutate(method = method)
+  gsea <- bind_rows(gsea, res)
+
+  res <- fgsea_run_batch(de, gs_brain,
+                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
+  res <- res %>% mutate(method = method)
+  gsea <- bind_rows(gsea, res)
+
+  res <- fgsea_run_batch(de, gs_ln,
+                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
+  res <- res %>% mutate(method = method)
+  gsea <- bind_rows(gsea, res)
+
+  res <- fgsea_run_batch(de, gs_kidney,
+                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
+  res <- res %>% mutate(method = method)
+  gsea <- bind_rows(gsea, res)
+
+  res <- fgsea_run_batch(de, gs_colon,
+                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
+  res <- res %>% mutate(method = method)
+  gsea <- bind_rows(gsea, res)
+
+  # filter out irrelevant genes in hallmark
+  gsea <- gsea %>%
+  filter(!(pathway %in% c("HALLMARK_ESTROGEN_RESPONSE_LATE", "HALLMARK_ESTROGEN_RESPONSE_EARLY", "HALLMARK_GLYCOLYSIS", "HALLMARK_P53_PATHWAY", "HALLMARK_HEME_METABOLISM", "HALLMARK_ADIPOGENESIS", "HALLMARK_MYC_TARGETS_V2", "HALLMARK_MYC_TARGETS_V1", "HALLMARK_OXIDATIVE_PHOSPHORYLATION", "HALLMARK_HYPOXIA", "HALLMARK_UV_RESPONSE_DN", "HALLMARK_MYOGENESIS", "HALLMARK_APICAL_SURFACE", "HALLMARK_APOPTOSIS", "HALLMARK_IL6_JAK_STAT3_SIGNALING", "HALLMARK_IL2_STAT5_SIGNALING", "HALLMARK_MITOTIC_SPINDLE", "HALLMARK_APICAL_JUNCTION", "HALLMARK_KRAS_SIGNALING_UP", "HALLMARK_KRAS_SIGNALING_DN", "HALLMARK_UNFOLDED_PROTEIN_RESPONSE", "HALLMARK_DNA_REPAIR", "HALLMARK_FATTY_ACID_METABOLISM", "HALLMARK_HEDGEHOG_SIGNALING", "HALLMARK_ANGIOGENESIS", "HALLMARK_E2F_TARGETS", "HALLMARK_G2M_CHECKPOINT", "HALLMARK_TNFA_SIGNALING_VIA_NFKB", "HALLMARK_PROTEIN_SECRETION", "HALLMARK_CHOLESTEROL_HOMEOSTASIS")))
+
+  return(gsea)
+}
+
 # Run full analysis
-#
-gsea <- NULL
-
-# hallmark
-res <- fgsea_run_batch(de, gs_hallmark, 
-                       fgsea_eps=0, fgsea_nperms=gsea_nperms, fgsea_nproc=10)
-gsea <- bind_rows(gsea, res)
-gsea_all <- gsea
-```
-
-# Downstream data analysis 
-## Highly variable genes: updated
-```{r updated highly variable genes, include=FALSE}
-most_var_genes <- function(x, ntop=500, span=0.50) {
-  tbl <- bind_cols(
-    u = rowMeans(x),
-    v = apply(x, 1, function(x) var(x))
-  )
-  lovar <- loess(v ~ u, data=tbl, span=span)
-  tbl$lovar <- predict(lovar)
-  tbl$vadj <- tbl$v - tbl$lovar
-  keep <- order(tbl$vadj, decreasing=TRUE)[seq_len(min(ntop, nrow(tbl)))]
-  return(x[keep,])
-}
-
-maplot <- function(y, ntop=1000, nlabel=100, myspan=0.5) {
-  tbl <- bind_cols(
-    g = rownames(y),
-    u = rowMeans(y),
-    v = apply(y, 1, function(x) var(x))
-  )
-  lovar <- loess(v ~ u, data=tbl, span=myspan)
-  tbl$lovar <- predict(lovar)
-  tbl$vadj <- tbl$v - tbl$lovar
-  tbl <- tbl %>% 
-    mutate(rank = min_rank(-vadj)) %>%
-    arrange(rank)
-  label_data <- filter(tbl, rank <= nlabel)
-  top_data <- filter(tbl, rank < ntop)
-  bot_data <- filter(tbl, rank >= ntop)
-  p <- ggplot(tbl) +
-    geom_point(data=bot_data, aes(u, v), alpha=0.5, size=1, color="#aaaaaa") +
-    geom_point(data=top_data, aes(u, v), alpha=0.8, size=2, color="#ffaaaa") +
-    geom_line(aes(u, lovar), color="blue") +
-    geom_text_repel(data=label_data, aes(u, v, label=g), color="black", size=3, max.overlaps=Inf) +
-    theme_minimal()
-  return(p)
-}
-
-heatmap_genes <- function(s, x, filename="heatmap.pdf", width=10, height=10) {
-  s <- s %>% mutate(
-    noise_quantile = ntile(desc(bg_auc), 8)
-  )
-  annot_col <- column_to_rownames(s, "aoi") %>%
-    select(patient, path_specimen, path_aoi, noise_quantile) %>%
-    as.data.frame()
-  p <- pheatmap(x,
-                scale="row",
-                show_rownames=TRUE,
-                show_colnames=FALSE,
-                border_color=NA,
-                clustering_distance_rows = "euclidean",
-                clustering_distance_cols = "euclidean",
-                clustering_method = "ward.D2",
-                # clustering_method = "average",
-                # clustering_distance_rows = "correlation",
-                # clustering_distance_cols = "correlation",
-                breaks = seq(-3, 3, 0.05),
-                color = colorRampPalette(c("purple3", "black", "yellow2"))(120),
-                annotation_col = annot_col,
-                # annotation_colors = color_scales,
-                fontsize=4,
-                filename=filename,
-                width=width,
-                height=height)
-  return(p)
-}
-
-y <- log2_norm_cpm_merged
-ntop <- 1000
-nlabel <- 100
-myspan <- 0.5
-
-# maplot
-p <- maplot(log2_norm_cpm_merged, ntop=ntop, nlabel=nlabel, myspan=myspan)
-f <- file.path(plot_dir, paste0("maplot"))
-ggsave(paste0(f, ".pdf"), plot=p, width=10, height=10)
-ggsave(paste0(f, ".png"), plot=p, width=10, height=10)
-
-# select most variable genes
-ntop <- 500
-myspan <- 0.5
-s <- fds$samples
-y <- most_var_genes(log2_norm_cpm_merged, ntop=ntop, span=myspan)
-
-# heatmap
-f <- file.path(plot_dir, "heatmap_highvar_genes.pdf")
-p <- heatmap_genes(s, y, f)
-```
-
-## PCA: updated
-```{r updated pca}
-# subset most variable genes
-ntop <- 250
-myspan <- 0.5
-s <- fds$samples
-y <- most_var_genes(log2_norm_cpm_merged, ntop=ntop, span=myspan)
-
-# plot dimensions
-width <- 12
-height <- 10
-
-# PCA
-npcs <- 50
-pca_res <- prcomp(t(y), center=TRUE, scale.=TRUE, rank.=npcs)
-pca_res.var <- round(100 * pca_res$sdev^2 / sum( pca_res$sdev^2 ), 2)
-pca_res.var_txt <- paste(colnames(pca_res$x), " (", paste(as.character(pca_res.var), "%", ")", sep=""), sep="")
-
-# fviz plots
-fviz_ntop <- 50
-
-# percent variance of each PC
-p <- fviz_eig(pca_res, col.var="blue", addlabels=TRUE)
-f <- file.path(plot_dir, paste0("pca_fviz_eig.pdf"))
-ggsave(f, p)
-
-# contributions of variables to each PC
-p <- fviz_contrib(pca_res, choice="var", axes=1, top=fviz_ntop) +
-  theme(axis.text.x = element_text(angle=90))
-f <- file.path(plot_dir, paste0("pca_fviz_contrib1.pdf"))
-ggsave(f, p)
-
-p <- fviz_contrib(pca_res, choice="var", axes=2, top=fviz_ntop) +
-  theme(axis.text.x = element_text(angle=90))
-f <- file.path(plot_dir, paste0("pca_fviz_contrib2.pdf"))
-ggsave(f, p)
-
-p <- fviz_contrib(pca_res, choice="var", axes=3, top=fviz_ntop) +
-  theme(axis.text.x = element_text(angle=90))
-f <- file.path(plot_dir, paste0("pca_fviz_contrib3.pdf"))
-ggsave(f, p)
-
-# contribution to first 3 PCs
-p <- fviz_contrib(pca_res, choice="var", axes = 1:3, top=fviz_ntop) +
-  theme(axis.text.x = element_text(angle=90))
-f <- file.path(plot_dir, paste0("pca_fviz_contrib123.pdf"))
-ggsave(f, p)
-
-# graph of variables contributing to PCA
-p <- fviz_pca_var(
-  pca_res,
-  axes = c(1,2),
-  col.var = "contrib", # Color by the quality of representation
-  gradient.cols = c("darkorchid4", "gold", "darkorange"),
-  repel=TRUE,
-  select.var = list(contrib=fviz_ntop)
-)
-f <- file.path(plot_dir, paste0("pca_fviz_pca_var_pc12.pdf"))
-ggsave(f, p, width=7, height=7)
-
-# graph of variables contributing to PCA
-p <- fviz_pca_var(
-  pca_res,
-  axes = c(1,3),
-  col.var = "contrib", # Color by the quality of representation
-  gradient.cols = c("darkorchid4", "gold", "darkorange"),
-  repel=TRUE,
-  select.var = list(contrib=fviz_ntop)
-)
-f <- file.path(plot_dir, paste0("pca_fviz_pca_var_pc13.pdf"))
-ggsave(f, p, width=7, height=7)
-
-# graph of variables contributing to PCA
-p <- fviz_pca_var(
-  pca_res,
-  axes = c(2,3),
-  col.var = "contrib", # Color by the quality of representation
-  gradient.cols = c("darkorchid4", "gold", "darkorange"),
-  repel=TRUE,
-  select.var = list(contrib=fviz_ntop)
-)
-f <- file.path(plot_dir, paste0("pca_fviz_pca_var_pc23.pdf"))
-ggsave(f, p, width=7, height=7)
-
-#
-# PCA plots
-#
-x <- bind_cols(s, as_tibble(pca_res$x))
-
-# bg_ks_dist
-p <- ggplot(x, aes(x=PC1, y=PC2, color=bg_auc)) +
-  geom_point(alpha = 0.8, size=3) +
-  scale_color_viridis_c() +
+gsea_all <- NULL
+gsea_none_q3 <- NULL
+gsea_none_tmm <- NULL
+gsea_none_rle <- NULL
+gsea_none_qn <- NULL
+
+gsea_bgsub_q3 <- NULL
+gsea_bgsub_tmm <- NULL
+gsea_bgsub_rle <- NULL
+gsea_bgsub_qn <- NULL
+
+gsea_qq_q3 <- NULL
+gsea_qq_tmm <- NULL
+gsea_qq_rle <- NULL
+gsea_qq_qn <- NULL
+
+# run stgeomx pipeline
+gsea_none_q3 <- stgeomx_gsea_pipeline(gsea_none_q3, de_none_q3, "none_q3")
+gsea_none_tmm <- stgeomx_gsea_pipeline(gsea_none_tmm, de_none_tmm, "none_tmm")
+gsea_none_rle <- stgeomx_gsea_pipeline(gsea_none_rle, de_none_rle, "none_rle")
+gsea_none_qn <- stgeomx_gsea_pipeline(gsea_none_qn, de_none_qn, "none_qn")
+
+gsea_bgsub_q3 <- stgeomx_gsea_pipeline(gsea_bgsub_q3, de_bgsub_q3, "bgsub_q3")
+gsea_bgsub_tmm <- stgeomx_gsea_pipeline(gsea_bgsub_tmm, de_bgsub_tmm, "bgsub_tmm")
+gsea_bgsub_rle <- stgeomx_gsea_pipeline(gsea_bgsub_rle, de_bgsub_rle, "bgsub_rle")
+gsea_bgsub_qn <- stgeomx_gsea_pipeline(gsea_bgsub_qn, de_bgsub_qn, "bgsub_qn")
+
+gsea_qq_q3 <- stgeomx_gsea_pipeline(gsea_qq_q3, de_qq_q3, "qq_q3")
+gsea_qq_tmm <- stgeomx_gsea_pipeline(gsea_qq_tmm, de_qq_tmm, "qq_tmm")
+gsea_qq_rle <- stgeomx_gsea_pipeline(gsea_qq_rle, de_qq_rle, "qq_rle")
+gsea_qq_qn <- stgeomx_gsea_pipeline(gsea_qq_qn, de_qq_qn, "qq_qn")
+
+# # write results for clusters
+# contrasts <- c("clusterPanc", "clusterBrain", "clusterLn", "clusterKidney", "clusterColon")
+# manuscript_pathways <- c(names(gs_hallmark))
+# res <- gsea_all %>% filter(contrast %in% contrasts, 
+#                            pathway %in% manuscript_pathways)
+
+# prefix <- "gsea_all"
+# f <- file.path(results_dir, paste0(prefix, "_results.tsv"))
+# data.table::fwrite(res, file=f, sep="\t", sep2=c("", " ", ""))
+# f <- file.path(results_dir, paste0(prefix, "_results.xlsx"))
+# write_xlsx(res, f)
+```
+### GSEA plots: helper functions
+```{r helpers}
+gsea_padj_cutoff <- 0.01
+gsea_log2fc_cutoff <- 1
+
+# contrasts
+contrasts <- c("clusterPanc", "clusterBrain", "clusterLn", "clusterKidney", "clusterColon")
+
+prep_gsea <- function(gsea) {
+  # prepare gsea results
+  gsea <- gsea %>%
+    filter(contrast %in% contrasts) %>%
+    mutate(padj = 2 * padj) %>%
+    mutate(contrast = factor(contrast, levels=contrasts),
+           neglog10padj = ifelse(padj < gsea_padj_cutoff, -log10(padj), NA),
+           sig = padj < gsea_padj_cutoff) %>%
+    group_by(contrast, pathway) %>%
+    slice_min(padj, n=1, with_ties=FALSE) %>%
+    ungroup()
+  return(gsea)
+}
+
+prep_hallmark <- function(gsea) {
+  # Hallmark
+  res <- gsea %>% filter(pathway %in% names(gs_hallmark)) %>%
+    mutate(pathway = gsub("HALLMARK_", "", pathway))
+  # only keep significant results (at least one contrast)
+  res <- res %>% group_by(pathway) %>%
+    filter(any(padj < gsea_padj_cutoff)) %>%
+    ungroup()
+  return(res)
+}
+
+prep_all <- function(gsea) {
+  # remove pathway prefixes
+  res <- gsea  %>% mutate(pathway = str_replace(pathway, "^HALLMARK_|^GOBP_", ""))
+  # only keep significant results (at least one contrast)
+  res <- res %>% group_by(pathway) %>%
+    filter(any(padj < gsea_padj_cutoff)) %>%
+    ungroup()
+  return(res)
+}
+```
+
+### GSEA plots: by cluster type
+
+```{r gsea plots}
+cluster = "clusterColon"
+
+plot_cluster <- function(res, pathway_levels, cluster) {
+  res_panc <- res %>% filter(contrast == cluster)
+
+  # Set the pathway levels to be the same as the reference (res_qq_qn)
+  res_panc$pathway <- factor(res_panc$pathway, levels=pathway_levels)
+
+  # Filter out any NA pathways
+  res_panc <- res_panc %>% filter(!is.na(pathway))
+
+  p <- ggplot(res_panc, aes(x=NES, y=pathway, fill=neglog10padj)) +
+    geom_col() +
+    scale_fill_viridis_c() +
     theme_bw() +
-  theme(aspect.ratio=1) +
-  theme(legend.text = element_text(size=7)) +
-  labs(x=pca_res.var_txt[1], y=pca_res.var_txt[2])
-f <- file.path(plot_dir, paste0("pca_bg_auc.pdf"))
-ggsave(f, plot=p, device="pdf", width=width, height=height)
-
-# patient
-p <- ggplot(x, aes(x=PC1, y=PC2, color=patient, shape=study)) +
-  geom_point(alpha = 0.8, size=3) +
-  # scale_color_manual(values = color_scales$patient) +
+    theme(axis.text.y = element_text(color="black", size=7)) +
+    theme(strip.text = element_text(size=8)) +
+    facet_wrap(~ contrast, nrow=1) +
+    labs(x="Normalized Enrichment Score (NES)", y="Pathway")
+
+  return (p)
+}
+
+# Prepare and process qq_qn data
+gsea_qq_qn <- prep_gsea(gsea_qq_qn)
+res_qq_qn <- prep_all(gsea_qq_qn)
+
+# Extract pathway levels from res_qq_qn to use for all plots
+pathway_levels <- res_qq_qn %>% filter(contrast == cluster) %>% arrange(NES) %>% pull(pathway)
+
+# Generate qq_qn plot using consistent pathway levels
+p_qq_qn <- plot_cluster(res_qq_qn, pathway_levels, cluster)
+
+# Prepare and process none_q3 data
+gsea_none_q3 <- prep_gsea(gsea_none_q3)
+res_none_q3 <- prep_all(gsea_none_q3)
+p_none_q3 <- plot_cluster(res_none_q3, pathway_levels, cluster)
+
+gsea_none_tmm <- prep_gsea(gsea_none_tmm)
+res_none_tmm <- prep_all(gsea_none_tmm)
+p_none_tmm <- plot_cluster(res_none_tmm, pathway_levels, cluster)
+
+gsea_none_rle <- prep_gsea(gsea_none_rle)
+res_none_rle <- prep_all(gsea_none_rle)
+p_none_rle <- plot_cluster(res_none_rle, pathway_levels, cluster)
+
+gsea_none_qn <- prep_gsea(gsea_none_qn)
+res_none_qn <- prep_all(gsea_none_qn)
+p_none_qn <- plot_cluster(res_none_qn, pathway_levels, cluster)
+
+gsea_bgsub_q3 <- prep_gsea(gsea_bgsub_q3)
+res_bgsub_q3 <- prep_all(gsea_bgsub_q3)
+p_bgsub_q3 <- plot_cluster(res_bgsub_q3, pathway_levels, cluster)
+
+gsea_bgsub_tmm <- prep_gsea(gsea_bgsub_tmm)
+res_bgsub_tmm <- prep_all(gsea_bgsub_tmm)
+p_bgsub_tmm <- plot_cluster(res_bgsub_tmm, pathway_levels, cluster)
+
+gsea_bgsub_rle <- prep_gsea(gsea_bgsub_rle)
+res_bgsub_rle <- prep_all(gsea_bgsub_rle)
+p_bgsub_rle <- plot_cluster(res_bgsub_rle, pathway_levels, cluster)
+
+gsea_bgsub_qn <- prep_gsea(gsea_bgsub_qn)
+res_bgsub_qn <- prep_all(gsea_bgsub_qn)
+p_bgsub_qn <- plot_cluster(res_bgsub_qn, pathway_levels, cluster)
+
+# Prepare and process qq
+gsea_qq_q3 <- prep_gsea(gsea_qq_q3)
+res_qq_q3 <- prep_all(gsea_qq_q3)
+p_qq_q3 <- plot_cluster(res_qq_q3, pathway_levels, cluster)
+
+gsea_qq_tmm <- prep_gsea(gsea_qq_tmm)
+res_qq_tmm <- prep_all(gsea_qq_tmm)
+p_qq_tmm <- plot_cluster(res_qq_tmm, pathway_levels, cluster)
+
+gsea_qq_rle <- prep_gsea(gsea_qq_rle)
+res_qq_rle <- prep_all(gsea_qq_rle)
+p_qq_rle <- plot_cluster(res_qq_rle, pathway_levels, cluster)
+
+gsea_qq_qn <- prep_gsea(gsea_qq_qn)
+res_qq_qn <- prep_all(gsea_qq_qn)
+p_qq_qn <- plot_cluster(res_qq_qn, pathway_levels, cluster)
+```
+
+### GSEA combined plot: cluster type
+``` {r GSEA combined plot}
+gsea_dir <- file.path(plot_dir, "gsea")
+
+# Grid column and row titles
+col_titles <- list(
+  wrap_elements(grid::textGrob('', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
+  wrap_elements(grid::textGrob('Q3', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
+  wrap_elements(grid::textGrob('TMM', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
+  wrap_elements(grid::textGrob('RLE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
+  wrap_elements(grid::textGrob('QN', gp = grid::gpar(fontsize = 18, fontface = "bold")))
+)
+
+row_titles <- list(
+  wrap_elements(grid::textGrob('NONE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
+  wrap_elements(grid::textGrob('BGSub', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
+  wrap_elements(grid::textGrob('QQ', gp = grid::gpar(fontsize = 18, fontface = "bold")))
+)
+
+# Create grid
+p <- wrap_plots(
+  col_titles[[1]], col_titles[[2]], col_titles[[3]], col_titles[[4]], col_titles[[5]],
+  row_titles[[1]], p_none_q3, p_none_tmm, p_none_rle, p_none_qn,
+  row_titles[[2]], p_bgsub_q3, p_bgsub_q3, p_bgsub_q3, p_bgsub_q3,
+  row_titles[[3]], p_qq_q3, p_qq_q3, p_qq_q3, p_qq_qn,
+  ncol = 5,
+  heights = c(0.1, 1, 1, 1),
+  widths = c(0.3, 1, 1, 1, 1)
+) +
+plot_layout(
+  guides = 'collect'
+) &
+theme(plot.margin = unit(c(0.5, 1, 1, 1), "cm"))
+
+# Add titles
+p <- p + 
+  plot_annotation(
+    title = "Effect of Background Correction and Normalization Methods on PCA"
+  ) & 
+  theme(
+    plot.title = element_text(size = 22, face = "bold", hjust = 0.5)
+  )
+
+f <- file.path(gsea_dir, "gsea_colon")
+ggsave(paste0(f, ".png"), p, width=35, height=20)
+ggsave(paste0(f, ".pdf"), p, width=35, height=20)
+```
+
+### GSEA plots: by bg + normalization
+``` {r gsea bg_norm}
+gsea_dir <- file.path(plot_dir, "gsea")
+
+contrasts <- c("clusterPanc", "clusterBrain", "clusterLn", "clusterKidney", "clusterColon")
+
+plot_clusters <- function(res) {
+  res_panc <- res %>% filter(contrast == "clusterPanc")
+  pathway_levels <- res_panc %>% filter(contrast == "clusterPanc") %>% arrange(NES) %>% pull(pathway)
+  res_panc$pathway <- factor(res_panc$pathway, levels=pathway_levels)
+
+  res_brain <- res %>% filter(contrast == "clusterBrain")
+  pathway_levels <- res_brain %>% filter(contrast == "clusterBrain") %>% arrange(NES) %>% pull(pathway)
+  res_brain$pathway <- factor(res_brain$pathway, levels=pathway_levels)
+
+  res_ln <- res %>% filter(contrast == "clusterLn")
+  pathway_levels <- res_ln %>% filter(contrast == "clusterLn") %>% arrange(NES) %>% pull(pathway)
+  res_ln$pathway <- factor(res_ln$pathway, levels=pathway_levels)
+
+  res_kidney <- res %>% filter(contrast == "clusterKidney")
+  pathway_levels <- res_kidney %>% filter(contrast == "clusterKidney") %>% arrange(NES) %>% pull(pathway)
+  res_kidney$pathway <- factor(res_kidney$pathway, levels=pathway_levels)
+
+  res_colon <- res %>% filter(contrast == "clusterColon")
+  pathway_levels <- res_colon %>% filter(contrast == "clusterColon") %>% arrange(NES) %>% pull(pathway)
+  res_colon$pathway <- factor(res_colon$pathway, levels=pathway_levels)
+
+  p_panc <- ggplot(res_panc, aes(x=NES, y=pathway, fill=neglog10padj)) +
+    geom_col() +
+    scale_fill_viridis_c() +
     theme_bw() +
-  theme(aspect.ratio=1) +
-  theme(legend.text = element_text(size=7)) +
-  labs(x=pca_res.var_txt[1], y=pca_res.var_txt[2])
-f <- file.path(plot_dir, paste0("pca_patient.pdf"))
-ggsave(f, plot=p, device="pdf", width=width, height=height)
-
-# region
-p <- ggplot(x, aes(x=PC1, y=PC2, color=region, shape=study)) +
-  geom_point(alpha = 0.8, size=3) +
-  # scale_color_manual(values = color_scales$patient) +
+    theme(axis.text.y = element_text(color="black", size=7)) +
+    theme(strip.text = element_text(size=8)) +
+    facet_wrap(~ contrast, nrow=1) +
+    labs(x="Normalized Enrichment Score (NES)", y="Pathway")
+  p_brain <- ggplot(res_brain, aes(x=NES, y=pathway, fill=neglog10padj)) +
+    geom_col() +
+    scale_fill_viridis_c() +
     theme_bw() +
-  theme(aspect.ratio=1) +
-  theme(legend.text = element_text(size=7)) +
-  labs(x=pca_res.var_txt[1], y=pca_res.var_txt[2])
-f <- file.path(plot_dir, paste0("pca_region.pdf"))
-ggsave(f, plot=p, device="pdf", width=width, height=height)
-
-# subset
-p <- ggplot(x, aes(x=PC1, y=PC2, color=subset, shape=study)) +
-  geom_point(alpha = 0.8, size=3) +
-  # scale_color_manual(values = color_scales$patient) +
+    theme(axis.text.y = element_text(color="black", size=7)) +
+    theme(strip.text = element_text(size=8)) +
+    facet_wrap(~ contrast, nrow=1) +
+    labs(x="Normalized Enrichment Score (NES)", y="Pathway")
+  p_ln <- ggplot(res_ln, aes(x=NES, y=pathway, fill=neglog10padj)) +
+    geom_col() +
+    scale_fill_viridis_c() +
     theme_bw() +
-  theme(aspect.ratio=1) +
-  theme(legend.text = element_text(size=7)) +
-  labs(x=pca_res.var_txt[1], y=pca_res.var_txt[2])
-f <- file.path(plot_dir, paste0("pca_subset.pdf"))
-ggsave(f, plot=p, device="pdf", width=width, height=height)
-
-# study
-p <- ggplot(x, aes(x=PC1, y=PC2, color=study)) +
-  geom_point(alpha = 0.8, size=3) +
-  # scale_color_manual(values=color_scales$path) +
+    theme(axis.text.y = element_text(color="black", size=7)) +
+    theme(strip.text = element_text(size=8)) +
+    facet_wrap(~ contrast, nrow=1) +
+    labs(x="Normalized Enrichment Score (NES)", y="Pathway")
+  p_kidney <- ggplot(res_kidney, aes(x=NES, y=pathway, fill=neglog10padj)) +
+    geom_col() +
+    scale_fill_viridis_c() +
     theme_bw() +
-  theme(aspect.ratio=1) +
-  theme(legend.text = element_text(size=7)) +
-  labs(x=pca_res.var_txt[1], y=pca_res.var_txt[2])
-f <- file.path(plot_dir, paste0("pca_study.pdf"))
-ggsave(f, plot=p, device="pdf", width=width, height=height)
-
-#
-# Plots of PC1, PC2, and PC3
-#
-pca_plot <- function(x, xvar, yvar, color_by, shape_by) {
-  color_by_string <- rlang::as_string(rlang::ensym(color_by))
-  shape_by_string <- rlang::as_string(rlang::ensym(shape_by))
-
-  # if (is.character(pull(x, {{color_by}}))) {
-  #   scale_color_func <- scale_color_manual(values = color_scales[[color_by_string]])
-  # } else {
-  #   scale_color_func <- scale_color_viridis_c()
-  # }
-
-  p <- ggplot(x, aes(x={{xvar}}, y={{yvar}}, color={{color_by}}, shape={{shape_by}})) +
-    geom_point(alpha = 0.8, size=3) +
-    #scale_color_func +
-    #scale_color_manual(values = color_scales[[color_by_string]]) +
-    #scale_shape_manual(values = shape_scales[[shape_by_string]]) +
+    theme(axis.text.y = element_text(color="black", size=7)) +
+    theme(strip.text = element_text(size=8)) +
+    facet_wrap(~ contrast, nrow=1) +
+    labs(x="Normalized Enrichment Score (NES)", y="Pathway")
+  p_colon <- ggplot(res_colon, aes(x=NES, y=pathway, fill=neglog10padj)) +
+    geom_col() +
+    scale_fill_viridis_c() +
     theme_bw() +
-    theme(aspect.ratio=1,
-          legend.text = element_text(size=7))
-    # coord_fixed() +
-  return(p)
-}
-
-# bg_auc
-p12 <- pca_plot(x, PC1, PC2, bg_auc, study)
-p13 <- pca_plot(x, PC1, PC3, bg_auc, study)
-p23 <- pca_plot(x, PC2, PC3, bg_auc, study)
-p <- p12 + p13 + p23 + 
-  plot_layout(guides = "collect")
-f <- file.path(plot_dir, paste0("pca_bg_auc_pc123.pdf"))
-ggsave(f, plot=p, device="pdf", width=12, height=6)
-
-# region
-p12 <- pca_plot(x, PC1, PC2, region, study)
-p13 <- pca_plot(x, PC1, PC3, region, study)
-p23 <- pca_plot(x, PC2, PC3, region, study)
-p <- p12 + p13 + p23 + 
-  plot_layout(guides = "collect")
-f <- file.path(plot_dir, paste0("pca_region_pc123.pdf"))
-ggsave(f, plot=p, device="pdf", width=12, height=6)
-
-# patient
-p12 <- pca_plot(x, PC1, PC2, patient, study)
-p13 <- pca_plot(x, PC1, PC3, patient, study)
-p23 <- pca_plot(x, PC2, PC3, patient, study)
-p <- p12 + p13 + p23 + 
-  plot_layout(guides = "collect")
-f <- file.path(plot_dir, paste0("pca_patient_pc123.pdf"))
-ggsave(f, plot=p, device="pdf", width=12, height=6)
-
-# study
-p12 <- pca_plot(x, PC1, PC2, study)
-p13 <- pca_plot(x, PC1, PC3, study)
-p23 <- pca_plot(x, PC2, PC3, study)
-p <- p12 + p13 + p23 + 
-  plot_layout(guides = "collect")
-f <- file.path(plot_dir, paste0("pca_study_pc123.pdf"))
-ggsave(f, plot=p, device="pdf", width=12, height=6)
-```
+    theme(axis.text.y = element_text(color="black", size=7)) +
+    theme(strip.text = element_text(size=8)) +
+    facet_wrap(~ contrast, nrow=1) +
+    labs(x="Normalized Enrichment Score (NES)", y="Pathway")
+
+  return (p_panc + p_brain + p_ln + p_kidney + p_colon)
+}
+
+gsea_none_q3 <- prep_gsea(gsea_none_q3)
+res_none_q3 <- prep_all(gsea_none_q3)
+p <- plot_clusters(res_none_q3)
+f <- file.path(gsea_dir, "gsea_none_q3")
+ggsave(paste0(f, ".png"), p, width=20, height=15)
+ggsave(paste0(f, ".pdf"), p, width=20, height=15)
+
+gsea_none_tmm <- prep_gsea(gsea_none_tmm)
+res_none_tmm <- prep_all(gsea_none_tmm)
+p <- plot_clusters(res_none_tmm)
+f <- file.path(gsea_dir, "gsea_none_tmm")
+ggsave(paste0(f, ".png"), p, width=20, height=15)
+ggsave(paste0(f, ".pdf"), p, width=20, height=15)
+
+gsea_none_rle <- prep_gsea(gsea_none_rle)
+res_none_rle <- prep_all(gsea_none_rle)
+p <- plot_clusters(res_none_rle)
+f <- file.path(gsea_dir, "gsea_none_rle")
+ggsave(paste0(f, ".png"), p, width=20, height=15)
+ggsave(paste0(f, ".pdf"), p, width=20, height=15)
+
+gsea_none_qn <- prep_gsea(gsea_none_qn)
+res_none_qn <- prep_all(gsea_none_qn)
+p <- plot_clusters(res_none_qn)
+f <- file.path(gsea_dir, "gsea_none_qn")
+ggsave(paste0(f, ".png"), p, width=20, height=15)
+ggsave(paste0(f, ".pdf"), p, width=20, height=15)
+
+gsea_bgsub_q3 <- prep_gsea(gsea_bgsub_q3)
+res_bgsub_q3 <- prep_all(gsea_bgsub_q3)
+p <- plot_clusters(res_bgsub_q3)
+f <- file.path(gsea_dir, "gsea_bgsub_q3")
+ggsave(paste0(f, ".png"), p, width=20, height=15)
+ggsave(paste0(f, ".pdf"), p, width=20, height=15)
+
+gsea_bgsub_tmm <- prep_gsea(gsea_bgsub_tmm)
+res_bgsub_tmm <- prep_all(gsea_bgsub_tmm)
+p <- plot_clusters(res_bgsub_tmm)
+f <- file.path(gsea_dir, "gsea_bgsub_tmm")
+ggsave(paste0(f, ".png"), p, width=20, height=15)
+ggsave(paste0(f, ".pdf"), p, width=20, height=15)
+
+gsea_bgsub_rle <- prep_gsea(gsea_bgsub_rle)
+res_bgsub_rle <- prep_all(gsea_bgsub_rle)
+p <- plot_clusters(res_bgsub_rle)
+f <- file.path(gsea_dir, "gsea_bgsub_rle")
+ggsave(paste0(f, ".png"), p, width=20, height=15)
+ggsave(paste0(f, ".pdf"), p, width=20, height=15)
+
+gsea_bgsub_qn <- prep_gsea(gsea_bgsub_qn)
+res_bgsub_qn <- prep_all(gsea_bgsub_qn)
+p <- plot_clusters(res_bgsub_qn)
+f <- file.path(gsea_dir, "gsea_bgsub_qn")
+ggsave(paste0(f, ".png"), p, width=20, height=15)
+ggsave(paste0(f, ".pdf"), p, width=20, height=15)
+
+gsea_qq_q3 <- prep_gsea(gsea_qq_q3)
+res_qq_q3 <- prep_all(gsea_qq_q3)
+p <- plot_clusters(res_qq_q3)
+f <- file.path(gsea_dir, "gsea_qq_q3")
+ggsave(paste0(f, ".png"), p, width=20, height=15)
+ggsave(paste0(f, ".pdf"), p, width=20, height=15)
+
+gsea_qq_tmm <- prep_gsea(gsea_qq_tmm)
+res_qq_tmm <- prep_all(gsea_qq_tmm)
+p <- plot_clusters(res_qq_tmm)
+f <- file.path(gsea_dir, "gsea_qq_tmm")
+ggsave(paste0(f, ".png"), p, width=20, height=15)
+ggsave(paste0(f, ".pdf"), p, width=20, height=15)
+
+gsea_qq_rle <- prep_gsea(gsea_qq_rle)
+res_qq_rle <- prep_all(gsea_qq_rle)
+p <- plot_clusters(res_qq_rle)
+f <- file.path(gsea_dir, "gsea_qq_rle")
+ggsave(paste0(f, ".png"), p, width=20, height=15)
+ggsave(paste0(f, ".pdf"), p, width=20, height=15)
+
+gsea_qq_qn <- prep_gsea(gsea_qq_qn)
+res_qq_qn <- prep_all(gsea_qq_qn)
+p <- plot_clusters(res_qq_qn)
+f <- file.path(gsea_dir, "gsea_qq_qn")
+ggsave(paste0(f, ".png"), p, width=20, height=15)
+ggsave(paste0(f, ".pdf"), p, width=20, height=15)
+```

### 2024-09-24 R/input.R
move code to replace NA with 0 into the prepare_input function. happens after counts is split into metadata / counts.

@@ -42,6 +42,11 @@ prepare_input <- function(samples, counts,
   # ensure samples and count tables are aligned
   stopifnot(all(pull(samples, {{ aoi_key_col }}) == colnames(counts)))
 
+  # if count is NA or blank, set to 0
+  counts <- counts %>%
+    replace(is.na(.), 0) %>%
+    replace(. == "", 0)
+
   # ensure no duplicate probes
   if (sum(duplicated(meta$probe)) > 0) {
     stop("duplicate probe found")
@@ -117,11 +122,6 @@ read_tsv <- function(sample_tsv_file,
   valid_cols <- col_names != ""
   counts <- counts[, valid_cols]
 
-  # Madison edit: if count is blank, set to 0
-  counts <- counts %>%
-    mutate_all(~replace(., is.na(.), 0)) %>%
-    mutate_all(~ifelse(. == "", 0, .))
-
   # prepare dataset
   ds <- prepare_input(samples, counts, slide_key_col, roi_key_col,
                       segment_key_col, aoi_key_col, probe_key_col,
@@ -171,11 +171,6 @@ read_xlsx <- function(xlsx_file,
   # read counts
   counts <- readxl::read_excel(xlsx_file, sheet=count_sheet,
                                .name_repair = "minimal")
-
-  # Madison edit: if count is blank, set to 0
-  counts <- counts %>%
-    mutate_all(~replace(., is.na(.), 0)) %>%
-    mutate_all(~ifelse(. == "", 0, .))
 
   # prepare dataset
   ds <- prepare_input(samples, counts, slide_key_col, roi_key_col,
@@ -241,7 +236,7 @@ read_dcc <- function(dcc_tar_file, dcc_dir) {
     tidyr::hoist("Scan_Attributes", SampleID="SampleID") %>%
     dplyr::select(-"Scan_Attributes") %>%
     tidyr::unnest("Code_Summary") %>%
-    tidyr::pivot_wider(names_from="SampleID", values_from="Count", values_fill=1)
+    tidyr::pivot_wider(names_from="SampleID", values_from="Count", values_fill=0)
 
   # cleanup
   if (file.exists(dcc_dir)) {

### 2024-10-19 inst/manuscript_figures.Rmd
gsea updates

@@
 ```
+
+### GSEA: all msigdb
+```{r all_msigdb}
+gsea_nperms <- 10000
+
+read_msigdb_json_file <- function(filename) {
+  x <- jsonlite::fromJSON(filename)
+  y <- tibble(gs_name = names(x), gs = x) %>% 
+    unnest_wider(gs) %>%
+    unnest_longer(geneSymbols) %>%
+    select(gs_cat = collection,
+           gs_name = gs_name,
+           gene_symbol = geneSymbols)
+}
+
+# read msigdb data
+
+f <- system.file("extdata", "msigdb.v2024.1.Hs.json", package = "stgeomx")
+msigdb_gene_sets <- read_msigdb_json_file(f)
+
+# determine gene universe
+msigdb_genes <- intersect(gene_universe, msigdb_gene_sets$gene_symbol)
+
+# filter genes not in our dataset
+msigdb_gene_sets <- msigdb_gene_sets %>%
+  dplyr::filter(gene_symbol %in% gene_universe) %>%
+  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
+
+# gene set categories
+gs_hallmark <- dplyr::filter(msigdb_gene_sets, gs_cat == "H")
+gs_hallmark <- split(x = gs_hallmark$gene_symbol, f = gs_hallmark$gs_name)
+gs_gobp <- dplyr::filter(msigdb_gene_sets, gs_cat == "C5:GO:BP")
+gs_gobp <- split(x = gs_gobp$gene_symbol, f = gs_gobp$gs_name)
+gs_3ca <- dplyr::filter(msigdb_gene_sets, gs_cat == "C4:3CA")
+gs_3ca <- split(x = gs_3ca$gene_symbol, f = gs_3ca$gs_name)
+
+cp_cat <- c("C2:CP:BIOCARTA", "C2:CP:KEGG_MEDICUS", "C2:CP:PID", "C2:CP:REACTOME",
+            "C2:CP:WIKIPATHWAYS")
+gs_cp <- dplyr::filter(msigdb_gene_sets, gs_cat %in% cp_cat)
+gs_cp <- split(x = gs_cp$gene_symbol, f = gs_cp$gs_name)
+gs_tft <- dplyr::filter(msigdb_gene_sets, gs_cat == "C3:TFT:GTRD")
+gs_tft <- split(x = gs_tft$gene_symbol, f = gs_tft$gs_name)
+
+# new gsea code
+get_fgsea_ranks_de <- function(xde, mycontrast) {
+  ranks <- xde %>%
+    dplyr::filter(contrast == mycontrast) %>%
+    select(gene, log2fc, padj) %>%
+    mutate(rank = log2fc)
+  ranks = sort(setNames(ranks$rank, ranks$gene), decreasing = TRUE)
+  return(ranks)
+}
+
+fgsea_run_batch <- function(xde, xgs,
+                            fgsea_eps=0,
+                            fgsea_nperms=1000) {
+  gsea <- NULL
+  for (xcontrast in unique(xde$contrast)) {
+    ranks <- get_fgsea_ranks_de(xde, xcontrast)
+    res <- fgsea(pathways=xgs,
+                 stats=ranks,
+                 eps=fgsea_eps,
+                 nPermSimple=fgsea_nperms)
+    res <- mutate(res, contrast = xcontrast)
+    gsea <- bind_rows(gsea, res)     
+  }
+  return(gsea)
+}
+
+stgeomx_gsea_pipeline <- function(gsea, de, method) {
+  res <- fgsea_run_batch(de, gs_hallmark,
+                       fgsea_eps=0, fgsea_nperms=gsea_nperms)
+  res <- res %>% mutate(method = method)
+  gsea <- bind_rows(gsea, res)
+
+  res <- fgsea_run_batch(de, gs_gobp,
+                       fgsea_eps=0, fgsea_nperms=gsea_nperms)
+  res <- res %>% mutate(method = method)
+  gsea <- bind_rows(gsea, res)
+
+  res <- fgsea_run_batch(de, gs_3ca,
+                       fgsea_eps=0, fgsea_nperms=gsea_nperms)
+  res <- res %>% mutate(method = method)
+  gsea <- bind_rows(gsea, res)
+
+  res <- fgsea_run_batch(de, gs_cp,
+                       fgsea_eps=0, fgsea_nperms=gsea_nperms)
+  res <- res %>% mutate(method = method)
+  gsea <- bind_rows(gsea, res)
+
+  res <- fgsea_run_batch(de, gs_tft,
+                       fgsea_eps=0, fgsea_nperms=gsea_nperms)
+  res <- res %>% mutate(method = method)
+  gsea <- bind_rows(gsea, res)
+
+  return(gsea)
+}
+
+# Run full analysis
+gsea_all <- NULL
+gsea_none_q3 <- NULL
+gsea_none_tmm <- NULL
+gsea_none_rle <- NULL
+gsea_none_qn <- NULL
+
+gsea_bgsub_q3 <- NULL
+gsea_bgsub_tmm <- NULL
+gsea_bgsub_rle <- NULL
+gsea_bgsub_qn <- NULL
+
+gsea_qq_q3 <- NULL
+gsea_qq_tmm <- NULL
+gsea_qq_rle <- NULL
+gsea_qq_qn <- NULL
+
+# run stgeomx pipeline
+gsea_none_q3 <- stgeomx_gsea_pipeline(gsea_none_q3, de_none_q3, "none_q3")
+gsea_none_tmm <- stgeomx_gsea_pipeline(gsea_none_tmm, de_none_tmm, "none_tmm")
+gsea_none_rle <- stgeomx_gsea_pipeline(gsea_none_rle, de_none_rle, "none_rle")
+gsea_none_qn <- stgeomx_gsea_pipeline(gsea_none_qn, de_none_qn, "none_qn")
+
+gsea_bgsub_q3 <- stgeomx_gsea_pipeline(gsea_bgsub_q3, de_bgsub_q3, "bgsub_q3")
+gsea_bgsub_tmm <- stgeomx_gsea_pipeline(gsea_bgsub_tmm, de_bgsub_tmm, "bgsub_tmm")
+gsea_bgsub_rle <- stgeomx_gsea_pipeline(gsea_bgsub_rle, de_bgsub_rle, "bgsub_rle")
+gsea_bgsub_qn <- stgeomx_gsea_pipeline(gsea_bgsub_qn, de_bgsub_qn, "bgsub_qn")
+
+gsea_qq_q3 <- stgeomx_gsea_pipeline(gsea_qq_q3, de_qq_q3, "qq_q3")
+gsea_qq_tmm <- stgeomx_gsea_pipeline(gsea_qq_tmm, de_qq_tmm, "qq_tmm")
+gsea_qq_rle <- stgeomx_gsea_pipeline(gsea_qq_rle, de_qq_rle, "qq_rle")
+gsea_qq_qn <- stgeomx_gsea_pipeline(gsea_qq_qn, de_qq_qn, "qq_qn")
+
+# write results for clusters
+contrasts <- c("clusterPanc", "clusterBrain", "clusterLn", "clusterKidney", "clusterColon")
+manuscript_pathways <- c(names(gs_hallmark), names(gs_gobp), names(gs_3ca), names(gs_cp), names(gs_tft))
+res <- gsea_all %>% filter(contrast %in% contrasts,
+                           pathway %in% manuscript_pathways)
+
+prefix <- "gsea_all"
+f <- file.path(results_dir, paste0(prefix, "_results.tsv"))
+data.table::fwrite(res, file=f, sep="\t", sep2=c("", " ", ""))
+f <- file.path(results_dir, paste0(prefix, "_results.xlsx"))
+write_xlsx(res, f)
+```
+
 ### GSEA plots: helper functions

### 2024-12-12 inst/manuscript_figures.Rmd
most recent updates

@@
 library(pheatmap)
+library(dplyr)
+library(tidyr)
+library(jsonlite)
 
@@
 working_dir <- "/Users/madisoncaldwell/Desktop/stgeomx"
+
+# load package
+install.packages(working_dir, repos = NULL, type = "source")
 
@@
   x <- NULL
+  res <- NULL
+
   for (coef in colnames(contrasts)) {
@@
     res <- res %>%
-      select(gene,
+      dplyr::select(gene,
              log2fc=logFC,
@@
  ```{r de heatmaps, include=FALSE}
-
-# heatmap functions
-plot_pheatmap <- function(m, 
-                          annot_col,
-                          # annot_row, 
-                          annot_colors,
-                          filename, width, height,
-                          xmin=-6, xmax=6) {
-  pheatmap(m,
-           scale = "none",
-           color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
-           breaks = seq(xmin, xmax, length.out=51),
-           annotation_col = annot_col,
-           # annotation_row = annot_row,
-           # annotation_colors = annot_colors,
-           border_color = NA,
-           cluster_rows = TRUE,
-           cluster_cols = TRUE,
-           clustering_distance_rows = "euclidean",
-           clustering_distance_cols = "euclidean",
-           clustering_method = "ward.D2",
-#           cellwidth = 6,
-#           cellheight = 6,
-           fontsize = 6,
-           fontsize_row = 6,
-           fontsize_col = 6,
-           filename = filename, 
-           width = width,
-           height = height)
-}
-
-# setup heatmap annotations
-# column annotations: show clusters, study
-# no row annotations
-gene_annot <- fds_qq_qn$meta
-# %>%
-#   left_join(hpa_pdac, by=c("gene"="gene_symbol")) %>%
-#   left_join(cptac_de_merged, by=c("gene"="gene_symbol")) %>%
-#   mutate(de_cptac_rna = replace_na(de_cptac_rna, "na"),
-#          de_cptac_prot = replace_na(de_cptac_prot, "na"))
-# annot_row <- gene_annot %>%
-#   select(gene) %>%
-#   column_to_rownames("gene") %>%
+# 
+# # heatmap functions
+# plot_pheatmap <- function(m, 
+#                           annot_col,
+#                           # annot_row, 
+#                           annot_colors,
+#                           filename, width, height,
+#                           xmin=-6, xmax=6) {
+#   pheatmap(m,
+#            scale = "none",
+#            color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
+#            breaks = seq(xmin, xmax, length.out=51),
+#            annotation_col = annot_col,
+#            # annotation_row = annot_row,
+#            # annotation_colors = annot_colors,
+#            border_color = NA,
+#            cluster_rows = TRUE,
+#            cluster_cols = TRUE,
+#            clustering_distance_rows = "euclidean",
+#            clustering_distance_cols = "euclidean",
+#            clustering_method = "ward.D2",
+# #           cellwidth = 6,
+# #           cellheight = 6,
+#            fontsize = 6,
+#            fontsize_row = 6,
+#            fontsize_col = 6,
+#            filename = filename, 
+#            width = width,
+#            height = height)
+# }
+# 
+# # setup heatmap annotations
+# # column annotations: show clusters, study
+# # no row annotations
+# gene_annot <- fds_qq_qn$meta
+# # %>%
+# #   left_join(hpa_pdac, by=c("gene"="gene_symbol")) %>%
+# #   left_join(cptac_de_merged, by=c("gene"="gene_symbol")) %>%
+# #   mutate(de_cptac_rna = replace_na(de_cptac_rna, "na"),
+# #          de_cptac_prot = replace_na(de_cptac_prot, "na"))
+# # annot_row <- gene_annot %>%
+# #   select(gene) %>%
+# #   column_to_rownames("gene") %>%
+# #   as.data.frame()
+# annot_col <- column_to_rownames(fds_qq_qn$samples, "aoi") %>%
+#   select(study, cluster_name) %>%
 #   as.data.frame()
-annot_col <- column_to_rownames(fds_qq_qn$samples, "aoi") %>%
-  select(study, cluster_name) %>%
-  as.data.frame()
-
-#
-# IPMN clusters
-#
-# de gene params
-padj_cutoff <- 0.01
-log2fc_cutoff <- 3
-ntop <- 10
-contrasts <- c("clusterPanc", "clusterLn", "clusterKidney", "clusterColon", "clusterBrain")
-
-g <- filter(de_qq_qn, contrast %in% contrasts,
-            abs(log2fc) > log2fc_cutoff, 
-            padj < padj_cutoff) %>%
-  group_by(contrast, de) %>%
-  slice_max(abs(log2fc), n=ntop) %>%
-  ungroup() %>%
-  distinct(gene) %>% 
-  pull(gene)
-ngenes <- length(g)
-cluster_marker_genes <- g
-
-# centered gene expression (subtract median)
-s <- fds_qq_qn$samples
-x <- log2_norm_cpm_qq_qn[g, s$aoi]
-u <- apply(x, 1, mean)
-x <- sweep(x, 1, u, FUN="-")
-xmin <- -6
-xmax <- 6
-
-# heatmap
-heatmap_dir <- file.path(plot_dir, "heatmap")
-
-f <- file.path(plot_dir,  paste0("heatmap_de_cluster_markers_", ngenes, "_genes"))
-h <- 6
-w <- 6
-p <- plot_pheatmap(x, 
-                   annot_col[s$aoi,], 
-                   annot_row[g,], 
-                   annot_colors=scale_color_continuous(),
-                   filename=paste0(f, ".png"), 
-                   width=w, height=h,
-                   xmin=xmin, xmax=xmax)
-save_pheatmap_pdf(p, filename=paste0(f, ".pdf"), width=w, height=h)
+# 
+# #
+# # IPMN clusters
+# #
+# # de gene params
+# padj_cutoff <- 0.01
+# log2fc_cutoff <- 3
+# ntop <- 10
+# contrasts <- c("clusterPanc", "clusterLn", "clusterKidney", "clusterColon", "clusterBrain")
+# 
+# g <- filter(de_qq_qn, contrast %in% contrasts,
+#             abs(log2fc) > log2fc_cutoff, 
+#             padj < padj_cutoff) %>%
+#   group_by(contrast, de) %>%
+#   slice_max(abs(log2fc), n=ntop) %>%
+#   ungroup() %>%
+#   distinct(gene) %>% 
+#   pull(gene)
+# ngenes <- length(g)
+# cluster_marker_genes <- g
+# 
+# # centered gene expression (subtract median)
+# s <- fds_qq_qn$samples
+# x <- log2_norm_cpm_qq_qn[g, s$aoi]
+# u <- apply(x, 1, mean)
+# x <- sweep(x, 1, u, FUN="-")
+# xmin <- -6
+# xmax <- 6
+# 
+# # heatmap
+# heatmap_dir <- file.path(plot_dir, "heatmap")
+# 
+# f <- file.path(plot_dir,  paste0("heatmap_de_cluster_markers_", ngenes, "_genes"))
+# h <- 6
+# w <- 6
+# p <- plot_pheatmap(x, 
+#                    annot_col[s$aoi,], 
+#                    annot_row[g,], 
+#                    annot_colors=scale_color_continuous(),
+#                    filename=paste0(f, ".png"), 
+#                    width=w, height=h,
+#                    xmin=xmin, xmax=xmax)
+# save_pheatmap_pdf(p, filename=paste0(f, ".pdf"), width=w, height=h)
 
@@
 for (gene in feature_plot_genes) {
-  print(gene)
   x <- bind_cols(s, as_tibble(pca_res_qq_qn$x), gene=unlist(y[gene,]))
@@
 
-plot_de_scatter2d <- function(tbl, xcontrast, ycontrast, nlabel=10, 
-                              highlight_genes=NULL) {
-  if (is.null(highlight_genes)) {
-    highlight_genes <- c()
-  }
-  xde <- sym(paste0("de_", xcontrast))
-  yde <- sym(paste0("de_", ycontrast))
-  xlog2fc <- sym(paste0("log2fc_", xcontrast))
-  ylog2fc <- sym(paste0("log2fc_", ycontrast))
-
-  x <- tbl %>% 
-    mutate(subtype = paste0("x", {{xde}}, "_", "y", {{yde}}),
-           log2fc_abs_x = case_when({{xde}} == "no" ~ 0,
-                                    TRUE ~ abs({{xlog2fc}})),
-           log2fc_abs_y = case_when({{yde}} == "no" ~ 0,
-                                    TRUE ~ abs({{ylog2fc}})),
-           log2fc_abs_max = pmax(log2fc_abs_x, log2fc_abs_y))
-
-  color_scale = c(xno_yno = "#888888",
-                  xup_yup = "#FF0000",
-                  xdn_ydn = "#0000aa",
-                  xup_ydn = "#eeee00",
-                  xdn_yup = "#ff00ff",
-                  xup_yno = "#32cd32",
-                  xdn_yno = "#670092",
-                  xno_yup = "#ff6700",
-                  xno_ydn = "#16ccc4")
-  size_scale = c(xno_yno = 1, 
-                 xup_yup = 2,
-                 xdn_ydn = 2,
-                 xup_ydn = 2,
-                 xdn_yup = 2,
-                 xup_yno = 2,
-                 xdn_yno = 2,
-                 xno_yup = 2,
-                 xno_ydn = 2)
-
-  annot_data <- filter(x, subtype != "xno_yno")
-  none_data <- filter(x, subtype == "xno_yno")
-  label_data <- annot_data %>% 
-    group_by(subtype) %>%
-    slice_max(log2fc_abs_max, n=nlabel) %>%
-    ungroup()
-  highlight_data <- filter(x, gene %in% highlight_genes)
-
-  p <- ggplot(x, aes(x={{xlog2fc}}, 
-                     y={{ylog2fc}},
-                     color=subtype, 
-                     size=subtype)) +
-    geom_point(data=none_data, alpha=0.4) + 
-    geom_point(data=annot_data, alpha=0.8) +
-    geom_text_repel(data=label_data, color="black", size=3, aes(label=gene), max.overlaps=Inf) + 
-    geom_point(data=highlight_data, pch=21, fill=NA, size=5, colour="#ffff00", stroke=1) + 
-    geom_point(data=highlight_data, pch=21, fill=NA, size=4, colour="#222222", stroke=1, alpha=0.5) +
-    geom_text_repel(data=highlight_data, color="black", size=3, aes(label=gene), max.overlaps=Inf) +     
-    scale_color_manual(values = color_scale) +
-    scale_size_manual(values = size_scale, guide = "none") + 
-    theme_minimal() +
-    xlab(rlang::as_string(xlog2fc)) +
-    ylab(rlang::as_string(ylog2fc)) +
-    labs(color = "DE Genes")
-  return(p)
-}
-
-# merge de analysis by gene
-de_merged <- de_qq_qn %>% select(gene, contrast, log2fc, padj, de) %>%
-  pivot_wider(names_from = contrast, values_from = c(log2fc, padj, de))
-nlabel <- 10
-width <- 8
-height <- 8
-
-#
-# IPMN Clusters
-#
-p <- plot_de_scatter2d(de_merged, 
-                       xcontrast="clusterColon", 
-                       ycontrast="clusterLn", 
-                       nlabel=nlabel) +
-  labs(title="IPMN Specimens", 
-       subtitle="clusterColon (x-axis) clusterLn (y-axis)")
-f <- file.path(plot_dir, "scatter2d_de_x_clusterColon_y_clusterLn")
-ggsave(paste0(f, ".pdf"), plot=p, width=8, height=8)
-ggsave(paste0(f, ".png"), plot=p, width=8, height=8)
+# plot_de_scatter2d <- function(tbl, xcontrast, ycontrast, nlabel=10, 
+#                               highlight_genes=NULL) {
+#   if (is.null(highlight_genes)) {
+#     highlight_genes <- c()
+#   }
+#   xde <- sym(paste0("de_", xcontrast))
+#   yde <- sym(paste0("de_", ycontrast))
+#   xlog2fc <- sym(paste0("log2fc_", xcontrast))
+#   ylog2fc <- sym(paste0("log2fc_", ycontrast))
+# 
+#   x <- tbl %>% 
+#     mutate(subtype = paste0("x", {{xde}}, "_", "y", {{yde}}),
+#            log2fc_abs_x = case_when({{xde}} == "no" ~ 0,
+#                                     TRUE ~ abs({{xlog2fc}})),
+#            log2fc_abs_y = case_when({{yde}} == "no" ~ 0,
+#                                     TRUE ~ abs({{ylog2fc}})),
+#            log2fc_abs_max = pmax(log2fc_abs_x, log2fc_abs_y))
+#   
+#   color_scale = c(xno_yno = "#888888",
+#                   xup_yup = "#FF0000",
+#                   xdn_ydn = "#0000aa",
+#                   xup_ydn = "#eeee00",
+#                   xdn_yup = "#ff00ff",
+#                   xup_yno = "#32cd32",
+#                   xdn_yno = "#670092",
+#                   xno_yup = "#ff6700",
+#                   xno_ydn = "#16ccc4")
+#   size_scale = c(xno_yno = 1, 
+#                  xup_yup = 2,
+#                  xdn_ydn = 2,
+#                  xup_ydn = 2,
+#                  xdn_yup = 2,
+#                  xup_yno = 2,
+#                  xdn_yno = 2,
+#                  xno_yup = 2,
+#                  xno_ydn = 2)
+#   
+#   annot_data <- filter(x, subtype != "xno_yno")
+#   none_data <- filter(x, subtype == "xno_yno")
+#   label_data <- annot_data %>% 
+#     group_by(subtype) %>%
+#     slice_max(log2fc_abs_max, n=nlabel) %>%
+#     ungroup()
+#   highlight_data <- filter(x, gene %in% highlight_genes)
+# 
+#   p <- ggplot(x, aes(x={{xlog2fc}}, 
+#                      y={{ylog2fc}},
+#                      color=subtype, 
+#                      size=subtype)) +
+#     geom_point(data=none_data, alpha=0.4) + 
+#     geom_point(data=annot_data, alpha=0.8) +
+#     geom_text_repel(data=label_data, color="black", size=3, aes(label=gene), max.overlaps=Inf) + 
+#     geom_point(data=highlight_data, pch=21, fill=NA, size=5, colour="#ffff00", stroke=1) + 
+#     geom_point(data=highlight_data, pch=21, fill=NA, size=4, colour="#222222", stroke=1, alpha=0.5) +
+#     geom_text_repel(data=highlight_data, color="black", size=3, aes(label=gene), max.overlaps=Inf) +     
+#     scale_color_manual(values = color_scale) +
+#     scale_size_manual(values = size_scale, guide = "none") + 
+#     theme_minimal() +
+#     xlab(rlang::as_string(xlog2fc)) +
+#     ylab(rlang::as_string(ylog2fc)) +
+#     labs(color = "DE Genes")
+#   return(p)
+# }
+# 
+# 
+# # merge de analysis by gene
+# de_merged <- de_qq_qn %>% select(gene, contrast, log2fc, padj, de) %>%
+#   pivot_wider(names_from = contrast, values_from = c(log2fc, padj, de))
+# nlabel <- 10
+# width <- 8
+# height <- 8
+# 
+# #
+# # IPMN Clusters
+# #
+# p <- plot_de_scatter2d(de_merged, 
+#                        xcontrast="clusterColon", 
+#                        ycontrast="clusterLn", 
+#                        nlabel=nlabel) +
+#   labs(title="IPMN Specimens", 
+#        subtitle="clusterColon (x-axis) clusterLn (y-axis)")
+# f <- file.path(plot_dir, "scatter2d_de_x_clusterColon_y_clusterLn")
+# ggsave(paste0(f, ".pdf"), plot=p, width=8, height=8)
+# ggsave(paste0(f, ".png"), plot=p, width=8, height=8)
 
@@
 ``` {r gsea}
+# BiocManager::install("fgsea")
+# 
+# gsea_nperms <- 10000
+# 
+# read_msigdb_json_file <- function(filename) {
+#   x <- jsonlite::fromJSON(filename)
+#   y <- tibble(gs_name = names(x), gs = x) %>% 
+#     unnest_wider(gs) %>%
+#     unnest_longer(geneSymbols) %>%
+#     select(gs_cat = collection,
+#            gs_name = gs_name,
+#            gene_symbol = geneSymbols)
+# }
+# 
+# # read msigdb data: hallmark
+# f <- system.file("extdata", "h.all.v2024.1.Hs.json", package = "stgeomx")
+# msigdb_gene_sets <- read_msigdb_json_file(f)
+# # determine gene universe
+# msigdb_genes <- intersect(fds_qq_qn$meta$gene, msigdb_gene_sets$gene_symbol)
+# # filter genes not in our dataset
+# msigdb_gene_sets <- msigdb_gene_sets %>%
+#   dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
+#   distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
+# # gene set categories
+# gs_hallmark <- dplyr::filter(msigdb_gene_sets, gs_cat == "H")
+# gs_hallmark <- split(x = gs_hallmark$gene_symbol, f = gs_hallmark$gs_name)
+# 
+# # read msigdb data: pancreas_endocrine
+# f <- system.file("extdata", "GOBP_ENDOCRINE_PANCREAS_DEVELOPMENT.v2024.1.Hs.json", package = "stgeomx")
+# gobp_end_pancreas_gene_sets <- read_msigdb_json_file(f)
+# # determine gene universe
+# gobp_end_pancreas_genes <- intersect(fds_qq_qn$meta$gene, gobp_end_pancreas_gene_sets$gene_symbol)
+# # filter genes not in our dataset
+# gobp_end_pancreas_gene_sets <- gobp_end_pancreas_gene_sets %>%
+#   dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
+#   distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
+# # gene set categories
+# gs_end_panc <- dplyr::filter(gobp_end_pancreas_gene_sets, gs_cat == "C5:GO:BP")
+# gs_end_panc <- split(x = gs_end_panc$gene_symbol, f = gs_end_panc$gs_name)
+# 
+# # read msigdb data: pancreas secretion
+# f <- system.file("extdata", "GOBP_INSULIN_SECRETION.v2024.1.Hs.json", package = "stgeomx")
+# gobp_pancreas_secretion_gene_sets <- read_msigdb_json_file(f)
+# # determine gene universe
+# gobp_pancreas_secretion_genes <- intersect(fds_qq_qn$meta$gene, gobp_pancreas_secretion_gene_sets$gene_symbol)
+# # filter genes not in our dataset
+# gobp_pancreas_secretion_gene_sets <- gobp_pancreas_secretion_gene_sets %>%
+#   dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
+#   distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
+# # gene set categories
+# gs_panc_sec <- dplyr::filter(gobp_pancreas_secretion_gene_sets, gs_cat == "C5:GO:BP")
+# gs_panc_sec <- split(x = gs_panc_sec$gene_symbol, f = gs_panc_sec$gs_name)
+# 
+# # read msigdb data: brain development gene sets
+# f <- system.file("extdata", "brain_genesets.v2024.1.Hs.json", package = "stgeomx")
+# gobp_brain_gene_sets <- read_msigdb_json_file(f)
+# # determine gene universe
+# gobp_brain_genes <- intersect(fds_qq_qn$meta$gene, gobp_brain_gene_sets$gene_symbol)
+# # filter genes not in our dataset
+# gobp_brain_gene_sets <- gobp_brain_gene_sets %>%
+#   dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
+#   distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
+# # gene set categories
+# gs_brain <- dplyr::filter(gobp_brain_gene_sets, gs_cat == "C5:GO:BP")
+# gs_brain <- split(x = gs_brain$gene_symbol, f = gs_brain$gs_name)
+# 
+# # read msigdb data: ln development gene sets
+# f <- system.file("extdata", "ln_genesets.v2024.1.Hs.json", package = "stgeomx")
+# gobp_ln_gene_sets <- read_msigdb_json_file(f)
+# # determine gene universe
+# gobp_ln_genes <- intersect(fds_qq_qn$meta$gene, gobp_ln_gene_sets$gene_symbol)
+# # filter genes not in our dataset
+# gobp_ln_gene_sets <- gobp_ln_gene_sets %>%
+#   dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
+#   distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
+# # gene set categories
+# gs_ln <- dplyr::filter(gobp_ln_gene_sets, gs_cat == "C5:GO:BP")
+# gs_ln <- split(x = gs_ln$gene_symbol, f = gs_ln$gs_name)
+# 
+# # read msigdb data: kidney development gene sets
+# f <- system.file("extdata", "kidney_genesets.v2024.1.Hs.json", package = "stgeomx")
+# gobp_kidney_gene_sets <- read_msigdb_json_file(f)
+# # determine gene universe
+# gobp_kidney_genes <- intersect(fds_qq_qn$meta$gene, gobp_kidney_gene_sets$gene_symbol)
+# # filter genes not in our dataset
+# gobp_kidney_gene_sets <- gobp_kidney_gene_sets %>%
+#   dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
+#   distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
+# # gene set categories
+# gs_kidney <- dplyr::filter(gobp_kidney_gene_sets, gs_cat == "C5:GO:BP")
+# gs_kidney <- split(x = gs_kidney$gene_symbol, f = gs_kidney$gs_name)
+# 
+# # read msigdb data: colon development gene sets
+# f <- system.file("extdata", "colon_genesets.v2024.1.Hs.json", package = "stgeomx")
+# gobp_colon_gene_sets <- read_msigdb_json_file(f)
+# # determine gene universe
+# gobp_colon_genes <- intersect(fds_qq_qn$meta$gene, gobp_colon_gene_sets$gene_symbol)
+# # filter genes not in our dataset
+# gobp_colon_gene_sets <- gobp_colon_gene_sets %>%
+#   dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
+#   distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
+# # gene set categories
+# gs_colon <- dplyr::filter(gobp_colon_gene_sets, gs_cat == "C5:GO:BP")
+# gs_colon <- split(x = gs_colon$gene_symbol, f = gs_colon$gs_name)
+# 
+# # new gsea code
+# get_fgsea_ranks_de <- function(xde, mycontrast) {
+#   ranks <- xde %>%
+#     dplyr::filter(contrast == mycontrast) %>%
+#     select(gene, log2fc, padj) %>%
+#     mutate(rank = log2fc)
+#   ranks = sort(setNames(ranks$rank, ranks$gene), decreasing = TRUE)
+#   return(ranks)
+# }
+#  
+#  
+# fgsea_run_batch <- function(xde, xgs,
+#                             fgsea_eps=0,
+#                             fgsea_nperms=1000) {
+#   gsea <- NULL
+#   for (xcontrast in unique(xde$contrast)) {
+#     ranks <- get_fgsea_ranks_de(xde, xcontrast)
+#     res <- fgsea(pathways=xgs,
+#                  stats=ranks,
+#                  eps=fgsea_eps,
+#                  nPermSimple=fgsea_nperms)
+#     res <- mutate(res, contrast = xcontrast)
+#     gsea <- bind_rows(gsea, res)     
+#   }
+#   return(gsea)
+# }
+# 
+# stgeomx_gsea_pipeline <- function(gsea, de, method) {
+#   res <- fgsea_run_batch(de, gs_hallmark,
+#                        fgsea_eps=0, fgsea_nperms=gsea_nperms)
+#   res <- res %>% mutate(method = method)
+#   gsea <- bind_rows(gsea, res)
+#   
+#   res <- fgsea_run_batch(de, gs_ex_panc,
+#                          fgsea_eps=0, fgsea_nperms=gsea_nperms)
+#   res <- res %>% mutate(method = method)
+#   gsea <- bind_rows(gsea, res)
+#   
+#   
+#   res <- fgsea_run_batch(de, gs_end_panc,
+#                          fgsea_eps=0, fgsea_nperms=gsea_nperms)
+#   res <- res %>% mutate(method = method)
+#   gsea <- bind_rows(gsea, res)
+#   
+#   res <- fgsea_run_batch(de, gs_panc_sec,
+#                          fgsea_eps=0, fgsea_nperms=gsea_nperms)
+#   res <- res %>% mutate(method = method)
+#   gsea <- bind_rows(gsea, res)
+#   
+#   res <- fgsea_run_batch(de, gs_brain,
+#                          fgsea_eps=0, fgsea_nperms=gsea_nperms)
+#   res <- res %>% mutate(method = method)
+#   gsea <- bind_rows(gsea, res)
+#   
+#   res <- fgsea_run_batch(de, gs_ln,
+#                          fgsea_eps=0, fgsea_nperms=gsea_nperms)
+#   res <- res %>% mutate(method = method)
+#   gsea <- bind_rows(gsea, res)
+#   
+#   res <- fgsea_run_batch(de, gs_kidney,
+#                          fgsea_eps=0, fgsea_nperms=gsea_nperms)
+#   res <- res %>% mutate(method = method)
+#   gsea <- bind_rows(gsea, res)
+#   
+#   res <- fgsea_run_batch(de, gs_colon,
+#                          fgsea_eps=0, fgsea_nperms=gsea_nperms)
+#   res <- res %>% mutate(method = method)
+#   gsea <- bind_rows(gsea, res)
+#   
+#   # filter out irrelevant genes in hallmark
+#   gsea <- gsea %>%
+#   filter(!(pathway %in% c("HALLMARK_ESTROGEN_RESPONSE_LATE", "HALLMARK_ESTROGEN_RESPONSE_EARLY", "HALLMARK_GLYCOLYSIS", "HALLMARK_P53_PATHWAY", "HALLMARK_HEME_METABOLISM", "HALLMARK_ADIPOGENESIS", "HALLMARK_MYC_TARGETS_V2", "HALLMARK_MYC_TARGETS_V1", "HALLMARK_OXIDATIVE_PHOSPHORYLATION", "HALLMARK_HYPOXIA", "HALLMARK_UV_RESPONSE_DN", "HALLMARK_MYOGENESIS", "HALLMARK_APICAL_SURFACE", "HALLMARK_APOPTOSIS", "HALLMARK_IL6_JAK_STAT3_SIGNALING", "HALLMARK_IL2_STAT5_SIGNALING", "HALLMARK_MITOTIC_SPINDLE", "HALLMARK_APICAL_JUNCTION", "HALLMARK_KRAS_SIGNALING_UP", "HALLMARK_KRAS_SIGNALING_DN", "HALLMARK_UNFOLDED_PROTEIN_RESPONSE", "HALLMARK_DNA_REPAIR", "HALLMARK_FATTY_ACID_METABOLISM", "HALLMARK_HEDGEHOG_SIGNALING", "HALLMARK_ANGIOGENESIS", "HALLMARK_E2F_TARGETS", "HALLMARK_G2M_CHECKPOINT", "HALLMARK_TNFA_SIGNALING_VIA_NFKB", "HALLMARK_PROTEIN_SECRETION", "HALLMARK_CHOLESTEROL_HOMEOSTASIS")))
+#   
+#   return(gsea)
+# }
+# 
+# # Run full analysis
+# gsea_all <- NULL
+# gsea_none_q3 <- NULL
+# gsea_none_tmm <- NULL
+# gsea_none_rle <- NULL
+# gsea_none_qn <- NULL
+# 
+# gsea_bgsub_q3 <- NULL
+# gsea_bgsub_tmm <- NULL
+# gsea_bgsub_rle <- NULL
+# gsea_bgsub_qn <- NULL
+# 
+# gsea_qq_q3 <- NULL
+# gsea_qq_tmm <- NULL
+# gsea_qq_rle <- NULL
+# gsea_qq_qn <- NULL
+# 
+# # run stgeomx pipeline
+# gsea_none_q3 <- stgeomx_gsea_pipeline(gsea_none_q3, de_none_q3, "none_q3")
+# gsea_none_tmm <- stgeomx_gsea_pipeline(gsea_none_tmm, de_none_tmm, "none_tmm")
+# gsea_none_rle <- stgeomx_gsea_pipeline(gsea_none_rle, de_none_rle, "none_rle")
+# gsea_none_qn <- stgeomx_gsea_pipeline(gsea_none_qn, de_none_qn, "none_qn")
+# 
+# gsea_bgsub_q3 <- stgeomx_gsea_pipeline(gsea_bgsub_q3, de_bgsub_q3, "bgsub_q3")
+# gsea_bgsub_tmm <- stgeomx_gsea_pipeline(gsea_bgsub_tmm, de_bgsub_tmm, "bgsub_tmm")
+# gsea_bgsub_rle <- stgeomx_gsea_pipeline(gsea_bgsub_rle, de_bgsub_rle, "bgsub_rle")
+# gsea_bgsub_qn <- stgeomx_gsea_pipeline(gsea_bgsub_qn, de_bgsub_qn, "bgsub_qn")
+# 
+# gsea_qq_q3 <- stgeomx_gsea_pipeline(gsea_qq_q3, de_qq_q3, "qq_q3")
+# gsea_qq_tmm <- stgeomx_gsea_pipeline(gsea_qq_tmm, de_qq_tmm, "qq_tmm")
+# gsea_qq_rle <- stgeomx_gsea_pipeline(gsea_qq_rle, de_qq_rle, "qq_rle")
+# gsea_qq_qn <- stgeomx_gsea_pipeline(gsea_qq_qn, de_qq_qn, "qq_qn")
+
+# # write results for clusters
+# contrasts <- c("clusterPanc", "clusterBrain", "clusterLn", "clusterKidney", "clusterColon")
+# manuscript_pathways <- c(names(gs_hallmark))
+# res <- gsea_all %>% filter(contrast %in% contrasts, 
+#                            pathway %in% manuscript_pathways)
+
+# prefix <- "gsea_all"
+# f <- file.path(results_dir, paste0(prefix, "_results.tsv"))
+# data.table::fwrite(res, file=f, sep="\t", sep2=c("", " ", ""))
+# f <- file.path(results_dir, paste0(prefix, "_results.xlsx"))
+# write_xlsx(res, f)
+```
+
+### GSEA: all msigdb
+```{r all_msigdb}
+BiocManager::install("fgsea")
+library(fgsea)
+library(dplyr)
+
 gsea_nperms <- 10000
@@
     unnest_longer(geneSymbols) %>%
-    select(gs_cat = collection,
+    dplyr::select(gs_cat = collection,
            gs_name = gs_name,
@@
 
-# read msigdb data: hallmark
-f <- system.file("extdata", "h.all.v2024.1.Hs.json", package = "stgeomx")
+# read msigdb data
+
+f <- system.file("extdata", "msigdb.v2024.1.Hs.json", package = "stgeomx")
 msigdb_gene_sets <- read_msigdb_json_file(f)
+
 # determine gene universe
 msigdb_genes <- intersect(fds_qq_qn$meta$gene, msigdb_gene_sets$gene_symbol)
+
 # filter genes not in our dataset
@@
   dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
-  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
-# gene set categories
-gs_hallmark <- dplyr::filter(msigdb_gene_sets, gs_cat == "H")
-gs_hallmark <- split(x = gs_hallmark$gene_symbol, f = gs_hallmark$gs_name)
-
-# read msigdb data: pancreas_endocrine
-f <- system.file("extdata", "GOBP_ENDOCRINE_PANCREAS_DEVELOPMENT.v2024.1.Hs.json", package = "stgeomx")
-gobp_end_pancreas_gene_sets <- read_msigdb_json_file(f)
-# determine gene universe
-gobp_end_pancreas_genes <- intersect(fds_qq_qn$meta$gene, gobp_end_pancreas_gene_sets$gene_symbol)
-# filter genes not in our dataset
-gobp_end_pancreas_gene_sets <- gobp_end_pancreas_gene_sets %>%
-  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
-  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
-# gene set categories
-gs_end_panc <- dplyr::filter(gobp_end_pancreas_gene_sets, gs_cat == "C5:GO:BP")
-gs_end_panc <- split(x = gs_end_panc$gene_symbol, f = gs_end_panc$gs_name)
-
-# read msigdb data: pancreas secretion
-f <- system.file("extdata", "GOBP_INSULIN_SECRETION.v2024.1.Hs.json", package = "stgeomx")
-gobp_pancreas_secretion_gene_sets <- read_msigdb_json_file(f)
-# determine gene universe
-gobp_pancreas_secretion_genes <- intersect(fds_qq_qn$meta$gene, gobp_pancreas_secretion_gene_sets$gene_symbol)
-# filter genes not in our dataset
-gobp_pancreas_secretion_gene_sets <- gobp_pancreas_secretion_gene_sets %>%
-  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
-  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
-# gene set categories
-gs_panc_sec <- dplyr::filter(gobp_pancreas_secretion_gene_sets, gs_cat == "C5:GO:BP")
-gs_panc_sec <- split(x = gs_panc_sec$gene_symbol, f = gs_panc_sec$gs_name)
-
-# read msigdb data: brain development gene sets
-f <- system.file("extdata", "brain_genesets.v2024.1.Hs.json", package = "stgeomx")
-gobp_brain_gene_sets <- read_msigdb_json_file(f)
-# determine gene universe
-gobp_brain_genes <- intersect(fds_qq_qn$meta$gene, gobp_brain_gene_sets$gene_symbol)
-# filter genes not in our dataset
-gobp_brain_gene_sets <- gobp_brain_gene_sets %>%
-  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
-  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
-# gene set categories
-gs_brain <- dplyr::filter(gobp_brain_gene_sets, gs_cat == "C5:GO:BP")
-gs_brain <- split(x = gs_brain$gene_symbol, f = gs_brain$gs_name)
-
-# read msigdb data: ln development gene sets
-f <- system.file("extdata", "ln_genesets.v2024.1.Hs.json", package = "stgeomx")
-gobp_ln_gene_sets <- read_msigdb_json_file(f)
-# determine gene universe
-gobp_ln_genes <- intersect(fds_qq_qn$meta$gene, gobp_ln_gene_sets$gene_symbol)
-# filter genes not in our dataset
-gobp_ln_gene_sets <- gobp_ln_gene_sets %>%
-  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
-  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
-# gene set categories
-gs_ln <- dplyr::filter(gobp_ln_gene_sets, gs_cat == "C5:GO:BP")
-gs_ln <- split(x = gs_ln$gene_symbol, f = gs_ln$gs_name)
-
-# read msigdb data: kidney development gene sets
-f <- system.file("extdata", "kidney_genesets.v2024.1.Hs.json", package = "stgeomx")
-gobp_kidney_gene_sets <- read_msigdb_json_file(f)
-# determine gene universe
-gobp_kidney_genes <- intersect(fds_qq_qn$meta$gene, gobp_kidney_gene_sets$gene_symbol)
-# filter genes not in our dataset
-gobp_kidney_gene_sets <- gobp_kidney_gene_sets %>%
-  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
-  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
-# gene set categories
-gs_kidney <- dplyr::filter(gobp_kidney_gene_sets, gs_cat == "C5:GO:BP")
-gs_kidney <- split(x = gs_kidney$gene_symbol, f = gs_kidney$gs_name)
-
-# read msigdb data: colon development gene sets
-f <- system.file("extdata", "colon_genesets.v2024.1.Hs.json", package = "stgeomx")
-gobp_colon_gene_sets <- read_msigdb_json_file(f)
-# determine gene universe
-gobp_colon_genes <- intersect(fds_qq_qn$meta$gene, gobp_colon_gene_sets$gene_symbol)
-# filter genes not in our dataset
-gobp_colon_gene_sets <- gobp_colon_gene_sets %>%
-  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
-  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
-# gene set categories
-gs_colon <- dplyr::filter(gobp_colon_gene_sets, gs_cat == "C5:GO:BP")
-gs_colon <- split(x = gs_colon$gene_symbol, f = gs_colon$gs_name)
-
-# new gsea code
-get_fgsea_ranks_de <- function(xde, mycontrast) {
-  ranks <- xde %>%
-    dplyr::filter(contrast == mycontrast) %>%
-    select(gene, log2fc, padj) %>%
-    mutate(rank = log2fc)
-  ranks = sort(setNames(ranks$rank, ranks$gene), decreasing = TRUE)
-  return(ranks)
-}
-
-fgsea_run_batch <- function(xde, xgs,
-                            fgsea_eps=0,
-                            fgsea_nperms=1000) {
-  gsea <- NULL
-  for (xcontrast in unique(xde$contrast)) {
-    ranks <- get_fgsea_ranks_de(xde, xcontrast)
-    res <- fgsea(pathways=xgs,
-                 stats=ranks,
-                 eps=fgsea_eps,
-                 nPermSimple=fgsea_nperms)
-    res <- mutate(res, contrast = xcontrast)
-    gsea <- bind_rows(gsea, res)     
-  }
-  return(gsea)
-}
-
-stgeomx_gsea_pipeline <- function(gsea, de, method) {
-  res <- fgsea_run_batch(de, gs_hallmark,
-                       fgsea_eps=0, fgsea_nperms=gsea_nperms)
-  res <- res %>% mutate(method = method)
-  gsea <- bind_rows(gsea, res)
-
-  res <- fgsea_run_batch(de, gs_ex_panc,
-                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
-  res <- res %>% mutate(method = method)
-  gsea <- bind_rows(gsea, res)
-
-  res <- fgsea_run_batch(de, gs_end_panc,
-                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
-  res <- res %>% mutate(method = method)
-  gsea <- bind_rows(gsea, res)
-
-  res <- fgsea_run_batch(de, gs_panc_sec,
-                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
-  res <- res %>% mutate(method = method)
-  gsea <- bind_rows(gsea, res)
-
-  res <- fgsea_run_batch(de, gs_brain,
-                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
-  res <- res %>% mutate(method = method)
-  gsea <- bind_rows(gsea, res)
-
-  res <- fgsea_run_batch(de, gs_ln,
-                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
-  res <- res %>% mutate(method = method)
-  gsea <- bind_rows(gsea, res)
-
-  res <- fgsea_run_batch(de, gs_kidney,
-                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
-  res <- res %>% mutate(method = method)
-  gsea <- bind_rows(gsea, res)
-
-  res <- fgsea_run_batch(de, gs_colon,
-                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
-  res <- res %>% mutate(method = method)
-  gsea <- bind_rows(gsea, res)
-
-  # filter out irrelevant genes in hallmark
-  gsea <- gsea %>%
-  filter(!(pathway %in% c("HALLMARK_ESTROGEN_RESPONSE_LATE", "HALLMARK_ESTROGEN_RESPONSE_EARLY", "HALLMARK_GLYCOLYSIS", "HALLMARK_P53_PATHWAY", "HALLMARK_HEME_METABOLISM", "HALLMARK_ADIPOGENESIS", "HALLMARK_MYC_TARGETS_V2", "HALLMARK_MYC_TARGETS_V1", "HALLMARK_OXIDATIVE_PHOSPHORYLATION", "HALLMARK_HYPOXIA", "HALLMARK_UV_RESPONSE_DN", "HALLMARK_MYOGENESIS", "HALLMARK_APICAL_SURFACE", "HALLMARK_APOPTOSIS", "HALLMARK_IL6_JAK_STAT3_SIGNALING", "HALLMARK_IL2_STAT5_SIGNALING", "HALLMARK_MITOTIC_SPINDLE", "HALLMARK_APICAL_JUNCTION", "HALLMARK_KRAS_SIGNALING_UP", "HALLMARK_KRAS_SIGNALING_DN", "HALLMARK_UNFOLDED_PROTEIN_RESPONSE", "HALLMARK_DNA_REPAIR", "HALLMARK_FATTY_ACID_METABOLISM", "HALLMARK_HEDGEHOG_SIGNALING", "HALLMARK_ANGIOGENESIS", "HALLMARK_E2F_TARGETS", "HALLMARK_G2M_CHECKPOINT", "HALLMARK_TNFA_SIGNALING_VIA_NFKB", "HALLMARK_PROTEIN_SECRETION", "HALLMARK_CHOLESTEROL_HOMEOSTASIS")))
-
-  return(gsea)
-}
-
-# Run full analysis
-gsea_all <- NULL
-gsea_none_q3 <- NULL
-gsea_none_tmm <- NULL
-gsea_none_rle <- NULL
-gsea_none_qn <- NULL
-
-gsea_bgsub_q3 <- NULL
-gsea_bgsub_tmm <- NULL
-gsea_bgsub_rle <- NULL
-gsea_bgsub_qn <- NULL
-
-gsea_qq_q3 <- NULL
-gsea_qq_tmm <- NULL
-gsea_qq_rle <- NULL
-gsea_qq_qn <- NULL
-
-# run stgeomx pipeline
-gsea_none_q3 <- stgeomx_gsea_pipeline(gsea_none_q3, de_none_q3, "none_q3")
-gsea_none_tmm <- stgeomx_gsea_pipeline(gsea_none_tmm, de_none_tmm, "none_tmm")
-gsea_none_rle <- stgeomx_gsea_pipeline(gsea_none_rle, de_none_rle, "none_rle")
-gsea_none_qn <- stgeomx_gsea_pipeline(gsea_none_qn, de_none_qn, "none_qn")
-
-gsea_bgsub_q3 <- stgeomx_gsea_pipeline(gsea_bgsub_q3, de_bgsub_q3, "bgsub_q3")
-gsea_bgsub_tmm <- stgeomx_gsea_pipeline(gsea_bgsub_tmm, de_bgsub_tmm, "bgsub_tmm")
-gsea_bgsub_rle <- stgeomx_gsea_pipeline(gsea_bgsub_rle, de_bgsub_rle, "bgsub_rle")
-gsea_bgsub_qn <- stgeomx_gsea_pipeline(gsea_bgsub_qn, de_bgsub_qn, "bgsub_qn")
-
-gsea_qq_q3 <- stgeomx_gsea_pipeline(gsea_qq_q3, de_qq_q3, "qq_q3")
-gsea_qq_tmm <- stgeomx_gsea_pipeline(gsea_qq_tmm, de_qq_tmm, "qq_tmm")
-gsea_qq_rle <- stgeomx_gsea_pipeline(gsea_qq_rle, de_qq_rle, "qq_rle")
-gsea_qq_qn <- stgeomx_gsea_pipeline(gsea_qq_qn, de_qq_qn, "qq_qn")
-
-# # write results for clusters
-# contrasts <- c("clusterPanc", "clusterBrain", "clusterLn", "clusterKidney", "clusterColon")
-# manuscript_pathways <- c(names(gs_hallmark))
-# res <- gsea_all %>% filter(contrast %in% contrasts, 
-#                            pathway %in% manuscript_pathways)
-
-# prefix <- "gsea_all"
-# f <- file.path(results_dir, paste0(prefix, "_results.tsv"))
-# data.table::fwrite(res, file=f, sep="\t", sep2=c("", " ", ""))
-# f <- file.path(results_dir, paste0(prefix, "_results.xlsx"))
-# write_xlsx(res, f)
-```
-
-### GSEA: all msigdb
-```{r all_msigdb}
-gsea_nperms <- 10000
-
-read_msigdb_json_file <- function(filename) {
-  x <- jsonlite::fromJSON(filename)
-  y <- tibble(gs_name = names(x), gs = x) %>% 
-    unnest_wider(gs) %>%
-    unnest_longer(geneSymbols) %>%
-    select(gs_cat = collection,
-           gs_name = gs_name,
-           gene_symbol = geneSymbols)
-}
-
-# read msigdb data
-
-f <- system.file("extdata", "msigdb.v2024.1.Hs.json", package = "stgeomx")
-msigdb_gene_sets <- read_msigdb_json_file(f)
-
-# determine gene universe
-msigdb_genes <- intersect(gene_universe, msigdb_gene_sets$gene_symbol)
-
-# filter genes not in our dataset
-msigdb_gene_sets <- msigdb_gene_sets %>%
-  dplyr::filter(gene_symbol %in% gene_universe) %>%
   distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
@@
     dplyr::filter(contrast == mycontrast) %>%
-    select(gene, log2fc, padj) %>%
+    dplyr::select(gene, log2fc, padj) %>%
     mutate(rank = log2fc)
